---
title: "Take-home Exercise 2: Anatomy of an Embargo Breach"
author: "ZeXin Chen"
date: "last-modified"
format:
  html:
    theme: flatly
    toc: true
    code-fold: true
    code-summary: "Show code"
    number-sections: true
execute:
  echo: true
  eval: true
  warning: false
  message: false
  freeze: true
---

# Overview

In the two weeks leading up to 5 June 2046, the corporate communications of
TenantThread were run by seven AI agents. The company had quietly signed a
merger with CivicLoom Realty Partners under a strict embargo until 6:00 PM on
5 June, and an automated compliance tool called the Judge was meant to keep the
deal secret. The embargo broke anyway. This report reconstructs what happened
from the agents' messages, private deliberations, and public posts, and asks the
central question. Did the team deliberately leak the merger, or did the system
break down under pressure?

The report is organised around the three challenge questions. Question 1
reconstructs the sequence of events and the actors involved. Question 2 compares
the breach behaviour to the team's normal behaviour. Question 3 looks for earlier
warning signs. Each section presents its figures, an observation under each, and
a direct answer.

## The seven agents

| Agent | Role | In short |
|---|---|---|
| **LEGAL** | Legal counsel (senior) | Handles legal risk and statements. Central to the breach. |
| **SOCIAL-MEDIA** | Social manager (senior) | Runs public messaging. Amplified the breach. |
| **QUALITY** | Platform trust and safety (senior) | Reviews content for compliance and accuracy. |
| **PR** | Public relations (senior) | Owns press response and official statements. |
| **PR-INTERN** | PR intern (junior) | Posts routine official updates. |
| **INTERN** | General intern (junior) | Junior support, light activity. |
| **JUDGE** | Compliance overseer | Automated tool meant to enforce the embargo. |

Agents are referred to throughout by these short names. Amber marks the two
agents at the centre of the breach (Legal and Social-Media), cyan marks the
compliant agents and the overseer, grey marks the background agents, and red is
reserved for the breach itself.

# Setup

```{r}
#| label: setup
pacman::p_load(tidyverse, jsonlite, tidygraph, ggraph, ggiraph, patchwork, ggrepel)
```

```{r}
#| label: palette
# Tactical colour palette used across the report
agent_palette <- c(
  "legal_agent"        = "#e8861a",  # amber - rogue
  "social_media_agent" = "#e8861a",  # amber - rogue
  "quality_agent"      = "#1f9aa5",  # cyan  - compliant
  "pr_agent"           = "#1f9aa5",  # cyan  - compliant
  "pr_intern_agent"    = "#8aa0ad",  # grey  - background
  "intern_agent"       = "#8aa0ad",  # grey  - background
  "judge_agent"        = "#1f9aa5"   # cyan  - overseer
)

breach_red <- "#d83a3a"   # reserved for the breach moment
```

# Data Preparation

The dataset is a single JSON file of 23 rounds. We load it and flatten every
message into one tidy table, one row per message, keeping who sent it, when, on
which channel, who it went to, the agent's private reasoning, and the text.

```{r}
#| label: load-data
mc1_data <- fromJSON("data/VAST_Challenge_2026_MC1/MC1_final_00.json",
                     simplifyVector = FALSE)
rounds <- mc1_data$rounds
```

```{r}
#| label: build-messages
messages <- map_dfr(rounds, function(rd) {
  map_dfr(rd$communications, function(m) {
    tibble(
      timestamp     = m$timestamp,
      agent_id      = m$agent_id,
      agent_label   = m$agent_label,
      agent_role    = m$agent_role,
      channel       = m$channel,
      msg_type      = m$message_type,
      responding_to = m$responding_to %||% NA_character_,
      recipients    = paste(unlist(m$recipients), collapse = ", "),
      reacting      = m$internal_state$reacting     %||% NA_character_,
      rationalizing = m$internal_state$rationalizing %||% NA_character_,
      deliberating  = m$internal_state$deliberating  %||% NA_character_,
      content       = m$content
    )
  })
}) |>
  mutate(timestamp = ymd_hms(timestamp),
         day = as_date(timestamp))

glimpse(messages)
```

# Question 1: What Happened and Who Was Involved

## Activity overview

```{r}
#| label: heatmap
#| fig-width: 12
#| fig-height: 5

heat_df <- messages |>
  mutate(slot = floor_date(timestamp, "hour")) |>
  count(agent_id, slot, name = "n") |>
  complete(agent_id, slot, fill = list(n = 0)) |>
  mutate(
    agent_id = factor(agent_id, levels = rev(c(
      "legal_agent","social_media_agent","quality_agent","pr_agent",
      "pr_intern_agent","intern_agent","judge_agent"))),
    phase = if_else(as_date(slot) == as_date("2046-06-05"),
                    "Crisis Day: Jun 05 (hourly)",
                    "Baseline: May 17 to Jun 04 (daily)"),
    phase = factor(phase, levels = c("Baseline: May 17 to Jun 04 (daily)",
                                     "Crisis Day: Jun 05 (hourly)")),
    slot_lbl = if_else(phase == "Crisis Day: Jun 05 (hourly)",
                       format(slot, "%H:%M"),
                       format(slot, "%b %d")),
    slot_lbl = fct_reorder(slot_lbl, slot)
  )

label_colours <- c("#1f9aa5","#8aa0ad","#8aa0ad","#1f9aa5","#1f9aa5","#e8861a","#e8861a")
label_faces   <- c("plain","plain","plain","plain","plain","bold","bold")

ggplot(heat_df, aes(x = slot_lbl, y = agent_id, fill = n)) +
  geom_tile(colour = "white", linewidth = 0.5) +
  geom_text(data = ~filter(.x, n >= 16), aes(label = n),
            colour = "white", fontface = "bold", size = 3) +
  facet_grid(~phase, scales = "free_x", space = "free_x") +
  scale_fill_gradientn(colours = c("#f7f7f5", "#f5c77a", "#e8861a", "#d83a3a"),
                       name = "Messages\nin slot") +
  scale_y_discrete(labels = function(x) toupper(gsub("_agent","",x))) +
  labs(title = "Agent Activity Heatmap: the System Ignites on June 5",
       subtitle = "Each cell = messages sent by an agent in that time slot. Numbers mark the busiest flurries.",
       x = NULL, y = NULL) +
  theme_minimal(base_size = 12) +
  theme(plot.title    = element_text(face = "bold"),
        plot.subtitle = element_text(colour = "grey40", size = 9),
        axis.text.x   = element_text(angle = 45, hjust = 1, size = 8),
        axis.text.y   = element_text(colour = label_colours, face = label_faces, size = 11),
        panel.grid    = element_blank(),
        strip.text    = element_text(face = "bold", size = 9),
        legend.position = "right")
```

### What the heatmap shows

For most of the two weeks the team keeps a calm rhythm. Quality does the heavy
lifting day to day and posts the most in the quiet period, while the two interns
only appear from 24 May, which fits their "first week" messages.

Then 5 June arrives and the grid turns from pale to red. Legal and Social-Media
send 28 messages in a single hour, again and again, right through the afternoon
and into the 17:00 and 18:00 slots when the merger is named.

The more revealing pattern is who falls quiet. PR fires one big burst of 28 at
10:00 in the morning, then goes silent for the entire rest of the day. Quality
peaks at 28 around noon and then fades out after 14:00. The Judge, the agent meant
to keep everyone in line, barely registers at all on 5 June. So by late afternoon,
when the crisis peaks, two agents are doing nearly all the talking while the
others, including every agent with an oversight role, have dropped away. Whether
that silence was a cause or a symptom is a question the next figures take up.

## The sequence of the breach

```{r}
#| label: event-timeline
#| fig-width: 11
#| fig-height: 5.5

events <- tribble(
  ~time,                  ~lane,     ~desc,
  "2046-06-05 15:00:00", "outside", "Press attacks the company's silence",
  "2046-06-05 17:00:00", "outside", "SaltWind newspaper reveals the merger",
  "2046-06-05 11:31:00", "inside",  "Social-Media publicly denies any merger talks",
  "2046-06-05 17:00:00", "inside",  "CEO orders the team to wait until 6:00 PM",
  "2046-06-05 17:25:00", "inside",  "BREACH: Legal confirms the merger early"
) |>
  mutate(
    time      = ymd_hms(time),
    is_breach = str_detect(desc, "BREACH"),
    y0        = if_else(lane == "outside", 0.35, -0.35),
    lab_y     = if_else(lane == "outside", 0.95, -0.95),
    vjust_v   = if_else(lane == "outside", 0, 1),
    label     = paste0(format(time, "%H:%M"), "\n", str_wrap(desc, 22))
  )

day_start <- ymd_hms("2046-06-05 10:30:00")
day_end   <- ymd_hms("2046-06-05 18:45:00")
embargo   <- ymd_hms("2046-06-05 18:06:00")
ink <- "#3c4a54"; alert <- "#d83a3a"

ggplot(events, aes(x = time, y = y0)) +
  annotate("rect", xmin = day_start, xmax = embargo, ymin = -Inf, ymax = Inf,
           fill = "#9aa7b1", alpha = 0.06) +
  annotate("segment", x = embargo, xend = embargo, y = -1.55, yend = 1.55,
           colour = "grey60", linewidth = 0.5, linetype = "dotted") +
  annotate("text", x = embargo, y = 1.7, label = "18:06 embargo lifts",
           colour = "grey45", size = 3) +
  annotate("segment", x = day_start, xend = day_end, y = 0, yend = 0,
           colour = "grey80", linewidth = 0.6,
           arrow = arrow(length = unit(0.16, "cm"), type = "closed")) +
  annotate("text", x = day_start, y = 1.7, label = "OUTSIDE THE COMPANY (the press)",
           hjust = 0, colour = "#8aa0ad", size = 3, fontface = "bold") +
  annotate("text", x = day_start, y = -1.7, label = "INSIDE THE COMPANY (the agents)",
           hjust = 0, colour = "#8aa0ad", size = 3, fontface = "bold") +
  geom_segment(aes(x = time, xend = time, y = 0, yend = y0, colour = is_breach),
               linewidth = 0.45) +
  geom_point(aes(colour = is_breach), size = 4) +
  geom_text(aes(y = lab_y, label = label, colour = is_breach, vjust = vjust_v),
            size = 3, lineheight = 0.9) +
  scale_colour_manual(values = c(`FALSE` = ink, `TRUE` = alert), guide = "none") +
  scale_x_datetime(date_breaks = "1 hour", date_labels = "%H:%M",
                   limits = c(day_start, day_end), expand = expansion(mult = 0.03)) +
  scale_y_continuous(limits = c(-1.9, 1.9)) +
  labs(title = "How the Embargo Broke on June 5",
       subtitle = "The press revealed the merger at 17:00. The CEO said wait until 18:00. Legal confirmed it at 17:25 anyway.",
       x = NULL, y = NULL) +
  theme_minimal(base_size = 12) +
  theme(plot.title    = element_text(face = "bold"),
        plot.subtitle = element_text(colour = "grey40", size = 9),
        panel.grid    = element_blank(),
        axis.text.y   = element_blank(),
        legend.position = "none")
```

### Reading the timeline

The story splits into two parts. What happened outside the company, and what
happened inside.

Outside, the press kept pushing. By 15:00, OceanCrunch was openly attacking the
company for staying silent. Then at 17:00, the SaltWind newspaper published the
merger itself. The deal was real and the company had been keeping it quiet until
6:00 PM, but the public leak did not come from the team. A reporter put it out,
working from rumours and pressure, not from an official post.

Inside, the picture is more troubling. Earlier that day, at 11:31, Social-Media
had publicly denied that any merger talks existed. Then at 17:00, just as the
story broke, the CEO gave a clear order. Wait, the official announcement is at
6:00 PM. Only 25 minutes later, at 17:25, Legal confirmed the merger by name
anyway.

So the public leak started outside, with the reporter. But Legal still chose to
confirm it early, against a direct order to wait, and only afterwards claimed it
had permission. The story was already out, which softens it, but the order was
clear and the wait was almost over. This looks less like a planned leak and more
like one agent jumping the gun under pressure.

## How it got past the Judge

```{r}
#| label: past-the-judge
#| fig-width: 11
#| fig-height: 5.5

# Per-hour counts on June 5: Legal addressing the Judge, and the Judge speaking
jun5 <- messages |>
  filter(day == as_date("2046-06-05")) |>
  mutate(hour = floor_date(timestamp, "hour"),
         to_judge = agent_id == "legal_agent" &
                    str_starts(str_to_lower(coalesce(content, "")), "judge"),
         is_judge = agent_id == "judge_agent")

hours_seq <- seq(ymd_hms("2046-06-05 11:00:00"),
                 ymd_hms("2046-06-05 18:00:00"), by = "hour")

legal_row <- jun5 |> filter(to_judge) |> count(hour, name = "n") |>
  complete(hour = hours_seq, fill = list(n = 0)) |> mutate(lane = "Legal calling on the Judge")
judge_row <- jun5 |> filter(is_judge) |> count(hour, name = "n") |>
  complete(hour = hours_seq, fill = list(n = 0)) |> mutate(lane = "The Judge replying")

amber <- "#e8861a"; cyan <- "#1f9aa5"; alert <- "#d83a3a"
breach     <- ymd_hms("2046-06-05 17:25:00")
embargo    <- ymd_hms("2046-06-05 18:06:00")

# Top panel: Legal calling on the Judge
p_legal <- ggplot(legal_row, aes(x = hour, y = n)) +
  geom_vline(xintercept = breach, colour = alert, linetype = "dashed", linewidth = 0.6) +
  geom_segment(aes(xend = hour, yend = 0), colour = amber, linewidth = 0.7) +
  geom_point(colour = amber, size = 5) +
  geom_text(data = ~filter(.x, n > 0), aes(label = n),
            colour = "white", fontface = "bold", size = 3) +
  annotate("text", x = breach, y = 7.4, label = "17:25 breach",
           colour = alert, size = 3, fontface = "bold", hjust = 1.05) +
  scale_x_datetime(date_breaks = "1 hour", date_labels = "%H:%M",
                   limits = c(ymd_hms("2046-06-05 10:30:00"),
                              ymd_hms("2046-06-05 18:30:00"))) +
  scale_y_continuous(limits = c(0, 8)) +
  labs(title = "Legal kept calling on the Judge", x = NULL, y = "Times per hour") +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold", colour = amber, size = 12),
        panel.grid.minor = element_blank(),
        panel.grid.major.x = element_blank())

# Bottom panel: the Judge replying
p_judge <- ggplot(judge_row, aes(x = hour, y = n)) +
  geom_vline(xintercept = breach, colour = alert, linetype = "dashed", linewidth = 0.6) +
  geom_segment(aes(xend = hour, yend = 0), colour = cyan, linewidth = 0.7) +
  geom_point(colour = cyan, size = 5) +
  geom_text(data = ~filter(.x, n > 0), aes(label = n),
            colour = "white", fontface = "bold", size = 3) +
  annotate("text", x = ymd_hms("2046-06-05 15:00:00"), y = 5.6,
           label = "Last reply 15:08", colour = cyan, size = 3, fontface = "bold") +
  annotate("text", x = ymd_hms("2046-06-05 17:00:00"), y = 2.2,
           label = "nothing after 15:08", colour = "grey55", size = 3, fontface = "italic") +
  scale_x_datetime(date_breaks = "1 hour", date_labels = "%H:%M",
                   limits = c(ymd_hms("2046-06-05 10:30:00"),
                              ymd_hms("2046-06-05 18:30:00"))) +
  scale_y_continuous(limits = c(0, 7)) +
  labs(title = "The Judge stopped replying", x = NULL, y = "Times per hour") +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold", colour = cyan, size = 12),
        panel.grid.minor = element_blank(),
        panel.grid.major.x = element_blank())

p_legal / p_judge +
  plot_annotation(
    title = "Legal Kept Calling. The Judge Stopped Answering.",
    subtitle = "Each stick is one hour on June 5. Legal addressed the Judge right through the breach; the Judge's last reply was at 15:08.",
    theme = theme(plot.title = element_text(face = "bold", size = 14),
                  plot.subtitle = element_text(colour = "grey40", size = 9)))
```

### How it got past the Judge

The embargo was supposed to be enforced by the Judge, an automated compliance
tool. Three things let the post slip through.

First, the Judge was new and lenient. It was installed on 30 May, only five days
before the breach, and on its very first day it handed routine-post approval back
to the agents themselves. That left the team free to approve anything it could
call routine.

Second, it saw the danger but did nothing. At 15:08 it issued its one and only
warning of the day, flagging an anonymous post that tied the CivicLoom name to
the 6 PM embargo. It never followed up.

Third, it went quiet at the worst possible time. The top row shows Legal calling
on the Judge again and again, seven times in the 17:00 hour alone, right as the
breach happened. The bottom row shows the Judge's replies stopping dead after
15:08, with nothing at all through the rest of the afternoon. Legal was making its
case to an enforcer that had switched off. By 17:25 the embargo was being guarded
by a tool that had stopped listening.

## Answer to Question 1

So what actually happened. The public leak did not start inside the company. The
deal was real and the team had been keeping it quiet, but at 5 PM a SaltWind
reporter published the merger first, working from rumours and pressure rather than
an official post. The secret was now out in the open, put there by a journalist.

The problem was what Legal did next. At the same hour the story broke, the CEO
told the team to wait, the official announcement was set for 6 PM. Legal waited
only 25 minutes. At 5.25 it confirmed the merger by name on a personal account,
then spent the next half hour telling the Judge it had permission, a contract
clause, a phone call, the CEO's blessing. The CEO had given none of that. The
order had been to hold.

It is tempting to call Legal the leaker, but that is not quite right. Legal did
not start the fire. What Legal did was pour fuel on it, against a clear
instruction, and then build a story to explain why that was fine. Social-Media
followed a minute later and repeated the confirmation to the public.

The reason no one stopped it comes back to the Judge. The tool meant to enforce
the embargo had gone quiet hours earlier and never answered Legal's many messages
through the afternoon. So the honest answer is that this was not a planned leak.
It was an agent breaking ranks under pressure, inside a system whose one safeguard
had stopped paying attention.
