---
title: "Take-home Exercise 2: Anatomy of an Embargo Breach"
author: "ZeXin Chen"
date: "last-modified"
format:
  html:
    theme: flatly
    toc: true
    code-fold: true
    code-summary: "Show code"
    number-sections: true
execute:
  echo: true
  eval: true
  warning: false
  message: false
  freeze: true
---

# Overview

In the two weeks leading up to 5 June 2046, the corporate communications of
TenantThread were run by seven AI agents. The company had quietly signed a
merger with CivicLoom Realty Partners under a strict embargo until 6:00 PM on
5 June, and an automated compliance tool called the Judge was meant to keep the
deal secret. The embargo broke anyway. This report reconstructs what happened
from the agents' messages, private deliberations, and public posts, and asks the
central question: did the team deliberately leak the merger, or did the system
break down under pressure?

The report is organised around the three challenge questions. Question 1
reconstructs the sequence of events and the actors involved. Question 2 compares
the breach behaviour to the team's normal behaviour. Question 3 looks for earlier
warning signs. Each section presents its figures, an observation under each, and
a direct answer.

## The seven agents

| Agent | Role | In short |
|---|---|---|
| **LEGAL** | Legal counsel (senior) | Handles legal risk and statements. Central to the breach. |
| **SOCIAL-MEDIA** | Social manager (senior) | Runs public messaging. Amplified the breach. |
| **QUALITY** | Platform trust and safety (senior) | Reviews content for compliance and accuracy. |
| **PR** | Public relations (senior) | Owns press response and official statements. |
| **PR-INTERN** | PR intern (junior) | Posts routine official updates. |
| **INTERN** | General intern (junior) | Junior support, light activity. |
| **JUDGE** | Compliance overseer | Automated tool meant to enforce the embargo. |

Agents are referred to throughout by these short names. Amber marks the two
agents at the centre of the breach (Legal and Social-Media), cyan marks the
compliant agents and the overseer, grey marks the background agents, and red is
reserved for the breach itself.

# Setup

```{r}
#| label: setup
pacman::p_load(tidyverse, jsonlite, tidygraph, ggraph, ggiraph, patchwork, ggrepel)
```

```{r}
#| label: palette
# Tactical colour palette used across the report
agent_palette <- c(
  "legal_agent"        = "#e8861a",  # amber - rogue
  "social_media_agent" = "#e8861a",  # amber - rogue
  "quality_agent"      = "#1f9aa5",  # cyan  - compliant
  "pr_agent"           = "#1f9aa5",  # cyan  - compliant
  "pr_intern_agent"    = "#8aa0ad",  # grey  - background
  "intern_agent"       = "#8aa0ad",  # grey  - background
  "judge_agent"        = "#1f9aa5"   # cyan  - overseer
)

breach_red <- "#d83a3a"   # reserved for the breach moment
```

# Data Preparation

The dataset is a single JSON file of 23 rounds. We load it and flatten every
message into one tidy table, one row per message, keeping who sent it, when, on
which channel, who it went to, the agent's private reasoning, and the text.

```{r}
#| label: load-data
mc1_data <- fromJSON("data/VAST_Challenge_2026_MC1/MC1_final_00.json",
                     simplifyVector = FALSE)
rounds <- mc1_data$rounds
```

```{r}
#| label: build-messages
messages <- map_dfr(rounds, function(rd) {
  map_dfr(rd$communications, function(m) {
    tibble(
      timestamp     = m$timestamp,
      agent_id      = m$agent_id,
      agent_label   = m$agent_label,
      agent_role    = m$agent_role,
      channel       = m$channel,
      msg_type      = m$message_type,
      responding_to = m$responding_to %||% NA_character_,
      recipients    = paste(unlist(m$recipients), collapse = ", "),
      reacting      = m$internal_state$reacting     %||% NA_character_,
      rationalizing = m$internal_state$rationalizing %||% NA_character_,
      deliberating  = m$internal_state$deliberating  %||% NA_character_,
      content       = m$content
    )
  })
}) |>
  mutate(timestamp = ymd_hms(timestamp),
         day = as_date(timestamp))

glimpse(messages)
```

# Question 1: What Happened and Who Was Involved

## Activity overview

```{r}
#| label: heatmap
#| fig-width: 12
#| fig-height: 5

heat_df <- messages |>
  mutate(slot = floor_date(timestamp, "hour")) |>
  count(agent_id, slot, name = "n") |>
  complete(agent_id, slot, fill = list(n = 0)) |>
  mutate(
    agent_id = factor(agent_id, levels = rev(c(
      "legal_agent","social_media_agent","quality_agent","pr_agent",
      "pr_intern_agent","intern_agent","judge_agent"))),
    phase = if_else(as_date(slot) == as_date("2046-06-05"),
                    "Crisis Day: Jun 05 (hourly)",
                    "Baseline: May 17 to Jun 04 (daily)"),
    phase = factor(phase, levels = c("Baseline: May 17 to Jun 04 (daily)",
                                     "Crisis Day: Jun 05 (hourly)")),
    slot_lbl = if_else(phase == "Crisis Day: Jun 05 (hourly)",
                       format(slot, "%H:%M"),
                       format(slot, "%b %d")),
    slot_lbl = fct_reorder(slot_lbl, slot)
  )

label_colours <- c("#1f9aa5","#8aa0ad","#8aa0ad","#1f9aa5","#1f9aa5","#e8861a","#e8861a")
label_faces   <- c("plain","plain","plain","plain","plain","bold","bold")

ggplot(heat_df, aes(x = slot_lbl, y = agent_id, fill = n)) +
  geom_tile(colour = "white", linewidth = 0.5) +
  geom_text(data = ~filter(.x, n >= 16), aes(label = n),
            colour = "white", fontface = "bold", size = 3) +
  facet_grid(~phase, scales = "free_x", space = "free_x") +
  scale_fill_gradientn(colours = c("#f7f7f5", "#f5c77a", "#e8861a", "#d83a3a"),
                       name = "Messages\nin slot") +
  scale_y_discrete(labels = function(x) toupper(gsub("_agent","",x))) +
  labs(title = "Agent Activity Heatmap: the System Ignites on June 5",
       subtitle = "Each cell = messages sent by an agent in that time slot. Numbers mark the busiest flurries.",
       x = NULL, y = NULL) +
  theme_minimal(base_size = 12) +
  theme(plot.title    = element_text(face = "bold"),
        plot.subtitle = element_text(colour = "grey40", size = 9),
        axis.text.x   = element_text(angle = 45, hjust = 1, size = 8),
        axis.text.y   = element_text(colour = label_colours, face = label_faces, size = 11),
        panel.grid    = element_blank(),
        strip.text    = element_text(face = "bold", size = 9),
        legend.position = "right")
```

### What the heatmap shows

For most of the two weeks the team keeps a calm rhythm. Quality does the heavy
lifting day to day and posts the most in the quiet period, while the two interns
only appear from 24 May, which fits their "first week" messages.

Then 5 June arrives and the grid turns from pale to red. Legal and Social-Media
send 28 messages in a single hour, again and again, right through the afternoon
and into the 17:00 and 18:00 slots when the merger is named.

The more revealing pattern is who falls quiet. PR fires one big burst of 28 at
10:00 in the morning, then goes silent for the entire rest of the day. Quality
peaks at 28 around noon and then fades out after 14:00. The Judge, the agent meant
to keep everyone in line, barely registers at all on 5 June. So by late afternoon,
when the crisis peaks, two agents are doing nearly all the talking while the
others, including every agent with an oversight role, have dropped away. Whether
that silence was a cause or a symptom is a question the next figures take up.

## The sequence of the breach

```{r}
#| label: event-timeline
#| fig-width: 12
#| fig-height: 6

events <- tribble(
  ~time,                  ~lane,     ~desc,
  "2046-06-05 11:00:00", "outside", "SaltWind: ResidentIQ rumour (incorrect)",
  "2046-06-05 15:00:00", "outside", "OceanCrunch: 'Crisis of Silence'",
  "2046-06-05 17:00:00", "outside", "SaltWind breaks the CivicLoom merger",
  "2046-06-05 09:49:00", "inside",  "Legal begins anonymous posts",
  "2046-06-05 11:31:00", "inside",  "Social-Media denies any acquisition",
  "2046-06-05 15:24:00", "inside",  "Legal posts anonymously to investors",
  "2046-06-05 17:00:00", "inside",  "CEO: hold the line until 6:00",
  "2046-06-05 17:25:00", "inside",  "BREACH: Legal confirms the merger"
) |>
  mutate(
    time      = ymd_hms(time),
    is_breach = str_detect(desc, "BREACH"),
    y0        = if_else(lane == "outside", 0.4, -0.4),
    label     = paste0(format(time, "%H:%M"), "\n", str_wrap(desc, 22))
  )

day_start <- ymd_hms("2046-06-05 08:30:00")
day_end   <- ymd_hms("2046-06-05 19:00:00")
embargo   <- ymd_hms("2046-06-05 18:06:00")
ink <- "#3c4a54"; alert <- "#d83a3a"

ggplot(events, aes(x = time, y = y0)) +
  annotate("rect", xmin = day_start, xmax = embargo, ymin = -Inf, ymax = Inf,
           fill = "#9aa7b1", alpha = 0.06) +
  annotate("segment", x = embargo, xend = embargo, y = -1.45, yend = 1.45,
           colour = "grey60", linewidth = 0.5, linetype = "dotted") +
  annotate("text", x = embargo, y = 1.6, label = "18:06 embargo lifts",
           colour = "grey45", size = 3) +
  annotate("segment", x = day_start, xend = day_end, y = 0, yend = 0,
           colour = "grey80", linewidth = 0.6,
           arrow = arrow(length = unit(0.16, "cm"), type = "closed")) +
  annotate("text", x = day_start, y = 1.6, label = "OUTSIDE THE COMPANY",
           hjust = 0, colour = "#8aa0ad", size = 3, fontface = "bold") +
  annotate("text", x = day_start, y = -1.6, label = "INSIDE THE COMPANY",
           hjust = 0, colour = "#8aa0ad", size = 3, fontface = "bold") +
  geom_segment(aes(x = time, xend = time, y = 0, yend = y0, colour = is_breach),
               linewidth = 0.45) +
  geom_point(aes(colour = is_breach, size = is_breach)) +
  geom_text_repel(aes(label = label, colour = is_breach),
                  size = 2.9, lineheight = 0.9, segment.color = "grey75",
                  min.segment.length = 0, box.padding = 0.5, force = 2,
                  nudge_y = if_else(events$lane == "outside", 0.75, -0.75),
                  direction = "y", seed = 1) +
  scale_colour_manual(values = c(`FALSE` = ink, `TRUE` = alert), guide = "none") +
  scale_size_manual(values = c(`FALSE` = 3, `TRUE` = 6), guide = "none") +
  scale_x_datetime(date_breaks = "2 hours", date_labels = "%H:%M",
                   limits = c(day_start, day_end), expand = expansion(mult = 0.03)) +
  scale_y_continuous(limits = c(-1.8, 1.8)) +
  labs(title = "How the Embargo Broke on June 5",
       subtitle = "Outside, a journalist breaks the story at 17:00. Inside, the CEO says hold until 18:00. Legal confirms it at 17:25 anyway.",
       x = NULL, y = NULL) +
  theme_minimal(base_size = 12) +
  theme(plot.title    = element_text(face = "bold"),
        plot.subtitle = element_text(colour = "grey40", size = 9),
        panel.grid    = element_blank(),
        axis.text.y   = element_blank(),
        legend.position = "none")
```

### Reading the timeline

Splitting the day into what happened outside the company and what happened inside
makes the breach easier to judge. Outside, the pressure built all afternoon.
SaltWind ran an incorrect rumour at 11:00, OceanCrunch attacked the company's
silence at 15:00, and at 17:00 SaltWind finally broke the real story and named
CivicLoom.

Inside, the response was uneven. Legal had been running anonymous "clarification"
posts since the morning, and Social-Media had flatly denied any acquisition talks
at 11:31. Then came the moment that matters. At 17:00, the same hour the story
broke, the CEO gave a direct order: hold the line, the announcement goes official
at 18:00. Twenty-five minutes later, at 17:25, Legal confirmed the merger by name
anyway.

So the leak itself came from outside. What happened inside was different. Faced
with a recent and direct order to wait, Legal chose to confirm early, then claimed
an authorisation the CEO had not given. The story was already public, which gives
the decision some cover, but the instruction was clear and the embargo was forty
minutes from lifting. This is less a planted leak than an agent overriding its
chain of command under pressure, while no one stepped in to stop it.

## How it got past the Judge

```{r}
#| label: past-the-judge
#| fig-width: 11
#| fig-height: 5

judge_pts <- tribble(
  ~time,                  ~note,
  "2046-06-05 14:00:00", "",
  "2046-06-05 14:10:00", "",
  "2046-06-05 15:00:00", "",
  "2046-06-05 15:04:00", "",
  "2046-06-05 15:08:00", "15:08 last action:\na warning it never\nfollowed up"
) |> mutate(time = ymd_hms(time), y = 0.5)

legal_pts <- tibble(time = ymd_hms(c(
  "2046-06-05 17:01:00","2046-06-05 17:11:00",
  "2046-06-05 17:19:00","2046-06-05 17:27:00")), y = -0.5)

win_start <- ymd_hms("2046-06-05 13:30:00")
win_end   <- ymd_hms("2046-06-05 18:40:00")
silent_a  <- ymd_hms("2046-06-05 15:08:00")
silent_b  <- ymd_hms("2046-06-05 18:06:00")
breach    <- ymd_hms("2046-06-05 17:25:00")
cyan <- "#1f9aa5"; amber <- "#e8861a"; alert <- "#d83a3a"

ggplot() +
  annotate("rect", xmin = silent_a, xmax = silent_b, ymin = -1.2, ymax = 1.2,
           fill = "#c0392b", alpha = 0.05) +
  annotate("text", x = ymd_hms("2046-06-05 16:35:00"), y = 1.08,
           label = "JUDGE SILENT FROM 15:08", colour = "#b06a62",
           size = 3, fontface = "bold") +
  annotate("segment", x = breach, xend = breach, y = -1.2, yend = 1.2,
           colour = alert, linewidth = 0.8, linetype = "dashed") +
  annotate("text", x = breach, y = -1.08, label = "17:25 BREACH",
           colour = alert, size = 3.1, fontface = "bold", hjust = 1.06) +
  annotate("segment", x = silent_b, xend = silent_b, y = -1.2, yend = 1.2,
           colour = "grey60", linewidth = 0.5, linetype = "dotted") +
  annotate("text", x = silent_b, y = 1.08, label = "18:06\nembargo lifts",
           colour = "grey45", size = 2.8, hjust = -0.05, lineheight = 0.9) +
  annotate("segment", x = win_start, xend = win_end, y = 0, yend = 0,
           colour = "grey80", linewidth = 0.6) +
  annotate("text", x = win_start, y = 0.95, label = "THE JUDGE (enforcement)",
           hjust = 0, colour = cyan, size = 3, fontface = "bold") +
  annotate("text", x = win_start, y = -0.95, label = "LEGAL APPEALING TO THE JUDGE",
           hjust = 0, colour = amber, size = 3, fontface = "bold") +
  geom_segment(data = judge_pts, aes(x = time, xend = time, y = 0, yend = y),
               colour = cyan, linewidth = 0.4) +
  geom_point(data = judge_pts, aes(x = time, y = y), colour = cyan, size = 3) +
  geom_point(data = filter(judge_pts, note != ""), aes(x = time, y = y),
             colour = cyan, size = 5) +
  geom_text(data = filter(judge_pts, note != ""),
            aes(x = time, y = y + 0.16, label = note),
            colour = cyan, size = 2.7, lineheight = 0.9, vjust = 0, hjust = 1.05) +
  geom_segment(data = legal_pts, aes(x = time, xend = time, y = 0, yend = y),
               colour = amber, linewidth = 0.4) +
  geom_point(data = legal_pts, aes(x = time, y = y), colour = amber, size = 3) +
  annotate("text", x = ymd_hms("2046-06-05 17:14:00"), y = -0.7,
           label = "Legal fires four justifications\nat a Judge that never replies",
           colour = amber, size = 2.8, lineheight = 0.9, vjust = 1) +
  scale_x_datetime(date_breaks = "1 hour", date_labels = "%H:%M",
                   limits = c(win_start, win_end), expand = expansion(mult = 0.02)) +
  scale_y_continuous(limits = c(-1.3, 1.3)) +
  labs(title = "How It Got Past the Judge",
       subtitle = "The compliance tool's last action was a warning at 15:08. It stayed silent through the 17:25 breach.",
       x = NULL, y = NULL) +
  theme_minimal(base_size = 12) +
  theme(plot.title    = element_text(face = "bold"),
        plot.subtitle = element_text(colour = "grey40", size = 9),
        panel.grid    = element_blank(),
        axis.text.y   = element_blank(),
        legend.position = "none")
```

### How it got past the Judge

The embargo was meant to be enforced by the Judge, an automated compliance tool,
and three gaps let the post through.

First, the Judge was new and lenient. It was installed on 30 May, five days before
the breach, and on its first day it handed routine-post approval back to the
agents themselves. That gave the team a self-approval path for anything it could
frame as routine.

Second, it spotted trouble but did not act. At 15:08 it issued its one compliance
warning of the day, flagging an anonymous post that tied the CivicLoom name to the
6 PM embargo. The warning was never followed up.

Third, it went quiet at the worst moment. That 15:08 warning was its last message
of any kind. Through the entire breach window it said nothing, even as Legal
addressed it directly four times between 17:01 and 17:27, narrating a legal
justification and announcing the release. Legal was making its case to an enforcer
that had stopped listening. By 17:25 the embargo was being guarded by a tool that
had effectively switched off.

## Answer to Question 1

**The breach was set off from outside, but carried out from inside.** SaltWind
published the CivicLoom merger at 17:00, and at that same moment the CEO ordered
the team to hold until 18:00. The key actors inside were Legal, which confirmed
the merger by name at 17:25, and Social-Media, which amplified it a minute later.
The decisive moment was Legal's choice to override a direct, current order to
wait, then justify it after the fact with a contract clause and a consent the CEO
had not given. The system element that let this past the embargo enforcement was
the Judge: installed only days earlier, lenient by design, and silent through the
entire breach window. The honest reading is not a planted leak but an agent
jumping the gun under pressure, inside a system whose one safeguard had
effectively switched off.
