---
title: "Hands-on Exercise 1: A Layered Grammar of Graphics"
author: "ZeXin Chen"
date: "April 26, 2026"
date-modified: "last-modified"
format:
  html:
    toc: true
    code-fold: true
    code-summary: "Show code"
execute:
  echo: true
  eval: true
  warning: false
  message: false
  freeze: true
---

# Getting Started

## Installing and Loading R Packages

The code chunk below uses `p_load()` of **pacman** package to check if tidyverse packages are installed in the computer. If they are, then they will be launched into R.

```{r}
pacman::p_load(tidyverse)
```

## Importing the Data

```{r}
exam_data <- read_csv("data/Exam_data.csv")
```

The dataset contains year-end examination grades of a cohort of Primary 3 students from a local school. It has seven attributes, four categorical (`ID`, `CLASS`, `GENDER`, `RACE`) and three continuous (`MATHS`, `ENGLISH`, `SCIENCE`).

# Geometric Objects

## Bar Chart with `geom_bar()`

A bar chart showing the count of students by race.

```{r}
ggplot(data = exam_data,
       aes(x = RACE)) +
  geom_bar()
```

## Dot Plot with `geom_dotplot()`

A dot plot of Maths scores, where each dot represents one observation.

```{r}
ggplot(data = exam_data,
       aes(x = MATHS)) +
  geom_dotplot(binwidth = 2.5,
               dotsize = 0.5) +
  scale_y_continuous(NULL, breaks = NULL)
```

## Histogram with `geom_histogram()`

A histogram of Maths scores coloured by gender.

```{r}
ggplot(data = exam_data,
       aes(x = MATHS,
           fill = GENDER)) +
  geom_histogram(bins = 20,
                 color = "grey30")
```

## Kernel Density Estimate with `geom_density()`

A kernel density plot comparing Maths score distributions between genders.

```{r}
ggplot(data = exam_data,
       aes(x = MATHS,
           colour = GENDER)) +
  geom_density()
```

## Box Plot with `geom_boxplot()`

A notched box plot of Maths scores by gender. Non-overlapping notches indicate differing medians.

```{r}
ggplot(data = exam_data,
       aes(y = MATHS,
           x = GENDER)) +
  geom_boxplot(notch = TRUE)
```

## Violin Plot with `geom_violin()`

A violin plot comparing the full distribution shape of Maths scores by gender.

```{r}
ggplot(data = exam_data,
       aes(y = MATHS,
           x = GENDER)) +
  geom_violin()
```

## Combining Geoms, Box Plot plus Jitter

Data points overlaid on box plots using `geom_point()` with jitter.

```{r}
ggplot(data = exam_data,
       aes(y = MATHS,
           x = GENDER)) +
  geom_boxplot() +
  geom_point(position = "jitter",
             size = 0.5)
```

# Statistics

## Adding Mean Points with `stat_summary()`

Mean values added to box plots using `stat_summary()`.

```{r}
ggplot(data = exam_data,
       aes(y = MATHS,
           x = GENDER)) +
  geom_boxplot() +
  stat_summary(geom = "point",
               fun = "mean",
               colour = "red",
               size = 4)
```

## Best Fit Line with `geom_smooth()`

A scatter plot of Maths vs English with a linear regression line.

```{r}
ggplot(data = exam_data,
       aes(x = MATHS,
           y = ENGLISH)) +
  geom_point() +
  geom_smooth(method = lm,
              linewidth = 0.5)
```

# Facets

## Trellis Plot with `facet_wrap()`

Histograms of Maths scores faceted by class.

```{r}
ggplot(data = exam_data,
       aes(x = MATHS)) +
  geom_histogram(bins = 20) +
  facet_wrap(~ CLASS)
```

# Coordinates & Themes

## Flipped Bar Chart with `coord_flip()`

A horizontal bar chart of race counts with `theme_minimal()`.

```{r}
ggplot(data = exam_data,
       aes(x = RACE)) +
  geom_bar() +
  coord_flip() +
  theme_minimal()
```

## Fixed Axis Range with `coord_cartesian()`

Scatter plot with both axes fixed from 0 to 100.

```{r}
ggplot(data = exam_data,
       aes(x = MATHS,
           y = ENGLISH)) +
  geom_point() +
  geom_smooth(method = lm,
              linewidth = 0.5) +
  coord_cartesian(xlim = c(0, 100),
                  ylim = c(0, 100))
```