---
title: "Hands-on Exercise 10"
subtitle: "Information Dashboard Design"
author: "ZeXin Chen"
date: "2026-06-22"
date-modified: "last-modified"
format:
  html:
    toc: true
    number-sections: true
execute:
  echo: true
  eval: true
  warning: false
  message: false
  freeze: true
---

# Overview

In this hands-on exercise we learn to design the building blocks of an
information dashboard in R. We cover three things:

- bullet charts and sparklines drawn with **ggplot2**
- static dashboard tables built with **gt** and **gtExtras**
- interactive dashboard tables built with **reactable** and **reactablefmtr**

A dashboard packs a lot of numbers into a small space, so the design principle
throughout is restraint. We use plain backgrounds, a single highlight colour for
the value that matters, and grey for everything that is context.

# Getting Started

## Installing and Loading Packages

```{r}
pacman::p_load(lubridate, ggthemes, reactable,
               reactablefmtr, gt, gtExtras, tidyverse)
```

The interactive sparklines in the last section need the **dataui** package, which
lives on GitHub rather than CRAN. Install it once with the line below, then it
loads with the others.

```{r}
#| eval: false
remotes::install_github("timelyportfolio/dataui")
```

```{r}
library(dataui)
```

# The Data

The data is a personal sales database for a coffee chain. The original file is in
Microsoft Access format, which needs the RODBC package and a 32 bit R session to
read. To avoid that, we use the prepared `CoffeeChain.rds` file, which holds the
same query table already converted to an R data frame.

```{r}
coffeechain <- read_rds("data/rds/CoffeeChain.rds")
```

```{r}
glimpse(coffeechain)
```

# Bullet Chart in ggplot2

A bullet chart compares an actual value against a target, using shaded bands for
context. It is a compact alternative to a gauge. First we aggregate sales and
budgeted sales by product.

```{r}
product <- coffeechain %>%
  group_by(`Product`) %>%
  summarise(`target` = sum(`Budget Sales`),
            `current` = sum(`Sales`)) %>%
  ungroup()
```

The chart layers grey context bands behind a black measure bar, with a red marker
for the target. Colour is used sparingly here: grey for the qualitative ranges,
black for the actual value, and a single red line for the target.

::: panel-tabset
## The Plot
```{r}
#| echo: false
ggplot(product, aes(Product, current)) +
  geom_col(aes(Product, max(target) * 1.01),
           fill = "grey85", width = 0.85) +
  geom_col(aes(Product, target * 0.75),
           fill = "grey60", width = 0.85) +
  geom_col(aes(Product, target * 0.5),
           fill = "grey50", width = 0.85) +
  geom_col(aes(Product, current),
           width = 0.35,
           fill = "black") +
  geom_errorbar(aes(y = target,
                    x = Product,
                    ymin = target,
                    ymax = target),
                width = .4,
                colour = "red",
                size = 1) +
  coord_flip()
```
## The Code
```{r}
#| eval: false
ggplot(product, aes(Product, current)) +
  geom_col(aes(Product, max(target) * 1.01),
           fill = "grey85", width = 0.85) +
  geom_col(aes(Product, target * 0.75),
           fill = "grey60", width = 0.85) +
  geom_col(aes(Product, target * 0.5),
           fill = "grey50", width = 0.85) +
  geom_col(aes(Product, current),
           width = 0.35,
           fill = "black") +
  geom_errorbar(aes(y = target,
                    x = Product,
                    ymin = target,
                    ymax = target),
                width = .4,
                colour = "red",
                size = 1) +
  coord_flip()
```
:::

# Sparklines in ggplot2

A sparkline is a small line chart with no axes, meant to show the shape of a
trend at a glance. We build one per product, faceted into a stack.

## Preparing the Data

```{r}
sales_report <- coffeechain %>%
  filter(Date >= "2013-01-01") %>%
  mutate(Month = month(Date)) %>%
  group_by(Month, Product) %>%
  summarise(Sales = sum(Sales)) %>%
  ungroup() %>%
  select(Month, Product, Sales)
```

We mark the lowest month, the highest month, and the final month so each
sparkline can be annotated.

```{r}
mins <- group_by(sales_report, Product) %>%
  slice(which.min(Sales))
maxs <- group_by(sales_report, Product) %>%
  slice(which.max(Sales))
ends <- group_by(sales_report, Product) %>%
  filter(Month == max(Month))
```

We also compute the 25th and 75th percentiles so we can shade the interquartile
band behind each line.

```{r}
quarts <- sales_report %>%
  group_by(Product) %>%
  summarise(quart1 = quantile(Sales, 0.25),
            quart2 = quantile(Sales, 0.75)) %>%
  right_join(sales_report)
```

## Drawing the Sparklines

The grey band shows the normal range, the line shows the trend, the red point
marks the lowest month and the blue point the highest. The Tufte theme strips
away everything that is not data.

::: panel-tabset
## The Plot
```{r}
#| echo: false
#| fig-height: 8
ggplot(sales_report, aes(x = Month, y = Sales)) +
  facet_grid(Product ~ ., scales = "free_y") +
  geom_ribbon(data = quarts, aes(ymin = quart1, max = quart2),
              fill = 'grey90') +
  geom_line(size = 0.3) +
  geom_point(data = mins, col = 'red') +
  geom_point(data = maxs, col = 'blue') +
  geom_text(data = mins, aes(label = Sales), vjust = -1) +
  geom_text(data = maxs, aes(label = Sales), vjust = 2.5) +
  geom_text(data = ends, aes(label = Sales), hjust = 0, nudge_x = 0.5) +
  geom_text(data = ends, aes(label = Product), hjust = 0, nudge_x = 1.0) +
  expand_limits(x = max(sales_report$Month) +
                  (0.25 * (max(sales_report$Month) - min(sales_report$Month)))) +
  scale_x_continuous(breaks = seq(1, 12, 1)) +
  scale_y_continuous(expand = c(0.1, 0)) +
  theme_tufte(base_size = 3, base_family = "Helvetica") +
  theme(axis.title = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks = element_blank(),
        strip.text = element_blank())
```
## The Code
```{r}
#| eval: false
ggplot(sales_report, aes(x = Month, y = Sales)) +
  facet_grid(Product ~ ., scales = "free_y") +
  geom_ribbon(data = quarts, aes(ymin = quart1, max = quart2),
              fill = 'grey90') +
  geom_line(size = 0.3) +
  geom_point(data = mins, col = 'red') +
  geom_point(data = maxs, col = 'blue') +
  geom_text(data = mins, aes(label = Sales), vjust = -1) +
  geom_text(data = maxs, aes(label = Sales), vjust = 2.5) +
  geom_text(data = ends, aes(label = Sales), hjust = 0, nudge_x = 0.5) +
  geom_text(data = ends, aes(label = Product), hjust = 0, nudge_x = 1.0) +
  expand_limits(x = max(sales_report$Month) +
                  (0.25 * (max(sales_report$Month) - min(sales_report$Month)))) +
  scale_x_continuous(breaks = seq(1, 12, 1)) +
  scale_y_continuous(expand = c(0.1, 0)) +
  theme_tufte(base_size = 3, base_family = "Helvetica") +
  theme(axis.title = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks = element_blank(),
        strip.text = element_blank())
```
:::

# Static Dashboard with gt and gtExtras

The **gt** package builds presentation tables, and **gtExtras** adds dashboard
elements such as inline bullet charts and sparklines. These tables are the static
building blocks of a report dashboard.

## Simple Bullet Chart Table

`gt_plt_bullet()` draws an inline bullet chart in each row, comparing the current
value against its target. The light blue bar is the actual value, the black line
the target.

```{r}
product %>%
  gt::gt() %>%
  gt_plt_bullet(column = current,
                target = target,
                width = 60,
                palette = c("lightblue", "black")) %>%
  gt_theme_538()
```

## Preparing the Monthly Report

For sparklines in a table, gtExtras needs the monthly sales for each product
collected into a list column.

```{r}
report <- coffeechain %>%
  mutate(Year = year(Date)) %>%
  filter(Year == "2013") %>%
  mutate(Month = month(Date,
                       label = TRUE,
                       abbr = TRUE)) %>%
  group_by(Product, Month) %>%
  summarise(Sales = sum(Sales)) %>%
  ungroup()
```

## Sparkline Table

`gt_plt_sparkline()` turns each product's list of monthly sales into an inline
sparkline.

```{r}
report %>%
  group_by(Product) %>%
  summarize('Monthly Sales' = list(Sales),
            .groups = "drop") %>%
  gt() %>%
  gt_plt_sparkline('Monthly Sales',
                   same_limit = FALSE)
```

## Adding Summary Statistics

We can place summary numbers next to the sparkline. First the spark list column,
then the min, max and average, then join them.

```{r}
spark <- report %>%
  group_by(Product) %>%
  summarize('Monthly Sales' = list(Sales),
            .groups = "drop")
```

```{r}
sales <- report %>%
  group_by(Product) %>%
  summarise("Min" = min(Sales, na.rm = T),
            "Max" = max(Sales, na.rm = T),
            "Average" = mean(Sales, na.rm = T))
```

```{r}
sales_data <- left_join(sales, spark)
```

```{r}
sales_data %>%
  gt() %>%
  gt_plt_sparkline('Monthly Sales',
                   same_limit = FALSE)
```

## Combining Bullet Chart and Sparklines

A full dashboard row can carry both the trend and the target comparison. We add
the target and actual totals, then draw a sparkline and a bullet chart side by
side.

```{r}
bullet <- coffeechain %>%
  filter(Date >= "2013-01-01") %>%
  group_by(`Product`) %>%
  summarise(`Target` = sum(`Budget Sales`),
            `Actual` = sum(`Sales`)) %>%
  ungroup()
```

```{r}
sales_data <- sales_data %>%
  left_join(bullet)
```

```{r}
sales_data %>%
  gt() %>%
  gt_plt_sparkline('Monthly Sales') %>%
  gt_plt_bullet(column = Actual,
                target = Target,
                width = 28,
                palette = c("lightblue", "black")) %>%
  gt_theme_538()
```

# Interactive Dashboard with reactable and reactablefmtr

The **reactable** package builds interactive tables, and **reactablefmtr** adds
interactive sparklines that respond to hovering. These are the building blocks of
a live dashboard.

## Preparing the List Column

Like gtExtras, reactablefmtr needs the monthly sales as a list column.

```{r}
report <- report %>%
  group_by(Product) %>%
  summarize(`Monthly Sales` = list(Sales))
```

## Basic Interactive Sparklines

`react_sparkline()` draws an interactive sparkline in each row. Hovering over a
line reveals the underlying values.

```{r}
reactable(
  report,
  columns = list(
    Product = colDef(maxWidth = 200),
    `Monthly Sales` = colDef(
      cell = react_sparkline(report)
    )
  )
)
```

## Showing All Rows

By default the table pages at 10 rows. Setting `defaultPageSize` to 13 shows
every product on one page.

```{r}
reactable(
  report,
  defaultPageSize = 13,
  columns = list(
    Product = colDef(maxWidth = 200),
    `Monthly Sales` = colDef(
      cell = react_sparkline(report)
    )
  )
)
```

## Adding Points and Labels

`highlight_points` marks the lowest and highest months in red and blue, and
`labels` prints the first and last values.

```{r}
reactable(
  report,
  defaultPageSize = 13,
  columns = list(
    Product = colDef(maxWidth = 200),
    `Monthly Sales` = colDef(
      cell = react_sparkline(
        report,
        highlight_points = highlight_points(
          min = "red", max = "blue"),
        labels = c("first", "last")
      )
    )
  )
)
```

## Adding a Reference Line

`statline` draws a reference line, here the mean of each product's monthly sales.

```{r}
reactable(
  report,
  defaultPageSize = 13,
  columns = list(
    Product = colDef(maxWidth = 200),
    `Monthly Sales` = colDef(
      cell = react_sparkline(
        report,
        highlight_points = highlight_points(
          min = "red", max = "blue"),
        statline = "mean"
      )
    )
  )
)
```

## Adding a Bandline

Instead of a reference line, a band can show the interquartile range behind the
sparkline.

```{r}
reactable(
  report,
  defaultPageSize = 13,
  columns = list(
    Product = colDef(maxWidth = 200),
    `Monthly Sales` = colDef(
      cell = react_sparkline(
        report,
        highlight_points = highlight_points(
          min = "red", max = "blue"),
        line_width = 1,
        bandline = "innerquartiles",
        bandline_color = "green"
      )
    )
  )
)
```

## Sparkbars Instead of Sparklines

The same data can be shown as small bars rather than a line, using
`react_sparkbar()`.

```{r}
reactable(
  report,
  defaultPageSize = 13,
  columns = list(
    Product = colDef(maxWidth = 200),
    `Monthly Sales` = colDef(
      cell = react_sparkbar(
        report,
        highlight_bars = highlight_bars(
          min = "red", max = "blue"),
        bandline = "innerquartiles",
        statline = "mean")
    )
  )
)
```
