---
title: "Hands-on Exercise 6"
subtitle: "Visualising and Analysing Network Data & Text Data"
author: "ZeXin Chen"
date: "2026-06-02"
date-modified: last-modified
format:
  html:
    theme: flatly
    toc: true
    code-fold: true
    code-summary: "Show code"
    number-sections: true
execute:
  echo: true
  eval: true
  warning: false
  message: false
  freeze: true
---

# Part A: Visualising and Analysing Network Data

In this section, we model, analyse and visualise network data using R. We build
graph objects with **tidygraph**, draw static network graphs with **ggraph**,
compute network metrics, and finally build an interactive network graph with
**visNetwork**.

## Getting Started

### Installing and Loading Packages

```{r}
pacman::p_load(igraph, tidygraph, ggraph,
               visNetwork, lubridate, clock,
               tidyverse, graphlayouts,
               concaveman, ggforce)
```

## The Data

The data is from an oil exploration and extraction company (GAStech). There are
two files: a **nodes** file (the 55 employees) and an **edges** file (the email
correspondences between them).

### Importing the Network Data

```{r}
GAStech_nodes <- read_csv("data/GAStech_email_node.csv")
GAStech_edges <- read_csv("data/GAStech_email_edge-v2.csv")
```

### Reviewing the Imported Data

```{r}
glimpse(GAStech_edges)
```

The `SentDate` field is read as a *character* type instead of a *date*. We fix
this before continuing.

### Wrangling Time

```{r}
GAStech_edges <- GAStech_edges %>%
  mutate(SendDate = dmy(SentDate)) %>%
  mutate(Weekday = wday(SentDate,
                        label = TRUE,
                        abbr = FALSE))
```

Here `dmy()` converts `SentDate` to a proper Date type, and `wday()` extracts the
day of the week as an ordered factor stored in a new `Weekday` column.

### Wrangling Attributes

The edges file contains one row per individual email, which is not very useful
for visualisation. We aggregate by sender, receiver and day of the week, keeping
only work-related emails and pairs with more than one email.

```{r}
GAStech_edges_aggregated <- GAStech_edges %>%
  filter(MainSubject == "Work related") %>%
  group_by(source, target, Weekday) %>%
    summarise(Weight = n()) %>%
  filter(source != target) %>%
  filter(Weight > 1) %>%
  ungroup()
```

## Creating Network Objects using tidygraph

**tidygraph** provides a tidy way to handle network data as two tables: one for
nodes and one for edges.

### Building the tbl_graph Object

```{r}
GAStech_graph <- tbl_graph(nodes = GAStech_nodes,
                           edges = GAStech_edges_aggregated,
                           directed = TRUE)
```

```{r}
GAStech_graph
```

## Plotting Static Network Graphs with ggraph

**ggraph** is an extension of **ggplot2** for network graphs. A ggraph plot has
three main parts: nodes, edges and layouts.

### Basic Network Graph

```{r}
ggraph(GAStech_graph) +
  geom_edge_link() +
  geom_node_point()
```

### Changing the Default Theme

`theme_graph()` removes axes, grids and borders for a cleaner network look.

```{r}
g <- ggraph(GAStech_graph) +
  geom_edge_link(aes()) +
  geom_node_point(aes())

g + theme_graph()
```

### Fruchterman-Reingold Layout

The `layout` argument controls how the nodes are positioned. Here we use the
popular Fruchterman-Reingold (`fr`) force-directed layout.

```{r}
g <- ggraph(GAStech_graph,
            layout = "fr") +
  geom_edge_link(aes()) +
  geom_node_point(aes())

g + theme_graph()
```

### Colouring Nodes by Department

Here colour encodes a meaningful variable — each node is coloured by the
employee's department.

```{r}
g <- ggraph(GAStech_graph,
            layout = "nicely") +
  geom_edge_link(aes()) +
  geom_node_point(aes(colour = Department,
                      size = 3))

g + theme_graph()
```

### Mapping Edge Weight to Thickness

The width of each edge is mapped to the `Weight` variable (number of emails),
and opacity is reduced to ease over-plotting.

```{r}
g <- ggraph(GAStech_graph,
            layout = "nicely") +
  geom_edge_link(aes(width = Weight),
                 alpha = 0.2) +
  scale_edge_width(range = c(0.1, 5)) +
  geom_node_point(aes(colour = Department),
                  size = 3)

g + theme_graph()
```

## Creating Facet Graphs

Faceting spreads nodes and edges out by an attribute, reducing over-plotting in a
meaningful way.

### Faceting by Weekday with facet_edges()

```{r}
set_graph_style()

g <- ggraph(GAStech_graph,
            layout = "nicely") +
  geom_edge_link(aes(width = Weight),
                 alpha = 0.2) +
  scale_edge_width(range = c(0.1, 5)) +
  geom_node_point(aes(colour = Department),
                  size = 2) +
  theme(legend.position = 'bottom')

g + facet_edges(~Weekday)
```

### Framed Facet Graph

```{r}
set_graph_style()

g <- ggraph(GAStech_graph,
            layout = "nicely") +
  geom_edge_link(aes(width = Weight),
                 alpha = 0.2) +
  scale_edge_width(range = c(0.1, 5)) +
  geom_node_point(aes(colour = Department),
                  size = 2)

g + facet_edges(~Weekday) +
  th_foreground(foreground = "grey80",
                border = TRUE) +
  theme(legend.position = 'bottom')
```

### Faceting by Department with facet_nodes()

```{r}
set_graph_style()

g <- ggraph(GAStech_graph,
            layout = "nicely") +
  geom_edge_link(aes(width = Weight),
                 alpha = 0.2) +
  scale_edge_width(range = c(0.1, 5)) +
  geom_node_point(aes(colour = Department),
                  size = 2)

g + facet_nodes(~Department) +
  th_foreground(foreground = "grey80",
                border = TRUE) +
  theme(legend.position = 'bottom')
```

## Network Metrics Analysis

### Computing and Visualising Betweenness Centrality

Betweenness centrality measures how often a node lies on the shortest path
between other nodes. From ggraph v2.0 onward, tidygraph algorithms can be called
directly inside ggraph.

```{r}
g <- GAStech_graph %>%
  ggraph(layout = "fr") +
  geom_edge_link(aes(width = Weight),
                 alpha = 0.2) +
  scale_edge_width(range = c(0.1, 5)) +
  geom_node_point(aes(colour = Department,
                      size = centrality_betweenness()))

g + theme_graph()
```

### Visualising Community

Community detection groups nodes that are more densely connected to each other.
Here `group_edge_betweenness()` is used to detect communities.

```{r}
g <- GAStech_graph %>%
  mutate(community = as.factor(
    group_edge_betweenness(
      weights = Weight,
      directed = TRUE))) %>%
  ggraph(layout = "fr") +
  geom_edge_link(
    aes(width = Weight),
    alpha = 0.2) +
  scale_edge_width(
    range = c(0.1, 5)) +
  geom_node_point(
    aes(colour = community))

g + theme_graph()
```

We can make the communities clearer using `geom_mark_hull()` of the **ggforce**
package to draw a hull around each community.

```{r}
g <- GAStech_graph %>%
  activate(nodes) %>%
  mutate(community = as.factor(
    group_optimal(weights = Weight)),
         betweenness_measure = centrality_betweenness()) %>%
  ggraph(layout = "fr") +
  geom_mark_hull(
    aes(x, y,
        group = community,
        fill = community),
    alpha = 0.2,
    expand = unit(0.3, "cm"),
    radius = unit(0.3, "cm")
  ) +
  geom_edge_link(aes(width = Weight),
                 alpha = 0.2) +
  scale_edge_width(range = c(0.1, 5)) +
  geom_node_point(aes(fill = Department,
                      size = betweenness_measure),
                      color = "black",
                      shape = 21)

g + theme_graph()
```

## Building Interactive Network Graph with visNetwork

**visNetwork** uses a nodes list and an edges list to create an interactive
graph that you can drag, zoom and explore.

### Data Preparation

```{r}
GAStech_edges_aggregated <- GAStech_edges %>%
  left_join(GAStech_nodes, by = c("sourceLabel" = "label")) %>%
  rename(from = id) %>%
  left_join(GAStech_nodes, by = c("targetLabel" = "label")) %>%
  rename(to = id) %>%
  filter(MainSubject == "Work related") %>%
  group_by(from, to) %>%
    summarise(weight = n()) %>%
  filter(from != to) %>%
  filter(weight > 1) %>%
  ungroup()
```

### Basic Interactive Network Graph

The basic call has no layout specified, so visNetwork runs its live
force-directed physics engine. With this many nodes and edges the simulation
keeps the nodes moving, so we add `visPhysics(stabilization = TRUE)` to run the
simulation once and let it settle into a stable position instead of jittering
continuously.

```{r}
visNetwork(GAStech_nodes,
           GAStech_edges_aggregated) %>%
  visPhysics(stabilization = TRUE)
```

### Working with Layout

A cleaner approach is to pre-compute a fixed layout with igraph using
`visIgraphLayout()`. This turns the live physics off entirely, so nodes only move
when you drag them.

```{r}
visNetwork(GAStech_nodes,
           GAStech_edges_aggregated) %>%
  visIgraphLayout(layout = "layout_with_fr")
```

### Colouring Nodes by Group

visNetwork looks for a field called `group` and colours nodes by its values. We
rename `Department` to `group`.

```{r}
GAStech_nodes <- GAStech_nodes %>%
  rename(group = Department)
```

```{r}
visNetwork(GAStech_nodes,
           GAStech_edges_aggregated) %>%
  visIgraphLayout(layout = "layout_with_fr") %>%
  visLegend() %>%
  visLayout(randomSeed = 123)
```

### Adding Edges and Interactivity

`visEdges()` styles the edges, and `visOptions()` adds interactivity such as
highlighting the nearest nodes on click and a dropdown to select a node by id.

```{r}
visNetwork(GAStech_nodes,
           GAStech_edges_aggregated) %>%
  visIgraphLayout(layout = "layout_with_fr") %>%
  visEdges(arrows = "to",
           smooth = list(enabled = TRUE,
                         type = "curvedCW")) %>%
  visOptions(highlightNearest = TRUE,
             nodesIdSelection = TRUE) %>%
  visLegend() %>%
  visLayout(randomSeed = 123)
```

# Part B: Visualising and Analysing Text Data

In this section, we process and visualise text data using the **tidytext**
framework. We import many text files, clean them, build a word cloud, compute
TF-IDF, and visualise word relationships as a network.

## Getting Started

### Installing and Loading Packages

```{r}
pacman::p_load(tidytext, widyr, wordcloud, DT,
               ggwordcloud, textplot, lubridate, hms,
               tidyverse, tidygraph, ggraph, igraph)
```

## Importing Multiple Text Files from Multiple Folders

### Creating a Folder List

```{r}
news20 <- "data/20news/"
```

### Defining a Function to Read a Folder

```{r}
read_folder <- function(infolder) {
  tibble(file = dir(infolder,
                    full.names = TRUE)) %>%
    mutate(text = map(file,
                      read_lines)) %>%
    transmute(id = basename(file),
              text) %>%
    unnest(text)
}
```

### Reading All Messages

```{r}
raw_text <- tibble(folder =
                     dir(news20,
                         full.names = TRUE)) %>%
  mutate(folder_out = map(folder,
                          read_folder)) %>%
  unnest(cols = c(folder_out)) %>%
  transmute(newsgroup = basename(folder),
            id, text)
```

## Initial EDA

The chart below shows the number of messages by newsgroup. The bars are sorted by
count so the differences between newsgroups read clearly.

```{r}
raw_text %>%
  group_by(newsgroup) %>%
  summarize(messages = n_distinct(id)) %>%
  ggplot(aes(messages, reorder(newsgroup, messages))) +
  geom_col(fill = "lightblue") +
  labs(y = NULL)
```

## Introducing tidytext

### Removing Headers and Signatures

Each message has a header and sometimes an automated signature that we don't want
in the analysis.

```{r}
cleaned_text <- raw_text %>%
  group_by(newsgroup, id) %>%
  filter(cumsum(text == "") > 0,
         cumsum(str_detect(
           text, "^--")) == 0) %>%
  ungroup()
```

### Removing Quoted Lines

```{r}
cleaned_text <- cleaned_text %>%
  filter(str_detect(text, "^[^>]+[A-Za-z\\d]")
         | text == "",
         !str_detect(text,
                     "writes(:|\\.\\.\\.)$"),
         !str_detect(text,
                     "^In article <")
  )
```

### Tokenising and Removing Stop Words

`unnest_tokens()` splits the text into individual words, and we remove common
stop words.

```{r}
usenet_words <- cleaned_text %>%
  unnest_tokens(word, text) %>%
  filter(str_detect(word, "[a-z']$"),
         !word %in% stop_words$word)
```

```{r}
usenet_words %>%
  count(word, sort = TRUE)
```

### Counting Words by Newsgroup

```{r}
words_by_newsgroup <- usenet_words %>%
  count(newsgroup, word, sort = TRUE) %>%
  ungroup()
```

## Visualising Words with Word Cloud

### Static Word Cloud with wordcloud

```{r}
wordcloud(words_by_newsgroup$word,
          words_by_newsgroup$n,
          max.words = 300)
```

### Interactive Table of Word Counts

```{r}
DT::datatable(words_by_newsgroup, filter = 'top')
```

### Word Cloud with ggwordcloud

```{r}
set.seed(1234)

words_by_newsgroup %>%
  group_by(newsgroup) %>%
  slice_max(n, n = 30) %>%
  ungroup() %>%
  ggplot(aes(label = word,
             size = n)) +
  geom_text_wordcloud() +
  scale_size_area(max_size = 6) +
  theme_minimal() +
  facet_wrap(~newsgroup)
```

## TF-IDF Analysis

TF-IDF (term frequency–inverse document frequency) reflects how important a word
is to a document within a collection.

### Computing TF-IDF

```{r}
tf_idf <- words_by_newsgroup %>%
  bind_tf_idf(word, newsgroup, n) %>%
  arrange(desc(tf_idf))
```

### Visualising TF-IDF as an Interactive Table

```{r}
DT::datatable(tf_idf, filter = 'top') %>%
  formatRound(columns = c('tf', 'idf',
                          'tf_idf'),
              digits = 3) %>%
  formatStyle(0,
              target = 'row',
              lineHeight = '25%')
```

### Visualising TF-IDF within Science Newsgroups

Facet bar charts show the top TF-IDF words for each science-related newsgroup.

```{r}
tf_idf %>%
  filter(str_detect(newsgroup, "^sci\\.")) %>%
  group_by(newsgroup) %>%
  slice_max(tf_idf,
            n = 12) %>%
  ungroup() %>%
  mutate(word = reorder(word,
                        tf_idf)) %>%
  ggplot(aes(tf_idf,
             word,
             fill = newsgroup)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~ newsgroup,
             scales = "free") +
  labs(x = "tf-idf",
       y = NULL)
```

## Counting and Correlating Pairs of Words

### Pairwise Correlation between Newsgroups

`pairwise_cor()` of **widyr** computes the correlation between newsgroups based on
common words.

```{r}
newsgroup_cors <- words_by_newsgroup %>%
  filter(n >= 5) %>%
  pairwise_cor(newsgroup,
               word,
               n,
               sort = TRUE)
```

### Visualising Correlation as a Network

```{r}
set.seed(2017)

newsgroup_cors %>%
  filter(correlation > .025) %>%
  graph_from_data_frame() %>%
  ggraph(layout = "fr") +
  geom_edge_link(aes(alpha = correlation,
                     width = correlation)) +
  geom_node_point(size = 6,
                  color = "lightblue") +
  geom_node_text(aes(label = name),
                 color = "red",
                 repel = TRUE) +
  theme_void()
```

## Bigram Analysis

### Creating Bigrams

A bigram is a pair of consecutive words.

```{r}
bigrams <- cleaned_text %>%
  unnest_tokens(bigram,
                text,
                token = "ngrams",
                n = 2)
```

### Cleaning Bigrams

We separate each bigram into two words and remove stop words from both.

```{r}
bigrams_separated <- bigrams %>%
  filter(bigram != 'NA') %>%
  separate(bigram, c("word1", "word2"),
           sep = " ")

bigrams_filtered <- bigrams_separated %>%
  filter(!word1 %in% stop_words$word) %>%
  filter(!word2 %in% stop_words$word)
```

### Counting Bigrams

```{r}
bigram_counts <- bigrams_filtered %>%
  count(word1, word2, sort = TRUE)
```

### Creating a Network Graph from Bigrams

```{r}
bigram_graph <- bigram_counts %>%
  filter(n > 3) %>%
  graph_from_data_frame()

bigram_graph
```

### Visualising the Bigram Network

```{r}
set.seed(1234)

a <- grid::arrow(type = "closed",
                 length = unit(.15,
                               "inches"))

ggraph(bigram_graph,
       layout = "fr") +
  geom_edge_link(aes(edge_alpha = n),
                 show.legend = FALSE,
                 arrow = a,
                 end_cap = circle(.07,
                                  'inches')) +
  geom_node_point(color = "lightblue",
                  size = 5) +
  geom_node_text(aes(label = name),
                 vjust = 1,
                 hjust = 1) +
  theme_void()
```
