---
title: "Hands-on Exercise 7"
subtitle: "Visualising and Analysing Time-oriented Data"
author: "ZeXin Chen"
date: "2026-06-09"
date-modified: last-modified
format:
  html:
    theme: flatly
    toc: true
    code-fold: true
    code-summary: "Show code"
    number-sections: true
execute:
  echo: true
  eval: true
  warning: false
  message: false
  freeze: true
---

# Overview

In this hands-on exercise we learn to create several time-oriented data
visualisations using R:

- a **calendar heatmap** using ggplot2
- a **cycle plot** using ggplot2
- a **slopegraph** using the CGPfunctions package
- a **horizon chart** using the ggHoriPlot package

# Getting Started

## Installing and Loading Packages

```{r}
pacman::p_load(scales, viridis, lubridate, ggthemes,
               gridExtra, readxl, knitr, data.table,
               CGPfunctions, ggHoriPlot, tidyverse)
```

# Plotting Calendar Heatmap

A calendar heatmap shows the intensity of an event across two time dimensions,
here the weekday and the hour of day.

## The Data

We use `eventlog.csv`, which holds 199,999 time-series cyber attack records by
country. There are three columns: `timestamp` (date-time), `source_country` (ISO
country code), and `tz` (time zone).

## Importing the Data

```{r}
attacks <- read_csv("data/eventlog.csv")
```

## Examining the Data Structure

It is good practice to examine the imported data frame before further analysis.
`kable()` gives a tidy preview.

```{r}
kable(head(attacks))
```

## Data Preparation

### Deriving Weekday and Hour Fields

Before plotting we derive two new fields, `wkday` and `hour`. We write a function
to do this, converting each timestamp to its real local time first.

```{r}
make_hr_wkday <- function(ts, sc, tz) {
  real_times <- ymd_hms(ts,
                        tz = tz[1],
                        quiet = TRUE)
  dt <- data.table(source_country = sc,
                   wkday = weekdays(real_times),
                   hour = hour(real_times))
  return(dt)
}
```

### Building the Cleaned Tibble

We order the weekday and hour fields as factors so they plot in a sensible
order.

```{r}
wkday_levels <- c('Saturday', 'Friday',
                  'Thursday', 'Wednesday',
                  'Tuesday', 'Monday',
                  'Sunday')

attacks <- attacks %>%
  group_by(tz) %>%
  do(make_hr_wkday(.$timestamp,
                   .$source_country,
                   .$tz)) %>%
  ungroup() %>%
  mutate(wkday = factor(
    wkday, levels = wkday_levels),
    hour  = factor(
      hour, levels = 0:23))
```

```{r}
kable(head(attacks))
```

## Building the Calendar Heatmap

We aggregate the attacks by weekday and hour, then plot a tile for each
combination. Colour here encodes the number of attacks, which is the whole point
of the heatmap.

```{r}
grouped <- attacks %>%
  count(wkday, hour) %>%
  ungroup() %>%
  na.omit()

ggplot(grouped,
       aes(hour,
           wkday,
           fill = n)) +
geom_tile(color = "white",
          size = 0.1) +
theme_tufte(base_family = "Helvetica") +
coord_equal() +
scale_fill_gradient(name = "# of attacks",
                    low = "sky blue",
                    high = "dark blue") +
labs(x = NULL,
     y = NULL,
     title = "Attacks by weekday and time of day") +
theme(axis.ticks = element_blank(),
      plot.title = element_text(hjust = 0.5),
      legend.title = element_text(size = 8),
      legend.text = element_text(size = 6))
```

## Building Multiple Calendar Heatmaps

We can build one heatmap per country for the four countries with the most
attacks.

### Identifying the Top Four Countries

```{r}
attacks_by_country <- count(
  attacks, source_country) %>%
  mutate(percent = percent(n/sum(n))) %>%
  arrange(desc(n))
```

### Preparing the Tidy Data Frame

```{r}
top4 <- attacks_by_country$source_country[1:4]
top4_attacks <- attacks %>%
  filter(source_country %in% top4) %>%
  count(source_country, wkday, hour) %>%
  ungroup() %>%
  mutate(source_country = factor(
    source_country, levels = top4)) %>%
  na.omit()
```

### Plotting the Faceted Heatmap

```{r}
ggplot(top4_attacks,
       aes(hour,
           wkday,
           fill = n)) +
  geom_tile(color = "white",
          size = 0.1) +
  theme_tufte(base_family = "Helvetica") +
  coord_equal() +
  scale_fill_gradient(name = "# of attacks",
                    low = "sky blue",
                    high = "dark blue") +
  facet_wrap(~source_country, ncol = 2) +
  labs(x = NULL, y = NULL,
     title = "Attacks on top 4 countries by weekday and time of day") +
  theme(axis.ticks = element_blank(),
        axis.text.x = element_text(size = 7),
        plot.title = element_text(hjust = 0.5),
        legend.title = element_text(size = 8),
        legend.text = element_text(size = 6))
```

# Plotting Cycle Plot

A cycle plot shows a time-series pattern and trend within and across a repeating
cycle, here visitor arrivals from Vietnam by month across years.

## Data Import

```{r}
air <- read_excel("data/arrivals_by_air.xlsx")
```

## Deriving Month and Year Fields

```{r}
air$month <- factor(month(air$`Month-Year`),
                    levels = 1:12,
                    labels = month.abb,
                    ordered = TRUE)
air$year <- year(ymd(air$`Month-Year`))
```

## Extracting the Target Country

```{r}
Vietnam <- air %>%
  select(`Vietnam`,
         month,
         year) %>%
  filter(year >= 2010)
```

## Computing Year Average Arrivals by Month

```{r}
hline.data <- Vietnam %>%
  group_by(month) %>%
  summarise(avgvalue = mean(`Vietnam`))
```

## Plotting the Cycle Plot

Each panel is one month, the black line shows the year-on-year trend, and the red
reference line marks that month's average. Colour is used only for the reference
line, to separate it from the data.

```{r}
ggplot() +
  geom_line(data = Vietnam,
            aes(x = year,
                y = `Vietnam`,
                group = month),
            colour = "black") +
  geom_hline(aes(yintercept = avgvalue),
             data = hline.data,
             linetype = 6,
             colour = "red",
             size = 0.5) +
  facet_grid(~month) +
  labs(axis.text.x = element_blank(),
       title = "Visitor arrivals from Vietnam by air, Jan 2010-Dec 2019") +
  xlab("") +
  ylab("No. of Visitors") +
  theme_tufte(base_family = "Helvetica")
```

# Plotting Slopegraph

A slopegraph compares values at two points in time and shows how each item moves
between them.

## Data Import

```{r}
rice <- read_csv("data/rice.csv")
```

## Plotting the Slopegraph

We convert `Year` to a factor and keep only 1961 and 1980, then use
`newggslopegraph()` of the CGPfunctions package.

```{r}
rice %>%
  mutate(Year = factor(Year)) %>%
  filter(Year %in% c(1961, 1980)) %>%
  newggslopegraph(Year, Yield, Country,
                Title = "Rice Yield of Top 11 Asian Counties",
                SubTitle = "1961-1980",
                Caption = "Prepared by: ZeXin Chen")
```

# Plotting Horizon Chart

A horizon chart is a compact way to compare many time-series at once. Each series
is split into colour bands by value, then the bands are layered so each series
takes very little vertical space. This makes it easy to scan many countries
together.

## Preparing the Data

We reshape the air arrivals data into a long format and derive a proper date
field, keeping data from 2010 onward.

```{r}
air_long <- air %>%
  pivot_longer(cols = -c(`Month-Year`, month, year),
               names_to = "country",
               values_to = "arrivals") %>%
  mutate(date = ymd(`Month-Year`)) %>%
  filter(year >= 2010)
```

## Plotting the Horizon Chart

`geom_horizon()` of the ggHoriPlot package builds the bands, and we facet by
country so each one gets its own compact strip. Colour here encodes the value
band, which is the core idea of the chart.

```{r}
#| fig-height: 12
air_long %>%
  ggplot() +
  geom_horizon(aes(x = date,
                   y = arrivals),
               origin = "midpoint",
               horizonscale = 6) +
  scale_fill_hcl(palette = 'RdBu') +
  facet_grid(country ~ .) +
  theme_few() +
  theme(panel.spacing.y = unit(0, "lines"),
        strip.text.y = element_text(size = 5,
                                    angle = 0,
                                    hjust = 0),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        panel.border = element_blank(),
        legend.position = 'none') +
  labs(title = 'Visitor arrivals by air, 2010 onward',
       x = NULL, y = NULL)
```
