---
title: "Hands-on Exercise 3"
author: "ZeXin Chen"
date: "2026-04-26"
date-modified: last-modified
format:
  html:
    toc: true
    code-fold: true
    code-summary: "Show code"
execute:
  echo: true
  eval: true
  warning: false
  message: false
  freeze: true
---

## Getting Started

### Install and load packages

```{r}
pacman::p_load(ggiraph, plotly, patchwork, DT, tidyverse,
               readxl, gifski, gapminder, gganimate)
```

### Import data

```{r}
exam_data <- read_csv("data/Exam_data.csv")
```

## Interactive Data Visualisation: ggiraph

### Tooltip effect

```{r}
p <- ggplot(data = exam_data, aes(x = MATHS)) +
  geom_dotplot_interactive(
    aes(tooltip = ID),
    stackgroups = TRUE,
    binwidth = 1,
    method = "histodot") +
  scale_y_continuous(NULL, breaks = NULL)

girafe(ggobj = p, width_svg = 6, height_svg = 6*0.618)
```

### Displaying multiple information on tooltip

```{r}
exam_data$tooltip <- c(paste0("Name = ", exam_data$ID,
                               "\n Class = ", exam_data$CLASS))

p <- ggplot(data = exam_data, aes(x = MATHS)) +
  geom_dotplot_interactive(
    aes(tooltip = exam_data$tooltip),
    stackgroups = TRUE,
    binwidth = 1,
    method = "histodot") +
  scale_y_continuous(NULL, breaks = NULL)

girafe(ggobj = p, width_svg = 8, height_svg = 8*0.618)
```

### Customising Tooltip style

```{r}
tooltip_css <- "background-color:white; font-style:bold; color:black;"

p <- ggplot(data = exam_data, aes(x = MATHS)) +
  geom_dotplot_interactive(
    aes(tooltip = ID),
    stackgroups = TRUE,
    binwidth = 1,
    method = "histodot") +
  scale_y_continuous(NULL, breaks = NULL)

girafe(ggobj = p,
       width_svg = 6,
       height_svg = 6*0.618,
       options = list(opts_tooltip(css = tooltip_css)))
```

### Displaying statistics on tooltip

```{r}
tooltip <- function(y, ymax, accuracy = .01) {
  mean <- scales::number(y, accuracy = accuracy)
  sem <- scales::number(ymax - y, accuracy = accuracy)
  paste("Mean maths scores:", mean, "+/-", sem)
}

gg_point <- ggplot(data = exam_data, aes(x = RACE)) +
  stat_summary(aes(y = MATHS,
                   tooltip = after_stat(tooltip(y, ymax))),
               fun.data = "mean_se",
               geom = GeomInteractiveCol,
               fill = "light blue") +
  stat_summary(aes(y = MATHS),
               fun.data = mean_se,
               geom = "errorbar", width = 0.2, size = 0.2)

girafe(ggobj = gg_point, width_svg = 8, height_svg = 8*0.618)
```

### Hover effect with data_id

```{r}
p <- ggplot(data = exam_data, aes(x = MATHS)) +
  geom_dotplot_interactive(
    aes(data_id = CLASS),
    stackgroups = TRUE,
    binwidth = 1,
    method = "histodot") +
  scale_y_continuous(NULL, breaks = NULL)

girafe(ggobj = p, width_svg = 6, height_svg = 6*0.618)
```

### Styling hover effect

```{r}
p <- ggplot(data = exam_data, aes(x = MATHS)) +
  geom_dotplot_interactive(
    aes(data_id = CLASS),
    stackgroups = TRUE,
    binwidth = 1,
    method = "histodot") +
  scale_y_continuous(NULL, breaks = NULL)

girafe(ggobj = p,
       width_svg = 6,
       height_svg = 6*0.618,
       options = list(
         opts_hover(css = "fill: #202020;"),
         opts_hover_inv(css = "opacity:0.2;")))
```

### Combining tooltip and hover effect

```{r}
p <- ggplot(data = exam_data, aes(x = MATHS)) +
  geom_dotplot_interactive(
    aes(tooltip = CLASS, data_id = CLASS),
    stackgroups = TRUE,
    binwidth = 1,
    method = "histodot") +
  scale_y_continuous(NULL, breaks = NULL)

girafe(ggobj = p,
       width_svg = 6,
       height_svg = 6*0.618,
       options = list(
         opts_hover(css = "fill: #202020;"),
         opts_hover_inv(css = "opacity:0.2;")))
```

### Click effect with onclick

```{r}
exam_data$onclick <- sprintf("window.open(\"%s%s\")",
  "https://www.moe.gov.sg/schoolfinder?journey=Primary%20school",
  as.character(exam_data$ID))

p <- ggplot(data = exam_data, aes(x = MATHS)) +
  geom_dotplot_interactive(
    aes(onclick = onclick),
    stackgroups = TRUE,
    binwidth = 1,
    method = "histodot") +
  scale_y_continuous(NULL, breaks = NULL)

girafe(ggobj = p, width_svg = 6, height_svg = 6*0.618)
```

### Coordinated Multiple Views with ggiraph

```{r}
p1 <- ggplot(data = exam_data, aes(x = MATHS)) +
  geom_dotplot_interactive(
    aes(data_id = ID),
    stackgroups = TRUE,
    binwidth = 1,
    method = "histodot") +
  coord_cartesian(xlim = c(0, 100)) +
  scale_y_continuous(NULL, breaks = NULL)

p2 <- ggplot(data = exam_data, aes(x = ENGLISH)) +
  geom_dotplot_interactive(
    aes(data_id = ID),
    stackgroups = TRUE,
    binwidth = 1,
    method = "histodot") +
  coord_cartesian(xlim = c(0, 100)) +
  scale_y_continuous(NULL, breaks = NULL)

girafe(code = print(p1 + p2),
       width_svg = 6,
       height_svg = 3,
       options = list(
         opts_hover(css = "fill: #202020;"),
         opts_hover_inv(css = "opacity:0.2;")))
```

## Interactive Data Visualisation: plotly

### Interactive scatter plot: plot_ly()

```{r}
plot_ly(data = exam_data,
        x = ~MATHS,
        y = ~ENGLISH,
        color = ~RACE)
```

### Interactive scatter plot: ggplotly()

```{r}
p <- ggplot(data = exam_data,
            aes(x = MATHS, y = ENGLISH, color = RACE)) +
  geom_point(size = 1) +
  coord_cartesian(xlim = c(0, 100), ylim = c(0, 100))

ggplotly(p)
```

### Coordinated Multiple Views with plotly

```{r}
d <- highlight_key(exam_data)

p1 <- ggplot(data = d, aes(x = MATHS, y = ENGLISH)) +
  geom_point(size = 1) +
  coord_cartesian(xlim = c(0, 100), ylim = c(0, 100))

p2 <- ggplot(data = d, aes(x = ENGLISH, y = SCIENCE)) +
  geom_point(size = 1) +
  coord_cartesian(xlim = c(0, 100), ylim = c(0, 100))

subplot(ggplotly(p1), ggplotly(p2))
```

## Interactive Data Table: DT

```{r}
DT::datatable(exam_data, class = "compact")
```

## Animated Data Visualisation: gganimate

### Import Global Population data

```{r}
col <- c("Country", "Continent")
globalPop <- read_xls("data/GlobalPopulation.xls", sheet = "Data") %>%
  mutate(across(col, as.factor)) %>%
  mutate(Year = as.integer(Year))
```

### Static bubble plot

```{r}
ggplot(globalPop, aes(x = Old, y = Young,
                      size = Population,
                      colour = Country)) +
  geom_point(alpha = 0.7, show.legend = FALSE) +
  scale_colour_manual(values = country_colors) +
  scale_size(range = c(2, 12)) +
  labs(title = 'Year: {frame_time}',
       x = '% Aged', y = '% Young')
```

### Animated bubble plot: gganimate

```{r}
ggplot(globalPop, aes(x = Old, y = Young,
                      size = Population,
                      colour = Country)) +
  geom_point(alpha = 0.7, show.legend = FALSE) +
  scale_colour_manual(values = country_colors) +
  scale_size(range = c(2, 12)) +
  labs(title = 'Year: {frame_time}',
       x = '% Aged', y = '% Young') +
  transition_time(Year) +
  ease_aes('linear')
```

### Animated bubble plot: ggplotly()

```{r}
gg <- ggplot(globalPop, aes(x = Old, y = Young,
                             size = Population,
                             colour = Country,
                             frame = Year)) +
  geom_point(alpha = 0.7, show.legend = FALSE) +
  scale_colour_manual(values = country_colors) +
  scale_size(range = c(2, 12)) +
  labs(x = '% Aged', y = '% Young') +
  theme(legend.position = 'none')

ggplotly(gg)
```

### Animated bubble plot: plot_ly()

```{r}
bp <- globalPop %>%
  plot_ly(x = ~Old,
          y = ~Young,
          size = ~Population,
          color = ~Continent,
          frame = ~Year,
          text = ~Country,
          hoverinfo = "text",
          type = 'scatter',
          mode = 'markers') %>%
  layout(showlegend = FALSE)

bp
```