---
title: "Hands-on Exercise 2"
author: "ZeXin Chen"
date: "2026-04-25"
date-modified: last-modified
format: html
execute:
  echo: true
  eval: true
  warning: false
  freeze: true
---

## Getting Started

### Install and load packages

```{r}
pacman::p_load(ggrepel, patchwork, ggthemes, hrbrthemes, tidyverse)
```

### Import data

```{r}
exam_data <- read_csv("data/Exam_data.csv")
```

## Beyond ggplot2 Annotation: ggrepel

### Without ggrepel

:::: panel-tabset
## The Plot
```{r}
#| echo: false
ggplot(exam_data, aes(x = MATHS, y = ENGLISH)) +
  geom_point() +
  geom_text(aes(label = ID), hjust = 0.5, vjust = -0.5) +
  coord_cartesian(xlim = c(0,100), ylim = c(0,100)) +
  ggtitle("English vs Maths scores - without ggrepel")
```

## The Code
```{r}
#| eval: false
ggplot(exam_data, aes(x = MATHS, y = ENGLISH)) +
  geom_point() +
  geom_text(aes(label = ID), hjust = 0.5, vjust = -0.5) +
  coord_cartesian(xlim = c(0,100), ylim = c(0,100)) +
  ggtitle("English vs Maths scores - without ggrepel")
```
::::

### With ggrepel

:::: panel-tabset
## The Plot
```{r}
#| echo: false
ggplot(exam_data, aes(x = MATHS, y = ENGLISH)) +
  geom_point() +
  geom_text_repel(aes(label = ID), size = 3) +
  coord_cartesian(xlim = c(0,100), ylim = c(0,100)) +
  ggtitle("English vs Maths scores - with ggrepel")
```

## The Code
```{r}
#| eval: false
ggplot(exam_data, aes(x = MATHS, y = ENGLISH)) +
  geom_point() +
  geom_text_repel(aes(label = ID), size = 3) +
  coord_cartesian(xlim = c(0,100), ylim = c(0,100)) +
  ggtitle("English vs Maths scores - with ggrepel")
```
::::

## Beyond ggplot2 Themes

### ggthemes — Economist theme

:::: panel-tabset
## The Plot
```{r}
#| echo: false
ggplot(exam_data, aes(x = MATHS)) +
  geom_histogram(bins = 20, fill = "grey", color = "black") +
  ggtitle("Distribution of Maths scores") +
  theme_economist()
```

## The Code
```{r}
#| eval: false
ggplot(exam_data, aes(x = MATHS)) +
  geom_histogram(bins = 20, fill = "grey", color = "black") +
  ggtitle("Distribution of Maths scores") +
  theme_economist()
```
::::

### hrbrthemes

:::: panel-tabset
## The Plot
```{r}
#| echo: false
ggplot(exam_data, aes(x = MATHS)) +
  geom_histogram(bins = 20, fill = "grey", color = "black") +
  ggtitle("Distribution of Maths scores") +
  theme_ipsum(axis_title_size = 18, base_size = 15, grid = "Y")
```

## The Code
```{r}
#| eval: false
ggplot(exam_data, aes(x = MATHS)) +
  geom_histogram(bins = 20, fill = "grey", color = "black") +
  ggtitle("Distribution of Maths scores") +
  theme_ipsum(axis_title_size = 18, base_size = 15, grid = "Y")
```
::::

## Beyond Single Graph: patchwork

### Two graphs side by side

:::: panel-tabset
## The Plot
```{r}
#| echo: false
p1 <- ggplot(exam_data, aes(x = MATHS)) +
  geom_histogram(bins = 20, color = "grey30", fill = "grey90") +
  ggtitle("Distribution of Maths scores")

p2 <- ggplot(exam_data, aes(x = ENGLISH)) +
  geom_histogram(bins = 20, color = "grey30", fill = "grey90") +
  ggtitle("Distribution of English scores")

p1 + p2
```

## The Code
```{r}
#| eval: false
p1 <- ggplot(exam_data, aes(x = MATHS)) +
  geom_histogram(bins = 20, color = "grey30", fill = "grey90") +
  ggtitle("Distribution of Maths scores")

p2 <- ggplot(exam_data, aes(x = ENGLISH)) +
  geom_histogram(bins = 20, color = "grey30", fill = "grey90") +
  ggtitle("Distribution of English scores")

p1 + p2
```
::::

### Three graphs combined

:::: panel-tabset
## The Plot
```{r}
#| echo: false
p3 <- ggplot(exam_data, aes(x = MATHS, y = ENGLISH)) +
  geom_point() +
  geom_smooth(method = lm, linewidth = 0.5) +
  coord_cartesian(xlim = c(0,100), ylim = c(0,100)) +
  ggtitle("English vs Maths scores")

(p1 / p2) | p3
```

## The Code
```{r}
#| eval: false
p3 <- ggplot(exam_data, aes(x = MATHS, y = ENGLISH)) +
  geom_point() +
  geom_smooth(method = lm, linewidth = 0.5) +
  coord_cartesian(xlim = c(0,100), ylim = c(0,100)) +
  ggtitle("English vs Maths scores")

(p1 / p2) | p3
```
::::

### With auto-tagging

:::: panel-tabset
## The Plot
```{r}
#| echo: false
((p1 / p2) | p3) +
  plot_annotation(tag_levels = 'I')
```

## The Code
```{r}
#| eval: false
((p1 / p2) | p3) +
  plot_annotation(tag_levels = 'I')
```
::::

### Inset plot

:::: panel-tabset
## The Plot
```{r}
#| echo: false
p3 + inset_element(p2, left = 0.02, bottom = 0.7, right = 0.5, top = 1)
```

## The Code
```{r}
#| eval: false
p3 + inset_element(p2, left = 0.02, bottom = 0.7, right = 0.5, top = 1)
```
::::