---
title: "Hands-on Exercise 5"
author: "ZeXin Chen"
date: "2025-05-14"
date-modified: last-modified
format:
  html:
    toc: true
    code-fold: true
    code-summary: "Show code"
execute:
  echo: true
  eval: true
  warning: false
  message: false
  freeze: true
---

# Ternary Plot

## Getting Started

### Installing and Loading Packages

```{r}
pacman::p_load(plotly, ggtern, tidyverse)
```

### Importing Data

```{r}
pop_data <- read_csv("data/respopagsex2000to2018_tidy.csv")
```

### Preparing the Data

```{r}
agpop_mutated <- pop_data %>%
  mutate(`Year` = as.character(Year)) %>%
  spread(AG, Population) %>%
  mutate(YOUNG = rowSums(.[4:8])) %>%
  mutate(ACTIVE = rowSums(.[9:16])) %>%
  mutate(OLD = rowSums(.[17:21])) %>%
  mutate(TOTAL = rowSums(.[22:24])) %>%
  filter(Year == 2018) %>%
  filter(TOTAL > 0)
```

## Plotting Ternary Diagram

### Static Ternary Plot

```{r}
ggtern(data = agpop_mutated,
       aes(x = YOUNG,
           y = ACTIVE,
           z = OLD)) +
  geom_point()
```

### Static Ternary Plot with Theme

```{r}
ggtern(data = agpop_mutated,
       aes(x = YOUNG,
           y = ACTIVE,
           z = OLD)) +
  geom_point() +
  labs(title = "Population structure, 2018") +
  theme_rgbw()
```

### Interactive Ternary Plot

```{r}
label <- function(txt) {
  list(
    text = txt,
    x = 0.1, y = 1,
    ax = 0, ay = 0,
    xref = "paper", yref = "paper",
    align = "center",
    font = list(family = "serif", size = 15, color = "white"),
    bgcolor = "#b3b3b3", bordercolor = "black", borderwidth = 2
  )
}

axis <- function(txt) {
  list(
    title = txt, tickformat = ".0%", tickfont = list(size = 10)
  )
}

ternaryAxes <- list(
  aaxis = axis("Young"),
  baxis = axis("Active"),
  caxis = axis("Old")
)

plot_ly(
  agpop_mutated,
  a = ~YOUNG,
  b = ~ACTIVE,
  c = ~OLD,
  color = I("black"),
  type = "scatterternary"
) %>%
  layout(
    annotations = label("Ternary Markers"),
    ternary = ternaryAxes
  )
```

# Visual Correlation Analysis

## Getting Started

### Installing and Loading Packages

```{r}
pacman::p_load(corrplot, ggstatsplot, tidyverse)
```

### Importing Data

```{r}
wine <- read_csv("data/wine_quality.csv")
```

## Correlation Matrix: pairs() Method

### Basic Correlation Matrix

```{r}
pairs(wine[,1:11])
```

### Lower Corner Only

```{r}
pairs(wine[,2:12], upper.panel = NULL)
```

### With Correlation Coefficients

```{r}
panel.cor <- function(x, y, digits=2, prefix="", cex.cor, ...) {
  usr <- par("usr")
  on.exit(par(usr))
  par(usr = c(0, 1, 0, 1))
  r <- abs(cor(x, y, use="complete.obs"))
  txt <- format(c(r, 0.123456789), digits=digits)[1]
  txt <- paste(prefix, txt, sep="")
  if(missing(cex.cor)) cex.cor <- 0.8/strwidth(txt)
  text(0.5, 0.5, txt, cex = cex.cor * (1 + r) / 2)
}

pairs(wine[,2:12],
      upper.panel = panel.cor)
```

## Visualising Correlation Matrix: ggcorrmat()

### Basic ggcorrmat Plot

```{r}
ggstatsplot::ggcorrmat(
  data = wine,
  cor.vars = 1:11)
```

### ggcorrmat with Customisation

```{r}
ggstatsplot::ggcorrmat(
  data = wine,
  cor.vars = 1:11,
  ggcorrplot.args = list(outline.color = "black",
                         hc.order = TRUE,
                         tl.cex = 10),
  title    = "Correlogram for wine dataset",
  subtitle = "Four pairs are not significant at p < 0.05"
)
```

## Building Multiple Plots

```{r}
grouped_ggcorrmat(
  data = wine,
  cor.vars = 1:11,
  grouping.var = type,
  type = "robust",
  p.adjust.method = "holm",
  plotgrid.args = list(ncol = 2),
  ggcorrplot.args = list(outline.color = "black",
                         hc.order = TRUE,
                         tl.cex = 10),
  annotation.args = list(
    tag_levels = "a",
    title = "Correlogram for wine dataset",
    subtitle = "The measures are: alcohol, sulphates, fixed acidity, citric acid, chlorides, residual sugar, density, free sulfur dioxide and volatile acidity",
    caption = "Dataset: UCI Machine Learning Repository"
  )
)
```

## Visualising Correlation Matrix: corrplot Package

### Computing Correlation Matrix

```{r}
wine.cor <- cor(wine[, 1:11])
```

### Basic corrplot

```{r}
corrplot(wine.cor)
```

### Using Ellipse Method

```{r}
corrplot(wine.cor,
         method = "ellipse")
```

### Lower Layout

```{r}
corrplot(wine.cor,
         method = "ellipse",
         type = "lower",
         diag = FALSE,
         tl.col = "black")
```

### Mixed Layout

```{r}
corrplot.mixed(wine.cor,
               lower = "ellipse",
               upper = "number",
               tl.pos = "lt",
               diag = "l",
               tl.col = "black")
```

### With Significance Test

```{r}
wine.sig = cor.mtest(wine.cor, conf.level = .95)

corrplot(wine.cor,
         method = "number",
         type = "lower",
         diag = FALSE,
         tl.col = "black",
         tl.srt = 45,
         p.mat = wine.sig$p,
         sig.level = .05)
```

### Reordering with AOE

```{r}
corrplot.mixed(wine.cor,
               lower = "ellipse",
               upper = "number",
               tl.pos = "lt",
               diag = "l",
               order = "AOE",
               tl.col = "black")
```

### Reordering with hclust

```{r}
corrplot(wine.cor,
         method = "ellipse",
         tl.pos = "lt",
         tl.col = "black",
         order = "hclust",
         hclust.method = "ward.D",
         addrect = 3)
```

# Heatmap for Visualising and Analysing Multivariate Data

## Getting Started

### Installing and Loading Packages

```{r}
pacman::p_load(seriation, dendextend, heatmaply, tidyverse)
```

### Importing Data

```{r}
wh <- read_csv("data/WHData-2018.csv")
```

### Preparing the Data

```{r}
row.names(wh) <- wh$Country
```

### Transforming to Matrix

```{r}
wh1 <- dplyr::select(wh, c(3, 7:12))
wh_matrix <- data.matrix(wh)
```

## Static Heatmap

### Basic Heatmap

```{r}
wh_heatmap <- heatmap(wh_matrix,
                      Rowv=NA, Colv=NA)
```

### Cluster Heatmap

```{r}
wh_heatmap <- heatmap(wh_matrix)
```

### Normalised Heatmap

```{r}
wh_heatmap <- heatmap(wh_matrix,
                      scale="column",
                      cexRow = 0.6,
                      cexCol = 0.8,
                      margins = c(10, 4))
```

## Interactive Heatmap

### Basic heatmaply

```{r}
heatmaply(wh_matrix[, -c(1, 2, 4, 5)])
```

### Scaling Method

```{r}
heatmaply(wh_matrix[, -c(1, 2, 4, 5)],
          scale = "column")
```

### Normalising Method

```{r}
heatmaply(normalize(wh_matrix[, -c(1, 2, 4, 5)]))
```

### Percentising Method

```{r}
heatmaply(percentize(wh_matrix[, -c(1, 2, 4, 5)]))
```

### Manual Clustering Approach

```{r}
heatmaply(normalize(wh_matrix[, -c(1, 2, 4, 5)]),
          dist_method = "euclidean",
          hclust_method = "ward.D")
```

### Statistical Clustering Approach

```{r}
wh_d <- dist(normalize(wh_matrix[, -c(1, 2, 4, 5)]), method = "euclidean")
dend_expend(wh_d)[[3]]
```

```{r}
wh_clust <- hclust(wh_d, method = "average")
num_k <- find_k(wh_clust)
plot(num_k)
```

```{r}
heatmaply(normalize(wh_matrix[, -c(1, 2, 4, 5)]),
          dist_method = "euclidean",
          hclust_method = "average",
          k_row = 3)
```

### Seriation: OLO

```{r}
heatmaply(normalize(wh_matrix[, -c(1, 2, 4, 5)]),
          seriate = "OLO")
```

### Seriation: GW

```{r}
heatmaply(normalize(wh_matrix[, -c(1, 2, 4, 5)]),
          seriate = "GW")
```

### Seriation: Mean

```{r}
heatmaply(normalize(wh_matrix[, -c(1, 2, 4, 5)]),
          seriate = "mean")
```

### Working with Colour Palettes

```{r}
heatmaply(normalize(wh_matrix[, -c(1, 2, 4, 5)]),
          seriate = "none",
          colors = Blues)
```

### Finishing Touch

```{r}
heatmaply(normalize(wh_matrix[, -c(1, 2, 4, 5)]),
          Colv=NA,
          seriate = "none",
          colors = Blues,
          k_row = 5,
          margins = c(NA,200,60,NA),
          fontsize_row = 4,
          fontsize_col = 5,
          main="World Happiness Score and Variables by Country, 2018 \nData Transformation using Normalise Method",
          xlab = "World Happiness Indicators",
          ylab = "World Countries")
```

# Visual Multivariate Analysis with Parallel Coordinates Plot

## Getting Started

### Installing and Loading Packages

```{r}
pacman::p_load(GGally, parallelPlot, tidyverse)
```

### Importing Data

```{r}
wh <- read_csv("data/WHData-2018.csv")
```

## Static Parallel Coordinates Plot

### Basic Parallel Coordinates

```{r}
ggparcoord(data = wh,
           columns = c(7:12))
```

### Parallel Coordinates with Boxplot

```{r}
ggparcoord(data = wh,
           columns = c(7:12),
           groupColumn = 2,
           scale = "uniminmax",
           alphaLines = 0.2,
           boxplot = TRUE,
           title = "Parallel Coordinates Plot of World Happiness Variables")
```

### Parallel Coordinates with Facet

```{r}
ggparcoord(data = wh,
           columns = c(7:12),
           groupColumn = 2,
           scale = "uniminmax",
           alphaLines = 0.2,
           boxplot = TRUE,
           title = "Multiple Parallel Coordinates Plots of World Happiness Variables by Region") +
  facet_wrap(~ Region)
```

### Rotating x-axis Labels

```{r}
ggparcoord(data = wh,
           columns = c(7:12),
           groupColumn = 2,
           scale = "uniminmax",
           alphaLines = 0.2,
           boxplot = TRUE,
           title = "Multiple Parallel Coordinates Plots of World Happiness Variables by Region") +
  facet_wrap(~ Region) +
  theme(axis.text.x = element_text(angle = 30, hjust=1))
```

## Interactive Parallel Coordinates Plot

### Preparing Data

```{r}
wh <- wh %>%
  select("Happiness score", c(7:12))
```

### Basic parallelPlot

```{r}
parallelPlot(wh,
             width = 320,
             height = 250)
```

### Rotating Axis Labels

```{r}
parallelPlot(wh,
             rotateTitle = TRUE)
```

### Changing Colour Scheme

```{r}
parallelPlot(wh,
             continuousCS = "YlOrRd",
             rotateTitle = TRUE)
```

### Parallel Coordinates with Histogram

```{r}
histoVisibility <- rep(TRUE, ncol(wh))
parallelPlot(wh,
             rotateTitle = TRUE,
             histoVisibility = histoVisibility)
```

# Treemap Visualisation with R

## Getting Started

### Installing and Loading Packages

```{r}
pacman::p_load(treemap, treemapify, tidyverse)
```

### Importing Data

```{r}
realis2018 <- read_csv("data/realis2018.csv")
```

## Data Wrangling

### Grouped Summaries with Pipe

```{r}
realis2018_summarised <- realis2018 %>%
  group_by(`Project Name`,`Planning Region`,
           `Planning Area`, `Property Type`,
           `Type of Sale`) %>%
  summarise(`Total Unit Sold` = sum(`No. of Units`, na.rm = TRUE),
            `Total Area` = sum(`Area (sqm)`, na.rm = TRUE),
            `Median Unit Price ($ psm)` = median(`Unit Price ($ psm)`, na.rm = TRUE),
            `Median Transacted Price` = median(`Transacted Price ($)`, na.rm = TRUE))
```

### Filtering Resale Condominiums

```{r}
realis2018_selected <- realis2018_summarised %>%
  filter(`Property Type` == "Condominium", `Type of Sale` == "Resale")
```

## Designing Treemap with treemap Package

### Basic Treemap

```{r}
treemap(realis2018_selected,
        index=c("Planning Region", "Planning Area", "Project Name"),
        vSize="Total Unit Sold",
        vColor="Median Unit Price ($ psm)",
        title="Resale Condominium by Planning Region and Area, 2017",
        title.legend = "Median Unit Price (S$ per sq. m)"
        )
```

### Value Type Treemap

```{r}
treemap(realis2018_selected,
        index=c("Planning Region", "Planning Area", "Project Name"),
        vSize="Total Unit Sold",
        vColor="Median Unit Price ($ psm)",
        type = "value",
        title="Resale Condominium by Planning Region and Area, 2017",
        title.legend = "Median Unit Price (S$ per sq. m)"
        )
```

### Value Type with RdYlBu Palette

```{r}
treemap(realis2018_selected,
        index=c("Planning Region", "Planning Area", "Project Name"),
        vSize="Total Unit Sold",
        vColor="Median Unit Price ($ psm)",
        type="value",
        palette="RdYlBu",
        title="Resale Condominium by Planning Region and Area, 2017",
        title.legend = "Median Unit Price (S$ per sq. m)"
        )
```

### Manual Type with Blues Palette

```{r}
treemap(realis2018_selected,
        index=c("Planning Region", "Planning Area", "Project Name"),
        vSize="Total Unit Sold",
        vColor="Median Unit Price ($ psm)",
        type="manual",
        palette="Blues",
        title="Resale Condominium by Planning Region and Area, 2017",
        title.legend = "Median Unit Price (S$ per sq. m)"
        )
```

### Squarified Algorithm

```{r}
treemap(realis2018_selected,
        index=c("Planning Region", "Planning Area", "Project Name"),
        vSize="Total Unit Sold",
        vColor="Median Unit Price ($ psm)",
        type="manual",
        palette="Blues",
        algorithm = "squarified",
        title="Resale Condominium by Planning Region and Area, 2017",
        title.legend = "Median Unit Price (S$ per sq. m)"
        )
```

### Using sortID

```{r}
treemap(realis2018_selected,
        index=c("Planning Region", "Planning Area", "Project Name"),
        vSize="Total Unit Sold",
        vColor="Median Unit Price ($ psm)",
        type="manual",
        palette="Blues",
        algorithm = "pivotSize",
        sortID = "Median Transacted Price",
        title="Resale Condominium by Planning Region and Area, 2017",
        title.legend = "Median Unit Price (S$ per sq. m)"
        )
```

## Designing Treemap using treemapify Package

### Basic treemapify

```{r}
ggplot(data=realis2018_selected,
       aes(area = `Total Unit Sold`,
           fill = `Median Unit Price ($ psm)`),
       layout = "scol",
       start = "bottomleft") +
  geom_treemap() +
  scale_fill_gradient(low = "light blue", high = "blue")
```

### Group by Planning Region

```{r}
ggplot(data=realis2018_selected,
       aes(area = `Total Unit Sold`,
           fill = `Median Unit Price ($ psm)`,
           subgroup = `Planning Region`),
       start = "topleft") +
  geom_treemap()
```

### Group by Planning Area with Boundary Lines

```{r}
ggplot(data=realis2018_selected,
       aes(area = `Total Unit Sold`,
           fill = `Median Unit Price ($ psm)`,
           subgroup = `Planning Region`,
           subgroup2 = `Planning Area`)) +
  geom_treemap() +
  geom_treemap_subgroup2_border(colour = "gray40",
                                size = 2) +
  geom_treemap_subgroup_border(colour = "gray20")
```

## Designing Interactive Treemap using d3treeR

### Installing d3treeR

```{r}
#| eval: false
library(devtools)
install_github("timelyportfolio/d3treeR")
```

### Interactive Treemap

```{r}
library(d3treeR)

tm <- treemap(realis2018_summarised,
        index=c("Planning Region", "Planning Area"),
        vSize="Total Unit Sold",
        vColor="Median Unit Price ($ psm)",
        type="value",
        title="Private Residential Property Sold, 2017",
        title.legend = "Median Unit Price (S$ per sq. m)"
        )

d3tree(tm, rootname = "Singapore")
```


