---
title: "Hands-on Exercise 9"
subtitle: "Visualising and Analysing Geographic Data"
author: "ZeXin Chen"
date: "2026-06-18"
date-modified: "last-modified"
format:
  html:
    toc: true
    number-sections: true
execute:
  echo: true
  eval: true
  warning: false
  message: false
  freeze: true
---

# Overview

In this hands-on exercise we learn to plot maps that carry data. We cover three
related skills:

- choropleth mapping, where regions are shaded by a value, using the **tmap** package
- geospatial point data, where outlets or events are drawn as proportional symbols
- analytical mapping, where we map rates and use percentile and box maps to surface outliers

The key new packages are **sf** for handling spatial data and **tmap** for drawing
thematic maps.

# Getting Started

## Installing and Loading Packages

```{r}
pacman::p_load(sf, tmap, tidyverse)
```

A quick note on the two new packages. **sf** reads and stores spatial data such as
boundaries and points, and **tmap** turns that data into maps. The **tidyverse**
packages handle the data wrangling along the way.

# Part A: Choropleth Mapping

A choropleth map shades enumeration units, such as countries or subzones, by a
value. Here we map the dependency ratio across Singapore planning subzones.

## The Data

Two datasets are used:

- the Master Plan 2014 Subzone Boundary, an ESRI shapefile holding the geographic
  boundaries of Singapore at the subzone level
- Singapore residents by planning area, subzone, age group and sex, an aspatial
  CSV that we will join to the boundaries

## Importing Geospatial Data

`st_read()` of the **sf** package reads the shapefile into a simple feature data
frame.

```{r}
mpsz <- st_read(dsn = "data/geospatial",
                layer = "MP14_SUBZONE_WEB_PL")
```

```{r}
mpsz
```

## Importing Attribute Data

```{r}
popdata <- read_csv("data/aspatial/respopagesextod2011to2020.csv")
```

## Data Preparation

We build a table of 2020 values with the fields we need: young, economy active,
aged, total, and a derived dependency ratio.

```{r}
popdata2020 <- popdata %>%
  filter(Time == 2020) %>%
  group_by(PA, SZ, AG) %>%
  summarise(`POP` = sum(`Pop`)) %>%
  ungroup() %>%
  pivot_wider(names_from = AG,
              values_from = POP) %>%
  mutate(YOUNG = rowSums(.[3:6]) + rowSums(.[12])) %>%
  mutate(`ECONOMY ACTIVE` = rowSums(.[7:11]) + rowSums(.[13:15])) %>%
  mutate(`AGED` = rowSums(.[16:21])) %>%
  mutate(`TOTAL` = rowSums(.[3:21])) %>%
  mutate(`DEPENDENCY` = (`YOUNG` + `AGED`) / `ECONOMY ACTIVE`) %>%
  select(`PA`, `SZ`, `YOUNG`,
         `ECONOMY ACTIVE`, `AGED`,
         `TOTAL`, `DEPENDENCY`)
```

Before joining, we convert the PA and SZ fields to uppercase so they match the
boundary data.

```{r}
popdata2020 <- popdata2020 %>%
  mutate(across(c(PA, SZ), toupper)) %>%
  filter(`ECONOMY ACTIVE` > 0)
```

Now we join the attribute table onto the boundary data using the subzone name as
the common key. Keeping `mpsz` on the left ensures the result stays a spatial
data frame.

```{r}
mpsz_pop2020 <- left_join(mpsz, popdata2020,
                          by = c("SUBZONE_N" = "SZ"))
```

## Choropleth Map Using qtm()

The quickest way to draw a choropleth map is `qtm()`. It gives a sensible default
in one line.

::: panel-tabset
## The Plot
```{r}
#| echo: false
tmap_mode("plot")
qtm(mpsz_pop2020,
    fill = "DEPENDENCY")
```
## The Code
```{r}
#| eval: false
tmap_mode("plot")
qtm(mpsz_pop2020,
    fill = "DEPENDENCY")
```
:::

## Choropleth Map Using tmap Elements

For more control over the look, we build the map from individual **tmap**
elements. Colour here encodes the dependency ratio, which is the point of the
map.

::: panel-tabset
## The Plot
```{r}
#| echo: false
tm_shape(mpsz_pop2020) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5,
                values = "brewer.blues"),
              fill.legend = tm_legend(
                title = "Dependency ratio")) +
  tm_title("Distribution of Dependency Ratio by planning subzone") +
  tm_layout(frame = TRUE) +
  tm_borders(fill_alpha = 0.5) +
  tm_compass(type = "8star", size = 2) +
  tm_grid(alpha = 0.2) +
  tm_credits("Source: URA Master Plan subzone boundary and Department of Statistics population data",
             position = c("left", "bottom"))
```
## The Code
```{r}
#| eval: false
tm_shape(mpsz_pop2020) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5,
                values = "brewer.blues"),
              fill.legend = tm_legend(
                title = "Dependency ratio")) +
  tm_title("Distribution of Dependency Ratio by planning subzone") +
  tm_layout(frame = TRUE) +
  tm_borders(fill_alpha = 0.5) +
  tm_compass(type = "8star", size = 2) +
  tm_grid(alpha = 0.2) +
  tm_credits("Source: URA Master Plan subzone boundary and Department of Statistics population data",
             position = c("left", "bottom"))
```
:::

## Data Classification Methods

`tmap` offers several ways to group values into classes. Below we compare the
jenks method with the equal method.

::: panel-tabset
## The Plot
```{r}
#| echo: false
tm_shape(mpsz_pop2020) +
  tm_polygons("DEPENDENCY",
              fill.scale = tm_scale_intervals(
                style = "jenks",
                n = 5)) +
  tm_borders(fill_alpha = 0.5)
```
## The Code
```{r}
#| eval: false
tm_shape(mpsz_pop2020) +
  tm_polygons("DEPENDENCY",
              fill.scale = tm_scale_intervals(
                style = "jenks",
                n = 5)) +
  tm_borders(fill_alpha = 0.5)
```
:::

::: panel-tabset
## The Plot
```{r}
#| echo: false
tm_shape(mpsz_pop2020) +
  tm_polygons("DEPENDENCY",
              fill.scale = tm_scale_intervals(
                style = "equal",
                n = 5)) +
  tm_borders(fill_alpha = 0.5)
```
## The Code
```{r}
#| eval: false
tm_shape(mpsz_pop2020) +
  tm_polygons("DEPENDENCY",
              fill.scale = tm_scale_intervals(
                style = "equal",
                n = 5)) +
  tm_borders(fill_alpha = 0.5)
```
:::

The jenks method spreads the classes to follow natural breaks in the data, while
the equal method splits the range into equal width bands, which can look very
different for skewed data.

## Colour Scheme

We can set the palette with the `values` argument. A minus prefix reverses it.

::: panel-tabset
## The Plot
```{r}
#| echo: false
tm_shape(mpsz_pop2020) +
  tm_polygons("DEPENDENCY",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5,
                values = "brewer.greens")) +
  tm_borders(fill_alpha = 0.5)
```
## The Code
```{r}
#| eval: false
tm_shape(mpsz_pop2020) +
  tm_polygons("DEPENDENCY",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5,
                values = "brewer.greens")) +
  tm_borders(fill_alpha = 0.5)
```
:::

## Small Multiple Choropleth Maps

Small multiples, also called facet maps, place several maps side by side so we
can compare a value across a grouping variable. Here we facet by region.

::: panel-tabset
## The Plot
```{r}
#| echo: false
tm_shape(mpsz_pop2020) +
  tm_polygons("DEPENDENCY",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5,
                values = "brewer.blues")) +
  tm_facets(by = "REGION_N",
            free.coords = TRUE) +
  tm_borders(fill_alpha = 0.5) +
  tm_layout(legend.show = FALSE)
```
## The Code
```{r}
#| eval: false
tm_shape(mpsz_pop2020) +
  tm_polygons("DEPENDENCY",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5,
                values = "brewer.blues")) +
  tm_facets(by = "REGION_N",
            free.coords = TRUE) +
  tm_borders(fill_alpha = 0.5) +
  tm_layout(legend.show = FALSE)
```
:::

# Part B: Visualising Geospatial Point Data

A proportional symbol map uses the size of a symbol to show the magnitude of a
value at a point. Here we map the winnings at Singapore Pools outlets.

## The Data

The dataset is `SGPools_svy21.csv`. Its XCOORD and YCOORD columns hold the outlet
coordinates in the Singapore SVY21 projected system.

## Importing the Aspatial Data

```{r}
sgpools <- read_csv("data/aspatial/SGPools_svy21.csv")
```

```{r}
list(sgpools)
```

## Creating a Spatial Point Data Frame

`st_as_sf()` converts the plain table into a spatial point data frame. The crs
3414 is the EPSG code for Singapore SVY21.

```{r}
sgpools_sf <- st_as_sf(sgpools,
                       coords = c("XCOORD", "YCOORD"),
                       crs = 3414)
```

```{r}
list(sgpools_sf)
```

## Interactive Point Symbol Map

We turn on the interactive view mode, then draw a basic point map.

::: panel-tabset
## The Plot
```{r}
#| echo: false
tmap_mode("view")
tm_shape(sgpools_sf) +
  tm_bubbles(fill = "red",
             size = 1,
             col = "black",
             lwd = 1)
```
## The Code
```{r}
#| eval: false
tmap_mode("view")
tm_shape(sgpools_sf) +
  tm_bubbles(fill = "red",
             size = 1,
             col = "black",
             lwd = 1)
```
:::

## Proportional Symbol Map

Assigning a numeric variable to `size` makes each bubble proportional to its
value, here the group 1 and group 2 winnings.

::: panel-tabset
## The Plot
```{r}
#| echo: false
tm_shape(sgpools_sf) +
  tm_bubbles(fill = "red",
             size = "Gp1Gp2 Winnings",
             col = "black",
             lwd = 1)
```
## The Code
```{r}
#| eval: false
tm_shape(sgpools_sf) +
  tm_bubbles(fill = "red",
             size = "Gp1Gp2 Winnings",
             col = "black",
             lwd = 1)
```
:::

## Adding Colour by Outlet Type

Mapping the outlet type to colour lets us compare branches and outlets while
keeping the proportional sizing.

::: panel-tabset
## The Plot
```{r}
#| echo: false
tm_shape(sgpools_sf) +
  tm_bubbles(fill = "OUTLET TYPE",
             size = "Gp1Gp2 Winnings",
             col = "black",
             lwd = 1)
```
## The Code
```{r}
#| eval: false
tm_shape(sgpools_sf) +
  tm_bubbles(fill = "OUTLET TYPE",
             size = "Gp1Gp2 Winnings",
             col = "black",
             lwd = 1)
```
:::

## Faceted Proportional Symbol Map

The view mode also works with facets. The sync argument keeps the zoom and pan in
step across the two maps.

::: panel-tabset
## The Plot
```{r}
#| echo: false
tm_shape(sgpools_sf) +
  tm_bubbles(fill = "OUTLET TYPE",
             size = "Gp1Gp2 Winnings",
             col = "black",
             lwd = 1) +
  tm_facets(by = "OUTLET TYPE",
            nrow = 1,
            sync = TRUE)
```
## The Code
```{r}
#| eval: false
tm_shape(sgpools_sf) +
  tm_bubbles(fill = "OUTLET TYPE",
             size = "Gp1Gp2 Winnings",
             col = "black",
             lwd = 1) +
  tm_facets(by = "OUTLET TYPE",
            nrow = 1,
            sync = TRUE)
```
:::

```{r}
tmap_mode("plot")
```

# Part C: Analytical Mapping

Analytical maps go beyond showing raw counts. We map rates, then use percentile
and box maps to highlight outliers. The data is water point information for
Nigeria at the local government area level.

## Importing the Data

The prepared dataset is an rds file, which `read_rds()` reads back into a spatial
data frame.

```{r}
NGA_wp <- read_rds("data/rds/NGA_wp.rds")
```

## Basic Choropleth of Counts

We start by mapping the total and functional water point counts side by side.

::: panel-tabset
## The Plot
```{r}
#| echo: false
p1 <- tm_shape(NGA_wp) +
  tm_polygons(fill = "wp_functional",
              fill.scale = tm_scale_intervals(
                style = "equal",
                n = 10,
                values = "brewer.blues")) +
  tm_borders(lwd = 0.1, fill_alpha = 1) +
  tm_title("Distribution of functional water point by LGAs")

p2 <- tm_shape(NGA_wp) +
  tm_polygons(fill = "total_wp",
              fill.scale = tm_scale_intervals(
                style = "equal",
                n = 10,
                values = "brewer.blues")) +
  tm_borders(lwd = 0.1, fill_alpha = 1) +
  tm_title("Distribution of total water point by LGAs")

tmap_arrange(p2, p1, nrow = 1)
```
## The Code
```{r}
#| eval: false
p1 <- tm_shape(NGA_wp) +
  tm_polygons(fill = "wp_functional",
              fill.scale = tm_scale_intervals(
                style = "equal",
                n = 10,
                values = "brewer.blues")) +
  tm_borders(lwd = 0.1, fill_alpha = 1) +
  tm_title("Distribution of functional water point by LGAs")

p2 <- tm_shape(NGA_wp) +
  tm_polygons(fill = "total_wp",
              fill.scale = tm_scale_intervals(
                style = "equal",
                n = 10,
                values = "brewer.blues")) +
  tm_borders(lwd = 0.1, fill_alpha = 1) +
  tm_title("Distribution of total water point by LGAs")

tmap_arrange(p2, p1, nrow = 1)
```
:::

## Choropleth Map for Rates

Mapping rates rather than counts is important because water points are not evenly
distributed. We derive the proportion of functional and non functional points
per area.

```{r}
NGA_wp <- NGA_wp %>%
  mutate(pct_functional = wp_functional / total_wp) %>%
  mutate(pct_nonfunctional = wp_nonfunctional / total_wp)
```

::: panel-tabset
## The Plot
```{r}
#| echo: false
tm_shape(NGA_wp) +
  tm_polygons("pct_functional",
              fill.scale = tm_scale_intervals(
                style = "equal",
                n = 10,
                values = "brewer.blues")) +
  tm_borders(lwd = 0.1, fill_alpha = 1) +
  tm_title("Rate map of functional water point by LGAs")
```
## The Code
```{r}
#| eval: false
tm_shape(NGA_wp) +
  tm_polygons("pct_functional",
              fill.scale = tm_scale_intervals(
                style = "equal",
                n = 10,
                values = "brewer.blues")) +
  tm_borders(lwd = 0.1, fill_alpha = 1) +
  tm_title("Rate map of functional water point by LGAs")
```
:::

## Percentile Map

A percentile map is a quantile map with six set categories: 0 to 1 percent, 1 to
10, 10 to 50, 50 to 90, 90 to 99, and 99 to 100 percent. We first drop records
with missing values, then write two small helper functions.

```{r}
NGA_wp <- NGA_wp %>%
  drop_na()
```

```{r}
get.var <- function(vname, df) {
  v <- df[vname] %>%
    st_set_geometry(NULL)
  v <- unname(v[, 1])
  return(v)
}
```

```{r}
percentmap <- function(vnam, df, legtitle = NA, mtitle = "Percentile Map") {
  percent <- c(0, .01, .1, .5, .9, .99, 1)
  var <- get.var(vnam, df)
  bperc <- quantile(var, percent)
  tm_shape(df) +
    tm_polygons() +
    tm_shape(df) +
    tm_polygons(vnam,
                title = legtitle,
                breaks = bperc,
                palette = "Blues",
                labels = c("< 1%", "1% - 10%", "10% - 50%",
                           "50% - 90%", "90% - 99%", "> 99%")) +
    tm_borders() +
    tm_layout(main.title = mtitle,
              title.position = c("right", "bottom"))
}
```

::: panel-tabset
## The Plot
```{r}
#| echo: false
percentmap("total_wp", NGA_wp)
```
## The Code
```{r}
#| eval: false
percentmap("total_wp", NGA_wp)
```
:::

## Box Map

A box map is a quartile map with extra categories for lower and upper outliers,
built on the logic of a boxplot. We start by looking at the distribution.

::: panel-tabset
## The Plot
```{r}
#| echo: false
ggplot(data = NGA_wp,
       aes(x = "",
           y = wp_nonfunctional)) +
  geom_boxplot()
```
## The Code
```{r}
#| eval: false
ggplot(data = NGA_wp,
       aes(x = "",
           y = wp_nonfunctional)) +
  geom_boxplot()
```
:::

We write a function to compute the break points, handling the case where there
are no lower or upper outliers.

```{r}
boxbreaks <- function(v, mult = 1.5) {
  qv <- unname(quantile(v))
  iqr <- qv[4] - qv[2]
  upfence <- qv[4] + mult * iqr
  lofence <- qv[2] - mult * iqr
  bb <- vector(mode = "numeric", length = 7)
  if (lofence < qv[1]) {
    bb[1] <- lofence
    bb[2] <- floor(qv[1])
  } else {
    bb[2] <- lofence
    bb[1] <- qv[1]
  }
  if (upfence > qv[5]) {
    bb[7] <- upfence
    bb[6] <- ceiling(qv[5])
  } else {
    bb[6] <- upfence
    bb[7] <- qv[5]
  }
  bb[3:5] <- qv[2:4]
  return(bb)
}
```

```{r}
boxmap <- function(vnam, df,
                   legtitle = NA,
                   mtitle = "Box Map",
                   mult = 1.5) {
  var <- get.var(vnam, df)
  bb <- boxbreaks(var)
  tm_shape(df) +
    tm_polygons() +
    tm_shape(df) +
    tm_fill(vnam, title = legtitle,
            breaks = bb,
            palette = "Blues",
            labels = c("lower outlier",
                       "< 25%",
                       "25% - 50%",
                       "50% - 75%",
                       "> 75%",
                       "upper outlier")) +
    tm_borders() +
    tm_layout(main.title = mtitle,
              title.position = c("left", "top"))
}
```

::: panel-tabset
## The Plot
```{r}
#| echo: false
tmap_mode("plot")
boxmap("wp_nonfunctional", NGA_wp)
```
## The Code
```{r}
#| eval: false
tmap_mode("plot")
boxmap("wp_nonfunctional", NGA_wp)
```
:::
