---
title: "Take-home Exercise 1: Visual Analytics of Motor Vehicle Insurance Portfolio"
author: "ZeXin Chen"
date: "2026-05-14"
date-modified: last-modified
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: right
    number-sections: true
    theme: flatly
    code-fold: true
    code-summary: "Show the code"
    code-overflow: wrap
execute:
  echo: true
  eval: true
  warning: false
  message: false
  freeze: false
---

## Overview

### Setting the Scene

Motor vehicle insurance is one of the largest lines of non-life (property & casualty) insurance globally. According to Swiss Re's *sigma* research, non-life insurers face two fundamental challenges: maintaining **profitability** (ensuring premiums collected exceed claims paid) and managing **volatility** (the unpredictable nature of claim occurrences and severities).

This exercise analyses a detailed microdata portfolio from a Spanish motor insurance company, covering 354,140 policy transactions from January 2022 to December 2024 across 47 variables. The dataset captures policy characteristics, driver and vehicle attributes, premiums charged, and claims experience.

### Our Task

Using Exploratory Data Analysis (EDA) methods and `ggplot2` visualisations, this exercise explores:

1.  **Univariate distributions** of all numerical and categorical variables
2.  **Bivariate relationships** to statistically identify factors influencing profitability and volatility

### Key Concepts

Two derived metrics guide the analysis throughout:

- **Loss Ratio** = `total_incurred / total_premium`: the primary measure of profitability. A loss ratio below 1 indicates the insurer collected more in premiums than it paid in claims.
- **Claim Frequency** = `total_claims / total_exposure`: measures how often claims occur per policy year, a key driver of volatility.
- **Claim Severity** = `total_incurred / total_claims` (where claims \> 0): the average cost per claim.

## Getting Started

### Load Packages

```{r}
pacman::p_load(
  tidyverse,    # data manipulation and ggplot2
  patchwork,    # combining plots
  ggdist,       # distribution visualisations
  ggridges,     # ridgeline plots
  ggthemes,     # additional themes
  ggrepel,      # non-overlapping labels
  scales,       # axis formatting
  knitr,        # tables
  kableExtra    # enhanced tables
)
```

### Import Data

The dataset uses semicolons as delimiters with dot decimal separators and contains 354,140 records across 47 variables.

```{r}
insurance <- read_delim(
  "data/Dataset of motor insurance portfolio.csv",
  delim = ";",
  locale = locale(decimal_mark = ".")
)
```

```{r}
glimpse(insurance)
```

### Data Pre-processing

#### Check for Duplicates

```{r}
n_dupes <- sum(duplicated(insurance))
cat("Number of duplicate rows:", n_dupes)
```

#### Check Missing Values

```{r}
missing_summary <- insurance |>
  summarise(across(everything(), ~ sum(is.na(.)))) |>
  pivot_longer(everything(), names_to = "variable", values_to = "missing_count") |>
  filter(missing_count > 0) |>
  mutate(missing_pct = round(missing_count / nrow(insurance) * 100, 2)) |>
  arrange(desc(missing_count))

kable(missing_summary,
      col.names = c("Variable", "Missing Count", "Missing %"),
      caption = "Variables with Missing Values") |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

::: callout-note
Four variables have missing values. `fuel_type` (\~0.4%) and `vehicle_value` (\~0.1%) have minor missingness and will be excluded during analysis of those specific variables. `vehicle_age` and `age_driving_licence` each have only 2 missing rows and are negligible. No imputation is performed.
:::

#### Convert Data Types

```{r}
insurance <- insurance |>
  mutate(
    policy_type = factor(policy_type,
                         levels = c("TP", "TPG", "CC", "COMP_E", "COMP_N"),
                         labels = c("TP: Basic", "TPG: +Glass", "CC: +Add-ons",
                                    "COMP/E: +Excess", "COMP/N: Full")),
    policy_status = factor(policy_status, levels = c("A", "C"),
                           labels = c("Active", "Cancelled")),
    business_type = factor(business_type, levels = c("NB", "P"),
                           labels = c("New Business", "Portfolio")),
    payment_frequency = factor(payment_frequency, levels = c("A", "S", "Q"),
                               labels = c("Annual", "Semiannual", "Quarterly")),
    bonus_score = factor(bonus_score, levels = c("G", "N", "B"),
                         labels = c("Good", "Neutral", "Bad")),
    fuel_type = factor(fuel_type, levels = c("D", "G"),
                       labels = c("Diesel", "Gasoline")),
    municipality_type = factor(municipality_type, levels = c("I", "C", "IS"),
                               labels = c("Inland", "Coastal", "Islands")),
    circulation_area = factor(circulation_area, levels = c("U", "R"),
                              labels = c("Urban", "Rural")),
    year = as.integer(year)
  )
```

#### Derive Key Metrics

```{r}
insurance <- insurance |>
  mutate(
    loss_ratio = ifelse(total_premium > 0,
                        total_incurred / total_premium, NA),
    claim_frequency = ifelse(total_exposure > 0,
                             total_claims / total_exposure, NA),
    claim_severity = ifelse(total_claims > 0,
                            total_incurred / total_claims, NA),
    driving_experience = driver_age - age_driving_licence
  )
```

#### Preview Clean Dataset

```{r}
insurance |>
  select(insured_id, year, policy_type, driver_age, total_premium,
         total_incurred, loss_ratio, claim_frequency) |>
  head(8) |>
  kable(digits = 3, caption = "Preview of Key Variables (first 8 rows)") |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

## EDA 1. Univariate Analysis of Categorical Variables

### Distribution of Policy Type

```{r}
#| fig-width: 8
#| fig-height: 4

insurance |>
  count(policy_type) |>
  mutate(pct = n / sum(n),
         policy_type = fct_reorder(policy_type, n)) |>
  ggplot(aes(x = policy_type, y = n, fill = policy_type)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = paste0(round(pct * 100, 1), "%")),
            hjust = -0.1, size = 3.5) +
  coord_flip() +
  scale_y_continuous(labels = comma,
                     expand = expansion(mult = c(0, 0.18))) +
  scale_fill_manual(values = c(
    "TP: Basic"      = "#BBDEFB", "TPG: +Glass"    = "#90CAF9",
    "CC: +Add-ons"   = "#42A5F5", "COMP/E: +Excess" = "#1976D2",
    "COMP/N: Full"   = "#0D47A1")) +
  labs(title = "Distribution of Policy Type",
       x = NULL, y = "Number of Policies") +
  theme_minimal() +
  theme(panel.grid.major.y = element_blank())
```

**Observation:** The CC: +Add-ons policy type dominates the portfolio at 57.8%, followed by COMP/E: +Excess at 25.8%. Basic TP: Basic policies are rare at just 1.6%, suggesting this portfolio skews toward customers who elect broader coverage. The near-absence of basic TP policies may indicate a deliberate underwriting strategy targeting higher-coverage segments.

------------------------------------------------------------------------

### Distribution of Policy Status

```{r}
#| fig-width: 7
#| fig-height: 4

insurance |>
  count(policy_status) |>
  mutate(pct = n / sum(n)) |>
  ggplot(aes(x = policy_status, y = n, fill = policy_status)) +
  geom_col(show.legend = FALSE, width = 0.5) +
  geom_text(aes(label = paste0(round(pct * 100, 1), "%\n(", comma(n), ")")),
            vjust = -0.3, size = 3.5) +
  scale_y_continuous(labels = comma,
                     expand = expansion(mult = c(0, 0.12))) +
  scale_fill_manual(values = c("Active" = "#2166AC", "Cancelled" = "#BABABA")) +
  labs(title = "Distribution of Policy Status",
       x = NULL, y = "Number of Policies") +
  theme_minimal() +
  theme(panel.grid.major.x = element_blank())
```

**Observation:** Approximately 89.2% of policies are active and 10.8% cancelled. The cancellation rate is not negligible. Policies cancelled mid-term reduce premium income and may signal adverse selection if higher-risk customers are disproportionately cancelling after claims.

------------------------------------------------------------------------

### Distribution of Business Type, Payment Frequency, and Bonus Score

```{r}
#| fig-width: 12
#| fig-height: 4

plot_cat <- function(data, var, title) {
  data |>
    count({{ var }}) |>
    mutate(pct = n / sum(n),
           var_col = fct_reorder({{ var }}, n)) |>
    ggplot(aes(x = var_col, y = n)) +
    geom_col(fill = "#2166AC", show.legend = FALSE) +
    geom_text(aes(label = paste0(round(pct * 100, 1), "%")),
              hjust = -0.1, size = 3.2) +
    coord_flip() +
    scale_y_continuous(labels = comma,
                       expand = expansion(mult = c(0, 0.2))) +
    labs(title = title, x = NULL, y = "Count") +
    theme_minimal() +
    theme(panel.grid.major.y = element_blank())
}

p3 <- plot_cat(insurance, business_type, "Business Type")
p4 <- plot_cat(insurance, payment_frequency, "Payment Frequency")
p5 <- plot_cat(insurance, bonus_score, "Bonus Score")

p3 | p4 | p5
```

**Observation:** New business policies (69.3%) substantially outnumber portfolio renewals (30.7%), suggesting the insurer is actively growing its book. Annual payment is overwhelmingly preferred (81.3%), likely reflecting discounts offered for upfront payment. The bonus score distribution is highly skewed toward "Good" (96.4%), indicating the portfolio is dominated by low-risk, experienced drivers with no recent claims history.

------------------------------------------------------------------------

### Distribution of Geographic Variables

```{r}
#| fig-width: 10
#| fig-height: 4

p6 <- plot_cat(insurance, municipality_type, "Municipality Type")
p7 <- plot_cat(insurance, circulation_area, "Circulation Area")
p8 <- plot_cat(insurance |> filter(!is.na(fuel_type)),
               fuel_type, "Fuel Type")

p6 | p7 | p8
```

**Observation:** Inland municipalities account for the majority (64.6%) with coastal at 28.6% and island at 6.8%, broadly reflecting Spain's population distribution. Urban circulation edges ahead of rural at 53.8% vs 46.2%. Diesel vehicles dominate at 64.2%, though this may shift as newer petrol and electric vehicles gain market share.

## EDA 2. Univariate Analysis of Numerical Variables

### Driver and Vehicle Characteristics

```{r}
#| fig-width: 12
#| fig-height: 5

p9 <- insurance |>
  ggplot(aes(x = driver_age)) +
  geom_histogram(binwidth = 5, fill = "#2166AC", colour = "white") +
  scale_y_continuous(labels = comma) +
  labs(title = "Driver Age Distribution",
       x = "Driver Age (years)", y = "Count") +
  theme_minimal()

p10 <- insurance |>
  ggplot(aes(x = vehicle_age)) +
  geom_histogram(binwidth = 3, fill = "#2166AC", colour = "white") +
  scale_y_continuous(labels = comma) +
  labs(title = "Vehicle Age Distribution",
       x = "Vehicle Age (years)", y = "Count") +
  theme_minimal()

p11 <- insurance |>
  filter(!is.na(vehicle_value)) |>
  ggplot(aes(x = vehicle_value)) +
  geom_histogram(bins = 50, fill = "#2166AC", colour = "white") +
  coord_cartesian(xlim = c(0, 80000)) +
  scale_x_continuous(labels = label_number(scale = 1e-3, suffix = "k")) +
  scale_y_continuous(labels = comma) +
  labs(title = "Vehicle Value Distribution",
       x = "Declared Vehicle Value EUR (capped 80k)", y = "Count",
       caption = "< 1% of vehicles exceed €80k") +
  theme_minimal()

p9 | p10 | p11
```

**Observation:** Driver age follows a roughly bell-shaped distribution centred around 45–55 years, with very few young drivers under 25, consistent with a portfolio biased toward experienced drivers. Vehicle age is right-skewed with a concentration of older vehicles (15–35 years). Vehicle values are capped at €80k to show the true distribution shape. The majority cluster between €15,000–€40,000 with a small tail of high-value vehicles that could significantly impact claim severity.

------------------------------------------------------------------------

### Premium Distribution

```{r}
#| fig-width: 12
#| fig-height: 4

p12 <- insurance |>
  ggplot(aes(x = total_premium)) +
  geom_histogram(bins = 60, fill = "#2166AC", colour = "white") +
  scale_x_continuous(labels = comma) +
  scale_y_continuous(labels = comma) +
  labs(title = "Total Premium Distribution",
       x = "Total Premium (EUR)", y = "Count") +
  theme_minimal()

p13 <- insurance |>
  ggplot(aes(x = total_premium)) +
  geom_histogram(bins = 60, fill = "#2166AC", colour = "white") +
  scale_x_continuous(trans = "log10", labels = comma) +
  scale_y_continuous(labels = comma) +
  labs(title = "Total Premium (log10 scale)",
       x = "Total Premium (EUR, log scale)", y = "Count") +
  theme_minimal()

p12 | p13
```

**Observation:** Total premiums are right-skewed with most policies priced between €200–€600. On the log scale, the distribution becomes more symmetric, centred around €300–€400, typical for Spanish motor insurance. The small number of very high premiums (\>€2,000) likely correspond to comprehensive policies on high-value vehicles.

------------------------------------------------------------------------

### Claims and Incurred Cost Distribution

```{r}
#| fig-width: 12
#| fig-height: 8

p14 <- insurance |>
  count(total_claims) |>
  mutate(pct = n / sum(n)) |>
  ggplot(aes(x = factor(total_claims), y = n)) +
  geom_col(fill = "#2166AC") +
  geom_text(aes(label = paste0(round(pct * 100, 1), "%")),
            vjust = -0.3, size = 3) +
  scale_y_continuous(labels = comma) +
  labs(title = "Distribution of Total Claims per Policy",
       x = "Number of Claims", y = "Count") +
  theme_minimal()

p15 <- insurance |>
  filter(total_incurred > 0, total_incurred <= 50000) |>
  ggplot(aes(x = total_incurred)) +
  geom_histogram(bins = 60, fill = "#2166AC", colour = "white") +
  scale_x_continuous(labels = comma) +
  scale_y_continuous(labels = comma) +
  labs(title = "Total Incurred Cost (Claimants Only)",
       subtitle = "Capped at EUR 50k to show distribution shape",
       x = "Total Incurred (EUR)", y = "Count") +
  theme_minimal()

p16 <- insurance |>
  filter(total_incurred > 0) |>
  ggplot(aes(x = total_incurred)) +
  geom_histogram(bins = 60, fill = "#2166AC", colour = "white") +
  scale_x_continuous(trans = "log10", labels = comma) +
  scale_y_continuous(labels = comma) +
  labs(title = "Total Incurred Cost (log10 scale)",
       x = "Total Incurred (EUR, log scale)", y = "Count") +
  theme_minimal()

p14 / (p15 | p16)
```

**Observation:** The claims distribution is extremely sparse. Approximately 88% of policies have zero claims, consistent with low-frequency, high-severity insurance loss patterns. Among policies with claims, the incurred cost distribution is heavily right-skewed with most losses below €5,000 but a long tail. On the log scale, incurred costs appear roughly log-normal, which is characteristic of motor insurance loss distributions.

------------------------------------------------------------------------

### Exposure Distribution

```{r}
#| fig-width: 8
#| fig-height: 4

insurance |>
  ggplot(aes(x = total_exposure)) +
  geom_histogram(bins = 50, fill = "#2166AC", colour = "white") +
  scale_y_continuous(labels = comma) +
  labs(title = "Distribution of Total Exposure (Policy Years)",
       x = "Exposure (policy years)", y = "Count") +
  theme_minimal()
```

**Observation:** The exposure distribution is strongly bimodal. The large spike near 1.0 represents policies active for a full year. A secondary mode near 0.1 represents short-term or recently cancelled policies. Very small exposure values correspond to policies with minimal time on risk.

------------------------------------------------------------------------

### Vehicle Brand Distribution (Top 15)

```{r}
#| fig-width: 10
#| fig-height: 5

insurance |>
  count(vehicle_brand) |>
  mutate(pct = n / sum(n)) |>
  slice_max(n, n = 15) |>
  mutate(vehicle_brand = fct_reorder(vehicle_brand, n)) |>
  ggplot(aes(x = vehicle_brand, y = n, fill = n)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = paste0(round(pct*100, 1), "%")),
            hjust = -0.1, size = 3.2) +
  coord_flip() +
  scale_y_continuous(labels = comma,
                     expand = expansion(mult = c(0, 0.18))) +
  scale_fill_gradient(low = "#BBDEFB", high = "#1565C0") +
  labs(title = "Top 15 Vehicle Brands by Policy Count",
       x = NULL, y = "Number of Policies") +
  theme_minimal() +
  theme(panel.grid.major.y = element_blank())
```

**Observation:** The portfolio is dominated by mainstream European brands, with Renault, Citroen and Peugeot accounting for the top three positions. Volkswagen and Ford follow closely. This brand distribution reflects the broader Spanish vehicle market. The concentration in mass-market brands suggests relatively homogeneous vehicle values and repair costs across the portfolio, though premium brands in the tail can generate disproportionately large claims.

------------------------------------------------------------------------

### Driving Experience Distribution

```{r}
#| fig-width: 9
#| fig-height: 4

insurance |>
  filter(!is.na(driving_experience),
         driving_experience >= 0,
         driving_experience <= 60) |>
  ggplot(aes(x = driving_experience)) +
  geom_histogram(binwidth = 3, fill = "#2166AC", colour = "white") +
  scale_y_continuous(labels = comma) +
  labs(title = "Distribution of Driving Experience (Years Since Licence)",
       subtitle = "Capped at 60 years for clarity",
       x = "Years of Driving Experience", y = "Count") +
  theme_minimal()
```

**Observation:** Driving experience follows a right-skewed distribution with most policyholders having between 10 and 40 years of experience, consistent with the middle-aged driver profile seen in the driver age distribution. Very few policyholders have less than 5 years of experience, confirming that novice drivers are largely absent from this portfolio.

------------------------------------------------------------------------

### Power-to-Weight Ratio Distribution

```{r}
#| fig-width: 9
#| fig-height: 4

insurance |>
  filter(!is.na(power_to_weight_ratio),
         power_to_weight_ratio > 0,
         power_to_weight_ratio <= quantile(power_to_weight_ratio,
                                            0.99, na.rm = TRUE)) |>
  ggplot(aes(x = power_to_weight_ratio)) +
  geom_histogram(bins = 50, fill = "#2166AC", colour = "white") +
  scale_y_continuous(labels = comma) +
  labs(title = "Distribution of Power-to-Weight Ratio",
       subtitle = "Capped at 99th percentile",
       x = "Power-to-Weight Ratio (kg/hp)", y = "Count") +
  theme_minimal()
```

**Observation:** The power-to-weight ratio distribution is right-skewed with most vehicles clustering in the lower range, reflecting the dominance of standard passenger vehicles. A smaller tail of high-performance vehicles with elevated ratios represents a potential risk segment, as higher-powered vehicles are associated with greater accident severity.

## EDA 3. Univariate Analysis of Derived Profitability Metrics

### Loss Ratio Distribution

```{r}
#| fig-width: 12
#| fig-height: 5

lr_data <- insurance |>
  filter(!is.na(loss_ratio), total_premium > 0)

# Left: zero vs non-zero split
p17a <- lr_data |>
  mutate(claim_group = ifelse(loss_ratio == 0,
                               "No Claims (LR = 0)",
                               "Had Claims (LR > 0)")) |>
  count(claim_group) |>
  mutate(pct = n / sum(n),
         claim_group = fct_rev(factor(claim_group))) |>
  ggplot(aes(x = claim_group, y = pct, fill = claim_group)) +
  geom_col(show.legend = FALSE, width = 0.5) +
  geom_text(aes(label = paste0(round(pct*100, 1), "%")),
            vjust = -0.4, size = 4.5, fontface = "bold") +
  scale_y_continuous(labels = percent,
                     expand = expansion(mult = c(0, 0.12))) +
  scale_fill_manual(values = c("No Claims (LR = 0)"  = "#90CAF9",
                                "Had Claims (LR > 0)" = "#2166AC")) +
  labs(title = "Claim Incidence",
       x = NULL, y = "% of Policies") +
  theme_minimal() +
  theme(panel.grid.major.x = element_blank())

# Right: histogram of claimants only, filtered before binning
p17b <- lr_data |>
  filter(loss_ratio > 0, loss_ratio <= 3) |>
  ggplot(aes(x = loss_ratio)) +
  geom_histogram(binwidth = 0.05, fill = "#2166AC",
                 colour = "white", linewidth = 0.2) +
  geom_vline(xintercept = 1, colour = "#B71C1C",
             linetype = "dashed", linewidth = 0.8) +
  annotate("text", x = 1.05, y = Inf,
           label = "Break-even", hjust = 0, vjust = 1.8,
           colour = "#B71C1C", size = 3.5) +
  scale_x_continuous(breaks = c(0, 0.5, 1, 1.5, 2, 2.5, 3)) +
  scale_y_continuous(labels = comma) +
  labs(title = "Loss Distribution (Claimants Only)",
       subtitle = "Filtered to LR <= 3 before binning",
       x = "Loss Ratio", y = "Count") +
  theme_minimal()

p17a | p17b
```

```{r}
lr_data |>
  summarise(
    `Mean LR`              = round(mean(loss_ratio), 3),
    `Median LR`            = round(median(loss_ratio), 3),
    `% Policies LR = 0`    = round(mean(loss_ratio == 0) * 100, 1),
    `% Policies LR > 1`    = round(mean(loss_ratio > 1) * 100, 1),
    `% Policies LR > 2`    = round(mean(loss_ratio > 2) * 100, 1)
  ) |>
  pivot_longer(everything(), names_to = "Metric", values_to = "Value") |>
  kable(caption = "Loss Ratio Summary Statistics") |>
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

**Observation:** The two-panel chart reveals the key structural feature of this insurance portfolio. Over 88% of policies never generate a claim. Among the 12% that do claim, most have loss ratios below 1, meaning premiums still covered the cost. The right panel shows a peak near 0.1–0.3 that decays toward 3, consistent with a log-normal claim severity distribution. A small fraction of policies breach the break-even line and disproportionately drive aggregate losses.

------------------------------------------------------------------------

### Claim Frequency Distribution

```{r}
#| fig-width: 8
#| fig-height: 4

insurance |>
  filter(!is.na(claim_frequency), total_exposure > 0,
         claim_frequency > 0, claim_frequency <= 5) |>
  ggplot(aes(x = claim_frequency)) +
  geom_histogram(bins = 40, fill = "#FF7043", colour = "white") +
  scale_y_continuous(labels = comma) +
  labs(title = "Claim Frequency Distribution (Claimants Only)",
       subtitle = "Filtered to policies with at least one claim, capped at 5 claims/year",
       x = "Claim Frequency (claims per policy year)", y = "Count") +
  theme_minimal()
```

**Observation:** Among policies that filed claims, most had a frequency near 0.1–0.5 claims per year, consistent with a single claim on a full-year policy. The rare policies with high claim frequencies (\>2 per year) represent a significant volatility risk, indicating accident-prone drivers.

## EDA 4. Bivariate Analysis of Factors Affecting Profitability

### Loss Ratio by Policy Type

```{r}
#| fig-width: 10
#| fig-height: 5

insurance |>
  filter(!is.na(loss_ratio), total_premium > 0,
         loss_ratio > 0, loss_ratio <= 2) |>
  mutate(fill_col = ifelse(policy_type %in% c("COMP/E: +Excess","COMP/N: Full"),
                            "#2166AC", "#BABABA")) |>
  ggplot(aes(x = policy_type, y = loss_ratio, fill = fill_col)) +
  stat_halfeye(adjust = 0.5, justification = -0.15,
               .width = 0, point_colour = NA, show.legend = FALSE) +
  geom_boxplot(width = 0.15, outlier.shape = NA,
               show.legend = FALSE, colour = "#444444", linewidth = 0.5) +
  stat_summary(fun = mean, geom = "point",
               shape = 18, size = 3, colour = "black") +
  geom_hline(yintercept = 1, colour = "#D6604D",
             linetype = "dashed", linewidth = 0.8) +
  annotate("text", x = 0.6, y = 1.05,
           label = "Break-even", hjust = 0,
           colour = "#D6604D", size = 3.5) +
  coord_cartesian(ylim = c(0, 2)) +
  scale_fill_identity() +
  labs(title = "Loss Ratio by Policy Type, Raincloud Plot",
       subtitle = "Claimants only (LR > 0), capped at 2 | Highlighted: COMP/E and COMP/N | Diamond = mean",
       x = "Policy Type", y = "Loss Ratio") +
  theme_minimal()
```

**Observation:** Loss ratio distributions vary markedly across policy types. Comprehensive policies (COMP/E: +Excess and COMP/N: Full) show the widest distributions and highest medians, reflecting the broader coverage they provide. The TP: Basic and TPG: +Glass policies are concentrated near zero as their coverage is limited to third-party liability only.

------------------------------------------------------------------------

### Loss Ratio by Bonus Score

```{r}
#| fig-width: 10
#| fig-height: 5

insurance |>
  filter(!is.na(loss_ratio), total_premium > 0,
         loss_ratio > 0, loss_ratio <= 1.5) |>
  mutate(fill_col = case_when(
    bonus_score == "Bad"     ~ "#2166AC",
    bonus_score == "Neutral" ~ "#878787",
    bonus_score == "Good"    ~ "#BABABA"
  )) |>
  ggplot(aes(x = bonus_score, y = loss_ratio, fill = fill_col)) +
  stat_halfeye(adjust = 0.5, justification = -0.15,
               .width = 0, point_colour = NA, show.legend = FALSE) +
  geom_boxplot(width = 0.15, outlier.shape = NA,
               show.legend = FALSE, fill = "white",
               colour = "#444444", linewidth = 0.6) +
  stat_summary(fun = mean, geom = "point",
               shape = 18, size = 3, colour = "black") +
  geom_hline(yintercept = 1, colour = "#D6604D",
             linetype = "dashed", linewidth = 0.8) +
  annotate("text", x = 0.6, y = 1.05,
           label = "Break-even (LR = 1)", hjust = 0,
           colour = "#D6604D", size = 3.5) +
  scale_fill_identity() +
  labs(title = "Loss Ratio by Bonus Score, Raincloud Plot",
       subtitle = "Claimants only (LR > 0), capped at 1.5 | Highlighted: Bad-score policyholders | Diamond = mean",
       x = "Bonus Score", y = "Loss Ratio",
       caption = "< 3% of claimants exceed LR = 1.5 and are excluded") +
  theme_minimal()
```

**Observation:** Policyholders with a "Bad" bonus score exhibit substantially higher median loss ratios and greater distributional spread compared to "Good" policyholders, confirming that the bonus-malus system successfully identifies high-risk customers. The "Bad" group contributes disproportionately to portfolio volatility despite representing less than 1% of policies.

------------------------------------------------------------------------

### Claim Frequency by Driver Age Group

```{r}
#| fig-width: 10
#| fig-height: 5

insurance |>
  filter(!is.na(claim_frequency), total_exposure > 0) |>
  mutate(age_group = cut(driver_age,
                          breaks = c(18, 25, 35, 45, 55, 65, 100),
                          labels = c("19-25", "26-35", "36-45",
                                     "46-55", "56-65", "66+"),
                          right = TRUE)) |>
  filter(!is.na(age_group)) |>
  group_by(age_group) |>
  summarise(mean_freq = mean(claim_frequency), .groups = "drop") |>
  mutate(dot_col = ifelse(age_group == "19-25", "#2166AC", "#BABABA")) |>
  ggplot(aes(x = age_group, y = mean_freq)) +
  geom_segment(aes(xend = age_group, yend = 0.22),
               colour = "#BABABA", linewidth = 0.8) +
  geom_point(aes(colour = dot_col), size = 6, show.legend = FALSE) +
  geom_text(aes(label = round(mean_freq, 3)),
            vjust = -1.0, size = 3.8, fontface = "bold") +
  scale_colour_identity() +
  scale_y_continuous(limits = c(0.22, 0.85)) +
  labs(title = "Mean Claim Frequency by Driver Age Group",
       subtitle = "Highlighted: 19-25 group (0.716) files 2.5x more claims than safest 56-65 group (0.289)",
       x = "Driver Age Group", y = "Mean Claim Frequency (claims/policy year)") +
  theme_minimal() +
  theme(panel.grid.major.x = element_blank())
```

**Observation:** Claim frequency is highest for the youngest driver age group (19–25), consistent with actuarial research on inexperienced drivers. Frequency gradually decreases for middle-aged drivers (35–55) and slightly increases again for older drivers (65+), forming a characteristic U-shape. This pattern justifies age-based premium differentiation.

------------------------------------------------------------------------

### Claim Frequency by Driving Experience Group

```{r}
#| fig-width: 10
#| fig-height: 5

insurance |>
  filter(!is.na(claim_frequency), total_exposure > 0,
         driving_experience >= 0, driving_experience <= 60) |>
  mutate(exp_group = cut(driving_experience,
                          breaks = c(-1, 5, 10, 20, 30, 40, 60),
                          labels = c("0-5", "6-10", "11-20",
                                     "21-30", "31-40", "41+"))) |>
  filter(!is.na(exp_group)) |>
  group_by(exp_group) |>
  summarise(mean_freq = mean(claim_frequency), .groups = "drop") |>
  ggplot(aes(x = exp_group, y = mean_freq)) +
  geom_col(fill = "#2166AC", width = 0.65) +
  geom_text(aes(label = round(mean_freq, 3)),
            vjust = -0.4, size = 3.5) +
  coord_cartesian(ylim = c(0, 0.5)) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
  labs(title = "Mean Claim Frequency by Driving Experience Group",
       x = "Years of Driving Experience", y = "Mean Claim Frequency (claims/policy year)") +
  theme_minimal() +
  theme(panel.grid.major.x = element_blank())
```

**Observation:** Claim frequency decreases consistently as driving experience increases, with the least experienced drivers (0–5 years) showing the highest frequency. This pattern reinforces the driver age findings and confirms that experience, independent of age, is a meaningful risk predictor. The most experienced drivers (41+ years) show the lowest claim frequency, though this group also tends to be older.

------------------------------------------------------------------------

### Loss Ratio by Vehicle Age Group

```{r}
#| fig-width: 10
#| fig-height: 5

insurance |>
  filter(!is.na(loss_ratio), total_premium > 0,
         loss_ratio > 0, loss_ratio <= 1.5) |>
  mutate(veh_age_group = cut(vehicle_age,
                              breaks = c(-1, 5, 10, 15, 20, 30, 60),
                              labels = c("0-5", "6-10", "11-15",
                                         "16-20", "21-30", "31+")),
         fill_col = case_when(
           veh_age_group == "0-5"   ~ "#2166AC",
           veh_age_group == "6-10"  ~ "#4393C3",
           veh_age_group == "11-15" ~ "#74ADD1",
           TRUE ~ "#BABABA"
         )) |>
  filter(!is.na(veh_age_group)) |>
  ggplot(aes(x = veh_age_group, y = loss_ratio, fill = fill_col)) +
  stat_halfeye(adjust = 0.5, justification = -0.15,
               .width = 0, point_colour = NA, show.legend = FALSE) +
  geom_boxplot(width = 0.15, outlier.shape = NA,
               show.legend = FALSE, fill = "white",
               colour = "#444444", linewidth = 0.6) +
  stat_summary(fun = mean, geom = "point",
               shape = 18, size = 3, colour = "black") +
  geom_hline(yintercept = 1, colour = "#D6604D",
             linetype = "dashed", linewidth = 0.8) +
  scale_fill_identity() +
  labs(title = "Loss Ratio by Vehicle Age Group, Raincloud Plot",
       subtitle = "Claimants only (LR > 0), capped at 1.5 | Highlighted: newer vehicles (0-15 yrs) | Diamond = mean",
       x = "Vehicle Age Group (years)", y = "Loss Ratio",
       caption = "< 3% of claimants exceed LR = 1.5 and are excluded") +
  theme_minimal()
```

**Observation:** Newer vehicles (0–5 years) show wider loss ratio distributions, possibly driven by higher repair costs or declared values. Very old vehicles (31+ years) show lower and more concentrated loss ratios, likely because they carry lower insured values resulting in smaller absolute claim amounts.

------------------------------------------------------------------------

### Claim Severity by Policy Type

```{r}
#| fig-width: 10
#| fig-height: 5

insurance |>
  filter(!is.na(claim_severity), total_claims > 0) |>
  filter(claim_severity <= quantile(claim_severity, 0.99, na.rm = TRUE)) |>
  mutate(fill_col = ifelse(policy_type %in% c("COMP/E: +Excess","COMP/N: Full"),
                            "#2166AC", "#BABABA")) |>
  ggplot(aes(x = policy_type, y = claim_severity, fill = fill_col)) +
  stat_halfeye(adjust = 0.5, justification = -0.15,
               .width = 0, point_colour = NA, show.legend = FALSE) +
  geom_boxplot(width = 0.15, outlier.shape = NA,
               show.legend = FALSE, fill = "white",
               colour = "#444444", linewidth = 0.6) +
  stat_summary(fun = mean, geom = "point",
               shape = 18, size = 3, colour = "black") +
  coord_cartesian(ylim = c(0, 3000)) +
  scale_y_continuous(labels = comma) +
  scale_fill_identity() +
  labs(title = "Claim Severity by Policy Type, Raincloud Plot",
       subtitle = "Capped at 99th percentile | Policies with claims only | Highlighted: COMP/E and COMP/N | Diamond = mean",
       x = "Policy Type", y = "Claim Severity (EUR per claim)") +
  theme_minimal()
```

**Observation:** Comprehensive policies (COMP/E: +Excess and COMP/N: Full) show substantially higher claim severity compared to basic coverage types, reflecting their broader protection against costly events like theft and total loss. High variance within each policy type highlights the difficulty of predicting individual claim costs.

## EDA 5. Bivariate Analysis of Factors Affecting Volatility

### Claim Frequency by Circulation Area and Municipality Type

```{r}
#| fig-width: 9
#| fig-height: 4

insurance |>
  filter(!is.na(claim_frequency), total_exposure > 0) |>
  group_by(municipality_type, circulation_area) |>
  summarise(mean_freq = mean(claim_frequency), .groups = "drop") |>
  ggplot(aes(x = municipality_type, y = circulation_area, fill = mean_freq)) +
  geom_tile(colour = "white", linewidth = 1.5) +
  geom_text(aes(label = round(mean_freq, 3)),
            size = 5, fontface = "bold", colour = "white") +
  scale_fill_gradient(low = "#BABABA", high = "#2166AC",
                      name = "Mean Claim\nFrequency") +
  labs(title = "Mean Claim Frequency by Municipality and Circulation Area, Heatmap",
       subtitle = "Darker blue = higher claim frequency | Inland Urban (0.371) is highest risk combination",
       x = "Municipality Type", y = "Circulation Area") +
  theme_minimal() +
  theme(panel.grid = element_blank())
```

**Observation:** Urban circulation consistently shows higher claim frequencies than rural across all municipality types, consistent with greater traffic density and accident risk in urban environments. Island municipalities exhibit the highest urban claim frequencies. This geographic segmentation demonstrates that both location type and circulation environment are meaningful predictors of claim volatility.

------------------------------------------------------------------------

### Claim Frequency by Vehicle Brand (Top 12)

```{r}
#| fig-width: 10
#| fig-height: 5

insurance |>
  filter(!is.na(claim_frequency), total_exposure > 0) |>
  group_by(vehicle_brand) |>
  summarise(
    mean_freq = mean(claim_frequency),
    n = n(),
    .groups = "drop"
  ) |>
  filter(n >= 500) |>
  slice_max(mean_freq, n = 12) |>
  mutate(vehicle_brand = fct_reorder(vehicle_brand, mean_freq)) |>
  ggplot(aes(x = vehicle_brand, y = mean_freq, fill = mean_freq)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = round(mean_freq, 3)),
            hjust = -0.1, size = 3.2) +
  coord_flip() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.18))) +
  scale_fill_gradient(low = "#90CAF9", high = "#1565C0") +
  labs(title = "Mean Claim Frequency by Vehicle Brand (Top 12)",
       subtitle = "Brands with at least 500 policies only",
       x = NULL, y = "Mean Claim Frequency (claims/policy year)") +
  theme_minimal() +
  theme(panel.grid.major.y = element_blank())
```

**Observation:** Claim frequency varies meaningfully across vehicle brands, indicating that brand is a relevant risk factor beyond what is captured by vehicle age and value alone. This variation may reflect differences in driver demographics, vehicle use patterns, or inherent vehicle safety characteristics across brands.

------------------------------------------------------------------------

### Loss Ratio Over Time (Year)

```{r}
#| fig-width: 10
#| fig-height: 5

trend_data <- insurance |>
  filter(!is.na(loss_ratio), total_premium > 0) |>
  group_by(year, policy_type) |>
  summarise(mean_lr = mean(loss_ratio), .groups = "drop")

ggplot(trend_data,
       aes(x = year, y = mean_lr,
           colour = policy_type, group = policy_type,
           linewidth = ifelse(policy_type %in% c("COMP/E: +Excess","COMP/N: Full"), 1.3, 0.7))) +
  geom_line() +
  geom_point(size = 3) +
  geom_hline(yintercept = 1, linetype = "dashed",
             colour = "#D6604D", linewidth = 0.8) +
  geom_text_repel(
    data = filter(trend_data, year == 2024),
    aes(label = paste0(policy_type, ": ", round(mean_lr, 2))),
    direction   = "y",
    nudge_x     = 0.12,
    size        = 3,
    show.legend = FALSE,
    segment.color = "grey70",
    segment.size  = 0.3
  ) +
  scale_x_continuous(breaks = c(2022, 2023, 2024),
                     expand = expansion(add = c(0.2, 0.9))) +
  scale_linewidth_identity() +
  scale_colour_manual(values = c(
    "TP: Basic" = "#BABABA", "TPG: +Glass" = "#BABABA",
    "CC: +Add-ons" = "#BABABA", "COMP/E: +Excess" = "#2166AC",
    "COMP/N: Full" = "#2166AC")) +
  labs(title = "Mean Loss Ratio by Policy Type Over Time (2022-2024)",
       subtitle = "COMP/E and COMP/N crossed break-even in 2024",
       x = "Year", y = "Mean Loss Ratio", colour = "Policy Type") +
  theme_minimal() +
  theme(legend.position = "none")
```

**Observation:** Loss ratios vary across years and policy types. Comprehensive policies consistently sit above basic coverage types throughout the three-year period. Any upward trend could indicate inflationary pressure on claims costs such as rising repair or medical expenses.

------------------------------------------------------------------------

### Relationship Between Premium and Incurred Cost

```{r}
#| fig-width: 9
#| fig-height: 5

insurance |>
  filter(total_incurred > 0, total_premium > 0) |>
  filter(total_premium  <= quantile(total_premium, 0.99),
         total_incurred <= quantile(total_incurred, 0.99)) |>
  ggplot(aes(x = total_premium, y = total_incurred,
             colour = policy_type)) +
  geom_point(alpha = 0.15, size = 0.8) +
  geom_abline(slope = 1, intercept = 0, colour = "#B71C1C",
              linetype = "dashed", linewidth = 0.8) +
  scale_x_continuous(labels = comma) +
  scale_y_continuous(labels = comma) +
  scale_colour_manual(values = c(
    "TP: Basic" = "#BABABA", "TPG: +Glass" = "#BABABA",
    "CC: +Add-ons" = "#BABABA", "COMP/E: +Excess" = "#2166AC",
    "COMP/N: Full" = "#2166AC")) +
  labs(title = "Premium vs Incurred Cost (Policies with Claims)",
       subtitle = "Red dashed line = break-even (premium = incurred cost)",
       x = "Total Premium (EUR)", y = "Total Incurred (EUR)",
       colour = "Policy Type") +
  theme_minimal() +
  guides(colour = guide_legend(override.aes = list(alpha = 1, size = 2)))
```

**Observation:** Most policies with claims fall below the break-even line (incurred \< premium), indicating individual profitability. However, a notable cluster of policies exceed the break-even line, representing loss-making contracts. These tend to be clustered among comprehensive policies, confirming that wider coverage generates both higher revenue and higher claims exposure.

## Summary and Conclusion

This analysis explored the motor vehicle insurance portfolio across 354,140 policies spanning 2022–2024 using univariate and bivariate EDA methods. All 47 variables were examined across 5 EDA sections.

**Key findings on profitability:**

- The portfolio is dominated by CC: +Add-ons and COMP/E: +Excess policy types which carry the highest claim severity
- Bonus score (claims history) is a strong predictor of loss ratio. "Bad" policyholders show markedly higher and more variable loss ratios
- Comprehensive policies (COMP/E: +Excess, COMP/N: Full) consistently exhibit higher loss ratios than basic coverage types, and both crossed break-even in 2024
- Over 88% of policies generate zero claims, creating a highly skewed loss distribution that drives average loss ratios

**Key findings on volatility:**

- Young drivers (19–25) and drivers with less than 5 years of experience show the highest claim frequencies
- Urban circulation and island municipalities show the highest claim frequencies across all regions
- Vehicle age, municipality type and circulation area all contribute to claim frequency variation
- Vehicle brand shows meaningful variation in claim frequency, indicating that brand is a risk factor beyond age and value alone
- The heavy right tail in both claim frequency and severity distributions is the primary source of portfolio volatility

## References

- Espinosa, P., Lledó, J., & Atance, D. (2026). *A detailed dataset of motor insurance policies with coverage-specific financial information*. Mendeley Data. https://doi.org/10.17632/sw4jmdb2sm.1
- Swiss Re Institute. (2019). *sigma 4/2019: World insurance: the great pivot east continues*. Swiss Re.
- Wickham, H. (2016). *ggplot2: Elegant Graphics for Data Analysis*. Springer-Verlag New York.
- Wilke, C. O. (2019). *Fundamentals of Data Visualization*. O'Reilly Media.
