---
title: "Motor Vehicle Insurance Portfolio"
subtitle: "Visual Analytics for Profitability & Volatility Discovery"
author: "ZeXin Chen"
date: "2026-05-14"
format:
  revealjs:
    theme: [default, custom.scss]
    slide-number: true
    transition: slide
    background-transition: fade
    show-notes: true
    footer: "ISSS608 Take-home Exercise 1 | ZeXin Chen | [↩ Back to Site](https://isss-608-vaa-eta.vercel.app)"
execute:
  echo: false
  eval: true
  warning: false
  message: false
  freeze: false
---

```{r}
#| label: setup
pacman::p_load(
  tidyverse, patchwork, ggdist,
  ggridges, ggthemes, colorspace,
  scales, knitr, kableExtra, ggrepel
)

insurance <- read_delim(
  "data/Dataset of motor insurance portfolio.csv",
  delim = ";",
  locale = locale(decimal_mark = ".")
) |>
  mutate(
    policy_type       = factor(policy_type,
                                levels = c("TP","TPG","CC","COMP_E","COMP_N"),
                                labels = c("TP: Basic",
                                           "TPG: +Glass",
                                           "CC: +Add-ons",
                                           "COMP/E: +Excess",
                                           "COMP/N: Full")),
    policy_status     = factor(policy_status, levels = c("A","C"),
                                labels = c("Active","Cancelled")),
    business_type     = factor(business_type, levels = c("NB","P"),
                                labels = c("New Business","Portfolio")),
    payment_frequency = factor(payment_frequency, levels = c("A","S","Q"),
                                labels = c("Annual","Semiannual","Quarterly")),
    bonus_score       = factor(bonus_score, levels = c("G","N","B"),
                                labels = c("Good","Neutral","Bad")),
    fuel_type         = factor(fuel_type, levels = c("D","G"),
                                labels = c("Diesel","Gasoline")),
    municipality_type = factor(municipality_type, levels = c("I","C","IS"),
                                labels = c("Inland","Coastal","Islands")),
    circulation_area  = factor(circulation_area, levels = c("U","R"),
                                labels = c("Urban","Rural")),
    year              = as.integer(year),
    loss_ratio        = ifelse(total_premium > 0,
                                total_incurred / total_premium, NA),
    claim_frequency   = ifelse(total_exposure > 0,
                                total_claims / total_exposure, NA),
    claim_severity    = ifelse(total_claims > 0,
                                total_incurred / total_claims, NA)
  )

# Consistent colour palette, blue highlight, grey for rest
clr_blue  <- "#2166AC"
clr_grey1 <- "#878787"
clr_grey2 <- "#BABABA"
clr_grey3 <- "#E0E0E0"
clr_red   <- "#D6604D"
clr_white <- "#FFFFFF"
```

## Contents {.smaller}

::::: columns
::: {.column width="50%"}
**Part 1: Dataset Overview**

- About the data
- Key variables and derived metrics

**Part 2: Portfolio Composition**

- Policy and geographic breakdown
- Driver and vehicle profile
:::

::: {.column width="50%"}
**Part 3: Profitability Analysis**

- Loss ratio distributions
- Key drivers of profitability

**Part 4: Volatility Analysis**

- Claim frequency patterns
- Geographic and temporal effects
:::
:::::

::: notes
Good morning everyone. Today I will be walking you through an exploratory visual analysis of a real motor insurance portfolio from Spain. The dataset covers three years of policy transactions and gives us a great opportunity to look at what actually drives profitability and claims volatility. We will move through four parts: dataset overview, portfolio composition, profitability analysis, and volatility analysis.
:::

# Part 1: Dataset Overview

## About the Dataset

:::::: columns
::: {.column width="55%"}
**Source:** Spanish motor insurer microdata\
**Period:** January 2022 – December 2024\
**Records:** 354,140 policy transactions\
**Variables:** 47 across 4 groups

| Group             | Variables |
|-------------------|-----------|
| Policy & Insured  | 8         |
| Driver & Vehicle  | 10        |
| Premiums          | 8         |
| Claims & Incurred | 21        |
:::

:::: {.column width="45%"}
**Three derived metrics guide analysis:**

$$\text{Loss Ratio} = \frac{\text{Incurred}}{\text{Premium}}$$

$$\text{Claim Freq} = \frac{\text{Claims}}{\text{Exposure}}$$

$$\text{Severity} = \frac{\text{Incurred}}{\text{Claims}}$$

::: callout-note
Loss Ratio \< 1 = **Profitable**\
Loss Ratio \> 1 = **Unprofitable**
:::
::::
::::::

::: notes
This is real anonymised data from a Spanish motor insurer. Each row is one policy in one year. Loss ratio is our profitability measure, below 1 means profitable. Claim frequency tells us how often accidents happen per year of exposure. Severity tells us the average cost per claim when one does occur.
:::

## Policy Type Reference Guide

| Code | Full Name | Coverage |
|----|----|----|
| **TP** | Third-Party Only | Liability to others only |
| **TPG** | Third-Party + Glass | Liability + windscreen |
| **CC** | Third-Party + Add-ons | Liability + theft, fire, glass, legal |
| **COMP/E** | Comprehensive with Excess | Full coverage, policyholder pays excess |
| **COMP/N** | Full Comprehensive | Full coverage, no excess |

::: callout-note
**CC, COMP/E and COMP/N** are the three broadest coverage types and together account for **88.2%** of the portfolio. These are the focus of our profitability analysis.
:::

::: notes
Before we look at charts with policy type codes, here is a quick reference guide. TP and TPG are basic policies covering only third-party liability. CC adds several extras like theft, fire and glass on top of liability. COMP with excess means full comprehensive coverage but the policyholder pays the first portion of any claim. Full comprehensive has no such excess. The three broadest types together account for 88.2 percent of all policies.
:::

## Data Quality

```{r}
#| fig-width: 9
#| fig-height: 3.5
#| fig-align: center

data.frame(
  variable    = c("vehicle_value", "fuel_type"),
  missing_pct = c(0.14, 0.37)
) |>
  ggplot(aes(x = reorder(variable, missing_pct),
             y = missing_pct)) +
  geom_col(fill = clr_blue, width = 0.45) +
  geom_text(aes(label = paste0(missing_pct, "%")),
            hjust = -0.15, size = 5.5, fontface = "bold",
            colour = clr_blue) +
  coord_flip(ylim = c(0, 0.5)) +
  scale_y_continuous(labels = function(x) paste0(x, "%")) +
  labs(title = "Dataset is Exceptionally Clean",
       subtitle = "Only 2 of 47 variables have missing values, both under 0.4%",
       x = NULL, y = "Missing (%)") +
  theme_minimal(base_size = 14) +
  theme(panel.grid.major.y = element_blank(),
        plot.title = element_text(face = "bold"))
```

- `fuel_type`: **0.37%** missing, excluded when fuel type is analysed
- `vehicle_value`: **0.14%** missing, excluded when vehicle value is analysed
- All remaining **45 variables** are 100% complete

::: notes
Only two of 47 variables have any missing values at all. Fuel type is missing for 0.37 percent and vehicle value for 0.14 percent. Both are handled by exclusion. Everything else is complete, which is exceptional for a real-world dataset of this size.
:::

# Part 2: Portfolio Composition

## Insight 1: CC Policies Dominate at 57.8%

```{r}
#| fig-width: 10
#| fig-height: 4.2
#| fig-align: center

insurance |>
  count(policy_type) |>
  mutate(pct = n / sum(n),
         policy_type = fct_reorder(policy_type, n),
         fill_col = ifelse(policy_type %in% c("CC: +Add-ons",
                                               "COMP/E: +Excess",
                                               "COMP/N: Full"),
                            clr_blue, clr_grey2)) |>
  ggplot(aes(x = policy_type, y = n, fill = fill_col)) +
  geom_col(show.legend = FALSE, width = 0.65) +
  geom_text(aes(label = paste0(round(pct*100,1), "%")),
            hjust = -0.05, size = 5, fontface = "bold") +
  coord_flip() +
  scale_y_continuous(labels = comma,
                     expand = expansion(mult = c(0, 0.25))) +
  scale_fill_identity() +
  labs(title = "Policy Type Distribution (n = 354,140)",
       subtitle = "Blue: three broadest coverage types (88.2% of portfolio)",
       x = NULL, y = "Number of Policies") +
  theme_minimal(base_size = 14) +
  theme(panel.grid.major.y = element_blank(),
        plot.title = element_text(face = "bold"),
        plot.subtitle = element_text(colour = "#555555"))
```

- **CC: +Add-ons** (third-party + theft, fire, glass) dominates at **57.8%** of all policies
- **COMP/E: +Excess** (comprehensive with excess) is second at **25.8%**
- **TP: Basic** (third-party only) is rare at just **1.6%**, most customers prefer broader coverage

::: notes
CC policies dominate at 57.8 percent. These combine third-party liability with add-ons like theft and glass cover. COMP/E at 25.8 percent is full comprehensive with an excess. What stands out is how rare basic TP policies are at just 1.6 percent. This tells us the insurer is largely serving customers who want broader protection.
:::

## Insight 2: 89.2% Active Policies, 96.4% Good Bonus Score

```{r}
#| fig-width: 11
#| fig-height: 4.2
#| fig-align: center

p_status <- insurance |>
  count(policy_status) |>
  mutate(pct = n / sum(n),
         fill_col = ifelse(policy_status == "Active", clr_blue, clr_grey2)) |>
  ggplot(aes(x = policy_status, y = n, fill = fill_col)) +
  geom_col(show.legend = FALSE, width = 0.5) +
  geom_text(aes(label = paste0(round(pct*100,1), "%")),
            vjust = -0.5, size = 6, fontface = "bold") +
  scale_y_continuous(labels = comma,
                     expand = expansion(mult = c(0, 0.15))) +
  scale_fill_identity() +
  labs(title = "Policy Status", x = NULL, y = "Count") +
  theme_minimal(base_size = 14) +
  theme(panel.grid.major.x = element_blank(),
        plot.title = element_text(face = "bold"))

p_bonus <- insurance |>
  count(bonus_score) |>
  mutate(pct = n / sum(n),
         fill_col = ifelse(bonus_score == "Good", clr_blue, clr_grey2)) |>
  ggplot(aes(x = bonus_score, y = n, fill = fill_col)) +
  geom_col(show.legend = FALSE, width = 0.5) +
  geom_text(aes(label = paste0(round(pct*100,1), "%")),
            vjust = -0.5, size = 6, fontface = "bold") +
  scale_y_continuous(labels = comma,
                     expand = expansion(mult = c(0, 0.15))) +
  scale_fill_identity() +
  labs(title = "Bonus Score (Claims History)",
       x = NULL, y = "Count") +
  theme_minimal(base_size = 14) +
  theme(panel.grid.major.x = element_blank(),
        plot.title = element_text(face = "bold"))

p_status | p_bonus
```

- **89.2%** active; cancellation rate of **10.8%** warrants monitoring
- **96.4%** of customers have a **Good** bonus score, excellent claims history
- Bad-score customers are only **1.3%** but represent elevated risk

::: notes
89 percent of policies are active and 11 percent cancelled. The bonus score on the right is very positive. 96 percent of customers have a Good score meaning no recent claims. This is great for profitability. The Bad group at 1.3 percent is small but important as a risk segment.
:::

## Insight 3: Middle-Aged Drivers, Older Vehicles

```{r}
#| fig-width: 11
#| fig-height: 4.2
#| fig-align: center

p_age <- insurance |>
  ggplot(aes(x = driver_age)) +
  geom_density(fill = clr_blue, colour = clr_white, alpha = 0.85) +
  geom_vline(xintercept = c(40, 60),
             colour = clr_red, linetype = "dashed", linewidth = 0.7) +
  annotate("rect", xmin = 40, xmax = 60,
           ymin = 0, ymax = Inf,
           fill = clr_red, alpha = 0.10) +
  annotate("text", x = 50, y = Inf,
           label = "Peak\n40-60 yrs",
           vjust = 1.5, size = 3.5,
           colour = clr_red, fontface = "bold") +
  labs(title = "Driver Age",
       x = "Age (years)", y = "Density") +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"))

p_vage <- insurance |>
  ggplot(aes(x = vehicle_age)) +
  geom_density(fill = clr_blue, colour = clr_white, alpha = 0.85) +
  geom_vline(xintercept = 25,
             colour = clr_red, linetype = "dashed", linewidth = 0.9) +
  annotate("text", x = 27, y = Inf,
           label = "Median\n25 yrs",
           vjust = 1.5, hjust = 0, size = 3.5,
           colour = clr_red, fontface = "bold") +
  labs(title = "Vehicle Age",
       x = "Age (years)", y = "Density") +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"))

p_val <- insurance |>
  filter(!is.na(vehicle_value)) |>
  ggplot(aes(x = vehicle_value)) +
  geom_density(fill = clr_blue, colour = clr_white, alpha = 0.85) +
  geom_vline(xintercept = 24730,
             colour = clr_red, linetype = "dashed", linewidth = 0.9) +
  annotate("text", x = 26000, y = Inf,
           label = "Median\n€24.7k",
           vjust = 1.5, hjust = 0, size = 3.5,
           colour = clr_red, fontface = "bold") +
  coord_cartesian(xlim = c(0, 80000)) +
  scale_x_continuous(labels = label_number(scale = 1e-3, suffix = "k")) +
  labs(title = "Vehicle Value (capped €80k)",
       x = "Value (EUR)", y = "Density") +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"))

p_age | p_vage | p_val
```

- Driver age peaks at **40-60 years** (mean: 49, median: 48), experienced, low-risk segment
- Vehicle age median is **25 years**, an older fleet than typical
- Vehicle value median is **€24,730**; 75th percentile at **€32,252**

::: notes
Using density curves with annotated peaks makes the distributions much clearer than histograms. The driver age peaks sharply at 40 to 60 years. Very few young drivers under 25 appear. The vehicle age median is 25 years which is quite old. Vehicle values are mostly in the 15 to 35 thousand euro range, with a long tail of high-value vehicles that can generate very large claims.
:::

## Insight 4: Inland-Dominant, Balanced Urban-Rural

```{r}
#| fig-width: 11
#| fig-height: 4.2
#| fig-align: center

p_mun <- insurance |>
  count(municipality_type) |>
  mutate(pct = n / sum(n),
         municipality_type = fct_reorder(municipality_type, n),
         fill_col = ifelse(municipality_type == "Inland",
                            clr_blue, clr_grey2)) |>
  ggplot(aes(x = municipality_type, y = pct, fill = fill_col)) +
  geom_col(show.legend = FALSE, width = 0.55) +
  geom_text(aes(label = paste0(round(pct*100,1), "%")),
            vjust = -0.5, size = 5, fontface = "bold") +
  scale_y_continuous(labels = percent,
                     expand = expansion(mult = c(0, 0.15))) +
  scale_fill_identity() +
  labs(title = "Municipality Type", x = NULL, y = NULL) +
  theme_minimal(base_size = 13) +
  theme(panel.grid.major.x = element_blank(),
        plot.title = element_text(face = "bold"))

p_circ <- insurance |>
  count(circulation_area) |>
  mutate(pct = n / sum(n),
         fill_col = ifelse(circulation_area == "Urban",
                            clr_blue, clr_grey2)) |>
  ggplot(aes(x = circulation_area, y = pct, fill = fill_col)) +
  geom_col(show.legend = FALSE, width = 0.45) +
  geom_text(aes(label = paste0(round(pct*100,1), "%")),
            vjust = -0.5, size = 5, fontface = "bold") +
  scale_y_continuous(labels = percent,
                     expand = expansion(mult = c(0, 0.15))) +
  scale_fill_identity() +
  labs(title = "Circulation Area", x = NULL, y = NULL) +
  theme_minimal(base_size = 13) +
  theme(panel.grid.major.x = element_blank(),
        plot.title = element_text(face = "bold"))

p_fuel <- insurance |>
  filter(!is.na(fuel_type)) |>
  count(fuel_type) |>
  mutate(pct = n / sum(n),
         fill_col = ifelse(fuel_type == "Diesel",
                            clr_blue, clr_grey2)) |>
  ggplot(aes(x = fuel_type, y = pct, fill = fill_col)) +
  geom_col(show.legend = FALSE, width = 0.45) +
  geom_text(aes(label = paste0(round(pct*100,1), "%")),
            vjust = -0.5, size = 5, fontface = "bold") +
  scale_y_continuous(labels = percent,
                     expand = expansion(mult = c(0, 0.15))) +
  scale_fill_identity() +
  labs(title = "Fuel Type", x = NULL, y = NULL) +
  theme_minimal(base_size = 13) +
  theme(panel.grid.major.x = element_blank(),
        plot.title = element_text(face = "bold"))

p_mun | p_circ | p_fuel
```

- **64.6%** inland, **28.6%** coastal, **6.8%** islands, reflects Spain's geography
- Urban vs rural split is nearly equal at **53.8% vs 46.2%**
- **64.2%** diesel vehicles, relevant for repair cost patterns

::: notes
The geographic breakdown shows inland dominance at 64.6 percent. Urban and rural circulation are fairly balanced at 54 and 46 percent. Diesel vehicles make up 64 percent of the fleet. These splits matter because urban and rural drivers have very different claim frequencies as we will see in Part 4.
:::

# Part 3: Profitability Analysis

## Insight 5: 88.1% of Policies Have Zero Claims

```{r}
#| fig-width: 11
#| fig-height: 4.2
#| fig-align: center

lr_data <- insurance |> filter(!is.na(loss_ratio), total_premium > 0)

p_zero <- lr_data |>
  mutate(claim_group = ifelse(loss_ratio == 0,
                               "No Claims\n(LR = 0)",
                               "Had Claims\n(LR > 0)"),
         claim_group = fct_rev(factor(claim_group))) |>
  count(claim_group) |>
  mutate(pct = n / sum(n),
         fill_col = ifelse(grepl("Had", claim_group),
                            clr_blue, clr_grey2)) |>
  ggplot(aes(x = claim_group, y = pct, fill = fill_col)) +
  geom_col(show.legend = FALSE, width = 0.5) +
  geom_text(aes(label = paste0(round(pct*100,1), "%")),
            vjust = -0.5, size = 6.5, fontface = "bold") +
  scale_y_continuous(labels = percent,
                     expand = expansion(mult = c(0, 0.15))) +
  scale_fill_identity() +
  labs(title = "Claim Incidence", x = NULL, y = "% of Policies") +
  theme_minimal(base_size = 13) +
  theme(panel.grid.major.x = element_blank(),
        plot.title = element_text(face = "bold"))

p_dist <- lr_data |>
  filter(loss_ratio > 0, loss_ratio <= 3) |>
  ggplot(aes(x = loss_ratio)) +
  geom_histogram(binwidth = 0.05, fill = clr_blue,
                 colour = clr_white, linewidth = 0.2) +
  geom_vline(xintercept = 1, colour = clr_red,
             linetype = "dashed", linewidth = 1) +
  annotate("text", x = 1.05, y = Inf,
           label = "Break-even", hjust = 0, vjust = 1.8,
           colour = clr_red, size = 4, fontface = "bold") +
  scale_x_continuous(breaks = c(0, 0.5, 1, 1.5, 2, 2.5, 3)) +
  scale_y_continuous(labels = comma) +
  labs(title = "Loss Distribution (Claimants Only, LR <= 3)",
       x = "Loss Ratio", y = "Count") +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"))

p_zero | p_dist
```

- **88.1%** of policies never file a claim; only **11.9%** generate any loss
- Among claimants, distribution peaks at **LR = 0.1-0.3** and decays toward break-even
- **9.0%** of ALL policies exceed break-even (LR \> 1), these drive total portfolio losses

::: notes
Two panels tell the full story. On the left, 88.1 percent of policies never filed a single claim. On the right, we see only claimants. The distribution peaks near 0.1 to 0.3 and decays toward break-even. Just 9 percent of all policies exceed LR equals 1. These are the policies that really matter for the portfolio bottom line.
:::

## Insight 6: Comprehensive Policies Have Higher Loss Ratios {.smaller}

```{r}
#| fig-width: 10
#| fig-height: 4.5
#| fig-align: center

insurance |>
  filter(!is.na(loss_ratio), total_premium > 0,
         loss_ratio > 0, loss_ratio <= 2) |>
  mutate(fill_col = ifelse(policy_type %in% c("COMP/E: +Excess","COMP/N: Full"),
                            clr_blue, clr_grey2)) |>
  ggplot(aes(x = policy_type, y = loss_ratio, fill = fill_col)) +
  stat_halfeye(adjust = 0.5, justification = -0.15,
               .width = 0, point_colour = NA,
               show.legend = FALSE) +
  geom_boxplot(width = 0.15, outlier.shape = NA,
               show.legend = FALSE, colour = "#444444",
               linewidth = 0.5) +
  stat_summary(fun = mean, geom = "point",
               shape = 18, size = 3,
               colour = "black", show.legend = FALSE) +
  geom_hline(yintercept = 1, colour = clr_red,
             linetype = "dashed", linewidth = 0.9) +
  annotate("text", x = 0.6, y = 1.05,
           label = "Break-even", hjust = 0,
           colour = clr_red, size = 3.8, fontface = "bold") +
  coord_cartesian(ylim = c(0, 2)) +
  scale_fill_identity() +
  labs(title = "Loss Ratio by Policy Type, Raincloud Plot (Claimants Only, LR <= 2)",
       subtitle = "Highlighted: Comprehensive policies | Diamond = mean | Box = IQR",
       x = "Policy Type", y = "Loss Ratio") +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"),
        plot.subtitle = element_text(colour = "#555555"))
```

- **COMP/N: Full** mean LR = **0.950**, median = **0.897**, highest among all policy types
- **COMP/E: +Excess** mean LR = **0.942**, median = **0.915**, second highest
- **TP: Basic** mean LR = **0.864**, median = **0.676**, narrowest distribution, lowest loss ratio

::: notes
This raincloud plot from Chapter 9 combines a half-density curve, boxplot, and mean diamond marker in one chart. The blue half-densities for COMP/E and COMP/N are visibly wider and taller than the grey ones. The mean diamond sits above the median box line for all types, showing the right-skew of insurance losses. Comprehensive policies have both higher medians and wider spreads.
:::

## Insight 7: Bad Claims History Predicts Higher Loss Ratios {.smaller}

```{r}
#| fig-width: 10
#| fig-height: 4.5
#| fig-align: center

insurance |>
  filter(!is.na(loss_ratio), total_premium > 0,
         loss_ratio > 0, loss_ratio <= 1.5) |>
  mutate(fill_col = case_when(
    bonus_score == "Bad"     ~ clr_blue,
    bonus_score == "Neutral" ~ clr_grey1,
    bonus_score == "Good"    ~ clr_grey2
  )) |>
  ggplot(aes(x = bonus_score, y = loss_ratio, fill = fill_col)) +
  stat_halfeye(adjust = 0.5, justification = -0.15,
               .width = 0, point_colour = NA,
               show.legend = FALSE) +
  geom_boxplot(width = 0.15, outlier.shape = NA,
               show.legend = FALSE, colour = "#444444",
               linewidth = 0.5) +
  stat_summary(fun = mean, geom = "point",
               shape = 18, size = 3,
               colour = "black", show.legend = FALSE) +
  geom_hline(yintercept = 1, colour = clr_red,
             linetype = "dashed", linewidth = 0.9) +
  annotate("text", x = 0.6, y = 1.06,
           label = "Break-even (LR = 1)", hjust = 0,
           colour = clr_red, size = 3.8, fontface = "bold") +
  scale_fill_identity() +
  labs(title = "Loss Ratio by Bonus Score, Raincloud Plot (Claimants, LR <= 1.5)",
       subtitle = "Highlighted: Bad-score policyholders | Diamond = mean | Box = IQR",
       x = "Bonus Score (Claims History)", y = "Loss Ratio",
       caption = "< 3% of claimants exceed LR = 1.5") +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"),
        plot.subtitle = element_text(colour = "#555555"),
        plot.caption  = element_text(colour = "#888888"))
```

- **Bad-score** median LR = **0.668**, IQR = **0.31–1.05**, wide spread crossing break-even
- **Good-score** median LR = **0.724**, IQR = **0.35–1.10**, both groups show high variance
- Bad policyholders are only **1.3%** of portfolio (n=433), the half-density shape shows heavier tail risk

::: notes
The raincloud plot highlights the Bad group in blue. The median difference between Bad and Good is actually modest. What the raincloud reveals is the distribution shape, the half-density for Bad policyholders is wider with a heavier upper tail, meaning more of them have high loss ratios. Both groups have similarly wide IQRs crossing break-even. The bonus-malus system identifies customers who are more likely to generate large claims, even if the average difference is not dramatic.
:::

## Insight 8: Comprehensive Policies Cost More Per Claim {.smaller}

```{r}
#| fig-width: 10
#| fig-height: 4.5
#| fig-align: center

insurance |>
  filter(!is.na(claim_severity), total_claims > 0) |>
  filter(claim_severity <=
           quantile(claim_severity, 0.99, na.rm = TRUE)) |>
  mutate(fill_col = ifelse(policy_type %in% c("COMP/E: +Excess","COMP/N: Full"),
                            clr_blue, clr_grey2)) |>
  ggplot(aes(x = policy_type, y = claim_severity, fill = fill_col)) +
  stat_halfeye(adjust = 0.5, justification = -0.15,
               .width = 0, point_colour = NA,
               show.legend = FALSE) +
  geom_boxplot(width = 0.15, outlier.shape = NA,
               show.legend = FALSE, colour = "#444444",
               linewidth = 0.5) +
  stat_summary(fun = mean, geom = "point",
               shape = 18, size = 3,
               colour = "black", show.legend = FALSE) +
  coord_cartesian(ylim = c(0, 3000)) +
  scale_y_continuous(labels = comma) +
  scale_fill_identity() +
  labs(title = "Claim Severity by Policy Type, Raincloud Plot (Capped at 99th Percentile)",
       subtitle = "Highlighted: Comprehensive policies | Diamond = mean | Box = IQR | Y-axis capped at €3,000",
       x = "Policy Type", y = "Claim Severity (EUR per claim)") +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"),
        plot.subtitle = element_text(colour = "#555555"),
        plot.margin = margin(10, 10, 20, 10))
```

- **COMP/E: +Excess** mean severity = **€753** per claim; median = **€547**, highest among all types
- **CC: +Add-ons** mean = **€651**, median = **€437**, moderate severity
- Wide gap between mean and median across ALL types confirms heavily right-skewed distributions

::: notes
Comprehensive policies generate not only more claims but also more expensive ones. COMP_E and COMP_N show wider half-density curves at higher severity levels. The large gap between the mean diamond and the median box line across all policy types confirms that a small number of very large claims drive average severity up. This unpredictability is the core source of underwriting risk.
:::

# Part 4: Volatility Analysis

## Insight 9: Young Drivers File Claims 2.7x More Often

```{r}
#| fig-width: 10
#| fig-height: 4.5
#| fig-align: center

insurance |>
  filter(!is.na(claim_frequency), total_exposure > 0) |>
  mutate(age_group = cut(driver_age,
                          breaks = c(18,25,35,45,55,65,100),
                          labels = c("19-25","26-35","36-45",
                                     "46-55","56-65","66+"))) |>
  filter(!is.na(age_group)) |>
  group_by(age_group) |>
  summarise(mean_freq = mean(claim_frequency), .groups = "drop") |>
  mutate(fill_col = ifelse(age_group == "19-25", clr_blue, clr_grey2)) |>
  ggplot(aes(x = age_group, y = mean_freq)) +
  geom_segment(aes(xend = age_group, yend = 0.22),
               colour = clr_grey2, linewidth = 0.8) +
  geom_point(aes(colour = fill_col), size = 7, show.legend = FALSE) +
  geom_text(aes(label = round(mean_freq, 3)),
            vjust = -1.0, size = 4.2, fontface = "bold") +
  scale_colour_identity() +
  scale_y_continuous(limits = c(0.22, 0.90),
                     breaks = seq(0.2, 0.9, 0.1)) +
  labs(title = "Mean Claim Frequency by Driver Age Group",
       subtitle = "Highlighted: 19-25 age group (highest) | Youngest drivers file 2.5x more claims than 56-65 group",
       x = "Driver Age Group",
       y = "Mean Claims per Policy Year") +
  theme_minimal(base_size = 14) +
  theme(panel.grid.major.x = element_blank(),
        plot.title = element_text(face = "bold"),
        plot.subtitle = element_text(colour = "#555555"))
```

- **19-25** group: mean frequency = **0.716**, highest of all age groups
- **56-65** group: mean frequency = **0.289**, safest segment
- Young drivers file claims **2.5x more often** than the 56-65 group (0.716 / 0.289 = 2.48x)

::: notes
A lollipop chart is cleaner than a bar chart for showing this comparison, less ink, same information, the dot draws the eye straight to the value. The 19-25 group highlighted in blue stands out clearly at 0.716. Then frequency drops steadily through middle age. Young drivers file claims 2.5 times more often than the safest group. Note the 19-25 sample is relatively small as very few young drivers appear in this portfolio.
:::

## Insight 10: Urban Areas Drive Higher Claim Frequency

```{r}
#| fig-width: 10
#| fig-height: 4.5
#| fig-align: center

geo_freq <- insurance |>
  filter(!is.na(claim_frequency), total_exposure > 0) |>
  group_by(municipality_type, circulation_area) |>
  summarise(mean_freq = mean(claim_frequency), .groups = "drop")

ggplot(geo_freq, aes(x = municipality_type,
                      y = circulation_area,
                      fill = mean_freq)) +
  geom_tile(colour = "white", linewidth = 1.5) +
  geom_text(aes(label = round(mean_freq, 3)),
            size = 5.5, fontface = "bold",
            colour = "white") +
  scale_fill_gradient(low = clr_grey3, high = clr_blue,
                      name = "Mean Claim\nFrequency") +
  labs(title = "Mean Claim Frequency by Municipality & Circulation Area, Heatmap",
       subtitle = "Darker blue = higher claim frequency",
       x = "Municipality Type", y = "Circulation Area") +
  theme_minimal(base_size = 14) +
  theme(plot.title = element_text(face = "bold"),
        plot.subtitle = element_text(colour = "#555555"),
        panel.grid = element_blank(),
        legend.position = "right")
```

- **Inland Urban** = **0.371**, highest claim frequency of all 6 combinations
- Urban always exceeds rural within each municipality type, traffic density effect
- **Island Rural** = **0.246**, lowest frequency, well-insulated from traffic risk

::: notes
A heatmap from Chapter 14 is ideal here, it shows all 6 municipality-circulation combinations in one clean grid. Darker blue means higher claim frequency. Inland Urban at 0.371 is the darkest cell and the highest risk combination. Island Rural at 0.246 is the lightest. The pattern is clear: urban always exceeds rural in every row, and inland urban poses the greatest volatility risk.
:::

## Insight 11: COMP/N and COMP/E Crossed Break-Even in 2024

```{r}
#| fig-width: 10
#| fig-height: 4.5
#| fig-align: center

trend_data <- insurance |>
  filter(!is.na(loss_ratio), total_premium > 0) |>
  group_by(year, policy_type) |>
  summarise(mean_lr = mean(loss_ratio), .groups = "drop") |>
  mutate(is_comp = policy_type %in% c("COMP/E: +Excess","COMP/N: Full"))

ggplot(trend_data,
       aes(x = year, y = mean_lr, group = policy_type,
           colour = ifelse(is_comp, clr_blue, clr_grey1),
           linewidth = ifelse(is_comp, 1.5, 0.7))) +
  geom_line() +
  geom_point(size = 3.5) +
  geom_hline(yintercept = 1, linetype = "dashed",
             colour = clr_red, linewidth = 0.9) +
  annotate("text", x = 2022.05, y = 1.03,
           label = "Break-even", hjust = 0,
           colour = clr_red, size = 4, fontface = "bold") +
  geom_text_repel(
    data = filter(trend_data, year == 2024),
    aes(label = paste0(policy_type, ": ", round(mean_lr, 2))),
    direction = "y", nudge_x = 0.12,
    size = 3.8, show.legend = FALSE,
    segment.color = clr_grey2, segment.size = 0.3
  ) +
  scale_x_continuous(breaks = c(2022, 2023, 2024),
                     expand = expansion(add = c(0.2, 0.9))) +
  scale_colour_identity() +
  scale_linewidth_identity() +
  labs(title = "Mean Loss Ratio by Policy Type: 2022-2024",
       subtitle = "Highlighted: COMP/E and COMP/N crossed break-even in 2024",
       x = "Year", y = "Mean Loss Ratio") +
  theme_minimal(base_size = 14) +
  theme(plot.title    = element_text(face = "bold"),
        plot.subtitle = element_text(colour = "#555555"),
        legend.position = "none")
```

- **COMP/N: Full**: LR rose from **0.706 (2022)** → **0.944 (2023)** → **1.171 (2024)**, crossed break-even
- **COMP/E: +Excess**: LR rose from **0.641 (2022)** → **0.783 (2023)** → **1.106 (2024)**, also crossed
- **CC, TP, TPG** remain below break-even throughout, basic coverage types stay profitable

::: notes
This is the most alarming finding in the analysis. COMP/N went from 0.706 in 2022 all the way to 1.171 in 2024. COMP/E similarly crossed from 0.641 to 1.106. Both comprehensive policy types are now paying out more in claims than they collect in premiums. This could be claims inflation from rising repair or medical costs. Basic coverage types CC, TP and TPG remain comfortably below break-even throughout.
:::

## Key Findings Summary

::::: columns
::: {.column width="50%"}
**Profitability**

- 🔵 **COMP/N LR: 0.71 → 1.17** (2022-2024)
- 🔵 **COMP/E LR: 0.64 → 1.11** (2022-2024)
- ⬜ Bad-score policyholders: IQR 0.31-1.05
- ⬜ **88.1%** of policies have zero claims
:::

::: {.column width="50%"}
**Volatility**

- 🔵 **19-25 drivers**: 0.716 claims/yr (2.5x vs safest group)
- 🔵 **Inland Urban**: 0.371 claims/yr (highest region)
- ⬜ Urban always \> Rural across all municipality types
- ⬜ Wide severity IQR across all policy types
:::
:::::

::: callout-important
**Core finding:** 88.1% of policies have zero claims. The portfolio's challenge is that COMP/N and COMP/E (full comprehensive policies) have now crossed break-even, the insurer is losing money on these segments.
:::

::: notes
To summarise. The biggest story is the trend: COMP/N rose from 0.71 to 1.17 and COMP/E from 0.64 to 1.11 over three years. Both are now loss-making. On the volatility side, young drivers at 0.716 and inland urban areas at 0.371 are the highest-frequency segments. Young drivers file claims 2.5 times more than the 56-65 group. The overarching finding is that 88 percent of policies are fine. The challenge is managing the small but growing share of loss-making comprehensive policies.
:::

## References {.smaller}

- Espinosa, P., Lledó, J., & Atance, D. (2026). *A detailed dataset of motor insurance policies with coverage-specific financial information*. Mendeley Data. https://doi.org/10.17632/sw4jmdb2sm.1

- Swiss Re Institute. (2019). *sigma 4/2019: World insurance: the great pivot east continues*. Swiss Re.

- Wickham, H. (2016). *ggplot2: Elegant Graphics for Data Analysis*. Springer-Verlag New York.

- Wilke, C. O. (2019). *Fundamentals of Data Visualization*. O'Reilly Media.

- Kay, M. (2024). *ggdist: Visualizations of Distributions and Uncertainty*. R package. https://mjskay.github.io/ggdist/
