---
title: "Motor Vehicle Insurance Portfolio"
subtitle: "Visual Analytics for Profitability & Volatility Discovery"
author: "ZeXin Chen"
date: "2026-05-14"
format:
  revealjs:
    theme: [default, custom.scss]
    slide-number: true
    transition: slide
    background-transition: fade
    show-notes: true
    footer: "ISSS608 Take-home Exercise 1 | ZeXin Chen | [↩ Back to Site](https://isss-608-vaa-eta.vercel.app)"
execute:
  echo: false
  eval: true
  warning: false
  message: false
  freeze: false
---

```{r}
#| label: setup
pacman::p_load(
  tidyverse, patchwork, ggdist,
  ggridges, scales, knitr, kableExtra, ggrepel
)

insurance <- read_delim(
  "data/Dataset of motor insurance portfolio.csv",
  delim = ";",
  locale = locale(decimal_mark = ".")
) |>
  mutate(
    policy_type       = factor(policy_type,
                                levels = c("TP","TPG","CC","COMP_E","COMP_N")),
    policy_status     = factor(policy_status, levels = c("A","C"),
                                labels = c("Active","Cancelled")),
    business_type     = factor(business_type, levels = c("NB","P"),
                                labels = c("New Business","Portfolio")),
    payment_frequency = factor(payment_frequency, levels = c("A","S","Q"),
                                labels = c("Annual","Semiannual","Quarterly")),
    bonus_score       = factor(bonus_score, levels = c("G","N","B"),
                                labels = c("Good","Neutral","Bad")),
    fuel_type         = factor(fuel_type, levels = c("D","G"),
                                labels = c("Diesel","Gasoline")),
    municipality_type = factor(municipality_type, levels = c("I","C","IS"),
                                labels = c("Inland","Coastal","Islands")),
    circulation_area  = factor(circulation_area, levels = c("U","R"),
                                labels = c("Urban","Rural")),
    year              = as.integer(year),
    loss_ratio        = ifelse(total_premium > 0,
                                total_incurred / total_premium, NA),
    claim_frequency   = ifelse(total_exposure > 0,
                                total_claims / total_exposure, NA),
    claim_severity    = ifelse(total_claims > 0,
                                total_incurred / total_claims, NA)
  )

# Color palette
clr_blue   <- "#2166AC"
clr_grey1  <- "#878787"
clr_grey2  <- "#BABABA"
clr_grey3  <- "#E0E0E0"
clr_white  <- "#FFFFFF"
clr_red    <- "#D6604D"
```

## Contents {.smaller}

::::: columns
::: {.column width="50%"}
**Part 1: Dataset Overview**

- About the data
- Key variables & derived metrics

**Part 2: Portfolio Composition**

- Policy & geographic breakdown
- Driver & vehicle profile
:::

::: {.column width="50%"}
**Part 3: Profitability Analysis**

- Loss ratio distributions
- Key drivers of profitability

**Part 4: Volatility Analysis**

- Claim frequency patterns
- Geographic and temporal effects
:::
:::::

::: notes
Good morning everyone. Today I will be walking you through an exploratory visual analysis of a real motor insurance portfolio from Spain. The dataset covers three years of policy transactions and gives us a great opportunity to look at what actually drives profitability and claims volatility in a non-life insurance book. We will move through four parts: first some context on the data, then a look at the portfolio makeup, followed by profitability findings, and finally what is driving volatility.
:::

# Part 1: Dataset Overview

## About the Dataset

:::::: columns
::: {.column width="55%"}
**Source:** Spanish motor insurer microdata\
**Period:** January 2022 – December 2024\
**Records:** 354,140 policy transactions\
**Variables:** 47 across 4 groups

| Group             | Variables |
|-------------------|-----------|
| Policy & Insured  | 8         |
| Driver & Vehicle  | 10        |
| Premiums          | 8         |
| Claims & Incurred | 21        |
:::

:::: {.column width="45%"}
**Three derived metrics guide analysis:**

$$\text{Loss Ratio} = \frac{\text{Incurred}}{\text{Premium}}$$

$$\text{Claim Freq} = \frac{\text{Claims}}{\text{Exposure}}$$

$$\text{Severity} = \frac{\text{Incurred}}{\text{Claims}}$$

::: callout-note
Loss Ratio \< 1 = **Profitable**\
Loss Ratio \> 1 = **Unprofitable**
:::
::::
::::::

::: notes
So what are we working with? This is real anonymised data from a Spanish motor insurer, covering about 354,000 policy transactions over three years. Each row represents one policy in one year. The three metrics on the right are what we will use throughout the presentation. Loss ratio is our profitability measure. Anything below 1 means the insurer made money on that policy. Claim frequency tells us how often accidents happen per year of exposure. Severity tells us how costly each claim is when it does happen.
:::

## Data Quality

```{r}
#| fig-width: 9
#| fig-height: 3.5
#| fig-align: center

data.frame(
  variable    = c("vehicle_value", "fuel_type"),
  missing_pct = c(0.14, 0.37),
  label       = c("0.14%", "0.37%")
) |>
  ggplot(aes(x = reorder(variable, missing_pct), y = missing_pct)) +
  geom_col(fill = clr_blue, width = 0.45) +
  geom_text(aes(label = label),
            hjust = -0.15, size = 5.5, fontface = "bold",
            colour = clr_blue) +
  coord_flip(ylim = c(0, 0.55)) +
  scale_y_continuous(labels = function(x) paste0(x, "%")) +
  labs(title = "Dataset is Exceptionally Clean",
       subtitle = "Only 2 of 47 variables have any missing values — both under 0.4%",
       x = NULL, y = "Missing (%)") +
  theme_minimal(base_size = 15) +
  theme(panel.grid.major.y = element_blank(),
        plot.title    = element_text(face = "bold", size = 17),
        plot.subtitle = element_text(colour = "#555555"),
        plot.margin   = margin(10, 30, 10, 10))
```

- `fuel_type`: **0.37%** missing (1,311 records) — excluded when fuel type is analysed
- `vehicle_value`: **0.14%** missing (500 records) — excluded when vehicle value is analysed
- All other **45 variables** are 100% complete

::: notes
Good news on data quality. Out of 47 variables, only two have any missing values at all. Fuel type is missing for less than half a percent of records, and vehicle value for just 0.14 percent. Everything else is complete and clean, which is quite rare for a real-world dataset of this size.
:::

# Part 2: Portfolio Composition

## Insight 1: CC Policies Dominate the Portfolio

```{r}
#| fig-width: 10
#| fig-height: 4.2
#| fig-align: center

insurance |>
  count(policy_type) |>
  mutate(
    pct         = n / sum(n),
    policy_type = fct_reorder(policy_type, n),
    fill_col    = ifelse(policy_type == "CC", clr_blue, clr_grey2)
  ) |>
  ggplot(aes(x = policy_type, y = n, fill = fill_col)) +
  geom_col(show.legend = FALSE, width = 0.65) +
  geom_text(aes(label = paste0(round(pct*100, 1), "%")),
            hjust = -0.05, size = 5, fontface = "bold") +
  coord_flip() +
  scale_y_continuous(labels = comma,
                     expand = expansion(mult = c(0, 0.25))) +
  scale_fill_identity() +
  labs(title = "Policy Type Distribution (n = 354,140)",
       x = NULL, y = "Number of Policies") +
  theme_minimal(base_size = 14) +
  theme(panel.grid.major.y = element_blank(),
        plot.title = element_text(face = "bold"))
```

- **CC** (Third-party + add-ons) is the dominant policy at **57.8%** of the portfolio
- **COMP_E** (Comprehensive with excess) is second at **25.8%**
- Basic **TP** (Third-party only) is rare at just **1.6%** — most customers prefer broader coverage

::: notes
CC policies dominate at 58 percent. These are third-party liability combined with several add-ons like theft and glass. COMP_E at 26 percent is full comprehensive with an excess. What stands out is how rare basic TP policies are at just 1.6 percent. This tells us the insurer is largely serving customers who want broader protection.
:::

## Insight 2: Portfolio is Stable with Low-Risk Customers

```{r}
#| fig-width: 11
#| fig-height: 4.2
#| fig-align: center

p_status <- insurance |>
  count(policy_status) |>
  mutate(
    pct      = n / sum(n),
    fill_col = ifelse(policy_status == "Active", clr_blue, clr_grey2)
  ) |>
  ggplot(aes(x = policy_status, y = n, fill = fill_col)) +
  geom_col(show.legend = FALSE, width = 0.5) +
  geom_text(aes(label = paste0(round(pct*100, 1), "%")),
            vjust = -0.5, size = 6, fontface = "bold") +
  scale_y_continuous(labels = comma,
                     expand = expansion(mult = c(0, 0.15))) +
  scale_fill_identity() +
  labs(title = "Policy Status", x = NULL, y = "Count") +
  theme_minimal(base_size = 14) +
  theme(panel.grid.major.x = element_blank(),
        plot.title = element_text(face = "bold"))

p_bonus <- insurance |>
  count(bonus_score) |>
  mutate(
    pct      = n / sum(n),
    fill_col = ifelse(bonus_score == "Good", clr_blue, clr_grey2)
  ) |>
  ggplot(aes(x = bonus_score, y = n, fill = fill_col)) +
  geom_col(show.legend = FALSE, width = 0.5) +
  geom_text(aes(label = paste0(round(pct*100, 1), "%")),
            vjust = -0.5, size = 6, fontface = "bold") +
  scale_y_continuous(labels = comma,
                     expand = expansion(mult = c(0, 0.15))) +
  scale_fill_identity() +
  labs(title = "Bonus Score (Claims History)", x = NULL, y = "Count") +
  theme_minimal(base_size = 14) +
  theme(panel.grid.major.x = element_blank(),
        plot.title = element_text(face = "bold"))

p_status | p_bonus
```

- **89.2%** of policies are active; cancellation rate of **10.8%** warrants monitoring
- **96.4%** of customers have a **Good** bonus score — no recent claims history
- Bad-score customers are only **1.3%** of the portfolio but are a key risk segment

::: notes
On the left, 89 percent of policies are active and 11 percent cancelled. On the right, the bonus score is striking. 96 percent of customers have a Good score, meaning they have had no recent claims. This is very positive for profitability. But it also means the Bad group at 1.3 percent is very small so we need to be careful drawing conclusions from it statistically.
:::

## Insight 3: Experienced Drivers, Older Vehicles

```{r}
#| fig-width: 11
#| fig-height: 4.2
#| fig-align: center

p_age <- insurance |>
  ggplot(aes(x = driver_age)) +
  geom_histogram(binwidth = 5, fill = clr_blue, colour = clr_white) +
  scale_y_continuous(labels = comma) +
  labs(title = "Driver Age", x = "Age (years)", y = "Count") +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"))

p_vage <- insurance |>
  ggplot(aes(x = vehicle_age)) +
  geom_histogram(binwidth = 3, fill = clr_grey1, colour = clr_white) +
  scale_y_continuous(labels = comma) +
  labs(title = "Vehicle Age", x = "Age (years)", y = "Count") +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"))

p_val <- insurance |>
  filter(!is.na(vehicle_value)) |>
  ggplot(aes(x = vehicle_value)) +
  geom_histogram(bins = 40, fill = clr_grey2, colour = clr_white) +
  coord_cartesian(xlim = c(0, 80000)) +
  scale_x_continuous(labels = label_number(scale = 1e-3, suffix = "k")) +
  scale_y_continuous(labels = comma) +
  labs(title = "Vehicle Value (EUR, capped 80k)", x = "Value", y = "Count") +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"))

p_age | p_vage | p_val
```

- Most drivers are aged **40-60 years** — an experienced, lower-risk segment
- Vehicle age peaks at **15-35 years** — an older fleet than typical
- Vehicle values concentrate between **€15k-€40k** with a high-value tail

::: notes
Three things jump out from these charts. First, the typical driver here is in their mid-40s to mid-50s. Very few young drivers under 25 appear in this book. Second, the vehicles are quite old with many in the 15 to 35 year range. Third, vehicle values mostly sit between 15 and 40 thousand euros, but there are expensive vehicles in the tail that can generate very large claims.
:::

## Insight 4: Urban-Heavy, Inland-Dominant Portfolio

```{r}
#| fig-width: 11
#| fig-height: 4.2
#| fig-align: center

p_mun <- insurance |>
  count(municipality_type) |>
  mutate(
    pct = n / sum(n),
    municipality_type = fct_reorder(municipality_type, n),
    fill_col = ifelse(municipality_type == "Inland", clr_blue, clr_grey2)
  ) |>
  ggplot(aes(x = municipality_type, y = pct, fill = fill_col)) +
  geom_col(show.legend = FALSE, width = 0.55) +
  geom_text(aes(label = paste0(round(pct*100, 1), "%")),
            vjust = -0.5, size = 5, fontface = "bold") +
  scale_y_continuous(labels = percent,
                     expand = expansion(mult = c(0, 0.15))) +
  scale_fill_identity() +
  labs(title = "Municipality Type", x = NULL, y = NULL) +
  theme_minimal(base_size = 13) +
  theme(panel.grid.major.x = element_blank(),
        plot.title = element_text(face = "bold"))

p_circ <- insurance |>
  count(circulation_area) |>
  mutate(
    pct = n / sum(n),
    fill_col = ifelse(circulation_area == "Urban", clr_blue, clr_grey2)
  ) |>
  ggplot(aes(x = circulation_area, y = pct, fill = fill_col)) +
  geom_col(show.legend = FALSE, width = 0.45) +
  geom_text(aes(label = paste0(round(pct*100, 1), "%")),
            vjust = -0.5, size = 5, fontface = "bold") +
  scale_y_continuous(labels = percent,
                     expand = expansion(mult = c(0, 0.15))) +
  scale_fill_identity() +
  labs(title = "Circulation Area", x = NULL, y = NULL) +
  theme_minimal(base_size = 13) +
  theme(panel.grid.major.x = element_blank(),
        plot.title = element_text(face = "bold"))

p_fuel <- insurance |>
  filter(!is.na(fuel_type)) |>
  count(fuel_type) |>
  mutate(
    pct = n / sum(n),
    fill_col = ifelse(fuel_type == "Diesel", clr_blue, clr_grey2)
  ) |>
  ggplot(aes(x = fuel_type, y = pct, fill = fill_col)) +
  geom_col(show.legend = FALSE, width = 0.45) +
  geom_text(aes(label = paste0(round(pct*100, 1), "%")),
            vjust = -0.5, size = 5, fontface = "bold") +
  scale_y_continuous(labels = percent,
                     expand = expansion(mult = c(0, 0.15))) +
  scale_fill_identity() +
  labs(title = "Fuel Type", x = NULL, y = NULL) +
  theme_minimal(base_size = 13) +
  theme(panel.grid.major.x = element_blank(),
        plot.title = element_text(face = "bold"))

p_mun | p_circ | p_fuel
```

- **64.6%** of policies are in inland municipalities; islands account for only **6.8%**
- Urban and rural circulation are split **53.8% vs 46.2%** — near equal exposure
- **64.2%** of vehicles use diesel — relevant for repair cost patterns

::: notes
The geographic mix shows inland as the dominant segment at 65 percent. Urban and rural circulation are fairly balanced. Diesel vehicles make up 64 percent of the fleet. These splits matter because urban and rural drivers have very different claim frequencies as we will see shortly.
:::

# Part 3: Profitability Analysis

## Insight 5: 88% of Policies Have Zero Claims

```{r}
#| fig-width: 11
#| fig-height: 4.2
#| fig-align: center

lr_data <- insurance |> filter(!is.na(loss_ratio), total_premium > 0)

p_zero <- lr_data |>
  mutate(claim_group = ifelse(loss_ratio == 0,
                               "No Claims (LR = 0)",
                               "Had Claims (LR > 0)"),
         claim_group = fct_rev(factor(claim_group))) |>
  count(claim_group) |>
  mutate(
    pct      = n / sum(n),
    fill_col = ifelse(grepl("No Claims", claim_group), clr_grey2, clr_blue)
  ) |>
  ggplot(aes(x = claim_group, y = pct, fill = fill_col)) +
  geom_col(show.legend = FALSE, width = 0.5) +
  geom_text(aes(label = paste0(round(pct*100, 1), "%")),
            vjust = -0.5, size = 6.5, fontface = "bold") +
  scale_y_continuous(labels = percent,
                     expand = expansion(mult = c(0, 0.15))) +
  scale_fill_identity() +
  labs(title = "Claim Incidence", x = NULL, y = "% of Policies") +
  theme_minimal(base_size = 13) +
  theme(panel.grid.major.x = element_blank(),
        plot.title = element_text(face = "bold"))

p_dist <- lr_data |>
  filter(loss_ratio > 0, loss_ratio <= 3) |>
  ggplot(aes(x = loss_ratio)) +
  geom_histogram(binwidth = 0.05, fill = clr_blue,
                 colour = clr_white, linewidth = 0.2) +
  geom_vline(xintercept = 1, colour = clr_red,
             linetype = "dashed", linewidth = 1) +
  annotate("text", x = 1.05, y = Inf,
           label = "Break-even", hjust = 0, vjust = 1.8,
           colour = clr_red, size = 4, fontface = "bold") +
  scale_x_continuous(breaks = c(0, 0.5, 1, 1.5, 2, 2.5, 3)) +
  scale_y_continuous(labels = comma) +
  labs(title = "Loss Distribution (Claimants Only, LR <= 3)",
       x = "Loss Ratio", y = "Count") +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"))

p_zero | p_dist
```

- **88.1%** of policies generate zero claims; only **11.9%** file any claim
- Among claimants, the distribution peaks near **LR = 0.1-0.3** and decays toward break-even
- **9.0%** of all policies exceed break-even (LR \> 1) — these drive total portfolio losses

::: notes
I have split this into two panels. On the left, 88 percent of policies never filed a single claim. On the right, we look only at the claimants. The distribution peaks near 0.1 to 0.3 and decays toward the break-even line. Only 9 percent of all policies exceed break-even. These are the policies that really matter for the portfolio bottom line.
:::

## Insight 6: Comprehensive Policies Have Higher Loss Ratios

```{r}
#| fig-width: 10
#| fig-height: 4.8
#| fig-align: center

insurance |>
  filter(!is.na(loss_ratio), total_premium > 0,
         loss_ratio > 0, loss_ratio <= 2) |>
  mutate(fill_col = ifelse(policy_type %in% c("COMP_E","COMP_N"),
                            clr_blue, clr_grey2)) |>
  ggplot(aes(x = policy_type, y = loss_ratio, fill = fill_col)) +
  geom_boxplot(outlier.shape = NA, show.legend = FALSE,
               colour = "#444444", linewidth = 0.5) +
  geom_hline(yintercept = 1, colour = clr_red,
             linetype = "dashed", linewidth = 1) +
  annotate("text", x = 0.6, y = 1.05,
           label = "Break-even (LR = 1)", hjust = 0,
           colour = clr_red, size = 4, fontface = "bold") +
  coord_cartesian(ylim = c(0, 2)) +
  scale_fill_identity() +
  labs(title = "Loss Ratio by Policy Type (Claimants Only, LR <= 2)",
       subtitle = "Highlighted: Comprehensive policies (COMP_E, COMP_N)",
       x = "Policy Type", y = "Loss Ratio") +
  theme_minimal(base_size = 14) +
  theme(plot.title = element_text(face = "bold"),
        plot.subtitle = element_text(colour = "#555555"))
```

- **COMP_E** median LR = **0.92**; **COMP_N** median LR = **0.90** — highest among all types
- **CC** median LR = **0.91**; **TPG** = **0.98**; **TP** = **0.68** — basic coverage has lower loss ratios
- Comprehensive policies show the widest IQR, indicating **greater profitability volatility**

::: notes
With the axis capped at 2 and using boxplots, the differences are clearly visible. COMP_E and COMP_N sit higher than TP and basic coverage types. The boxplot shows that comprehensive policies not only have higher medians but also much wider spreads, confirming they are the main source of profitability uncertainty.
:::

## Insight 7: Bad Claims History Predicts Higher Loss Ratios

```{r}
#| fig-width: 10
#| fig-height: 4.8
#| fig-align: center

insurance |>
  filter(!is.na(loss_ratio), total_premium > 0,
         loss_ratio > 0, loss_ratio <= 1.5) |>
  mutate(fill_col = case_when(
    bonus_score == "Good"    ~ clr_grey2,
    bonus_score == "Neutral" ~ clr_grey1,
    bonus_score == "Bad"     ~ clr_blue
  )) |>
  ggplot(aes(x = bonus_score, y = loss_ratio, fill = fill_col)) +
  geom_violin(alpha = 0.8, show.legend = FALSE, trim = TRUE, colour = NA) +
  geom_boxplot(width = 0.1, outlier.shape = NA,
               show.legend = FALSE, fill = clr_white,
               colour = "#444444", linewidth = 0.6) +
  geom_hline(yintercept = 1, colour = clr_red,
             linetype = "dashed", linewidth = 1) +
  annotate("text", x = 0.6, y = 1.06,
           label = "Break-even (LR = 1)", hjust = 0,
           colour = clr_red, size = 4, fontface = "bold") +
  scale_fill_identity() +
  labs(title = "Loss Ratio by Bonus Score (Claimants Only, LR <= 1.5)",
       subtitle = "Highlighted: Bad-score policyholders",
       x = "Bonus Score (Claims History)", y = "Loss Ratio",
       caption = "< 3% of claimants exceed LR = 1.5") +
  theme_minimal(base_size = 14) +
  theme(plot.title = element_text(face = "bold"),
        plot.subtitle = element_text(colour = "#555555"),
        plot.caption  = element_text(colour = "#888888"))
```

- **Bad-score** median LR = **0.67**; IQR = 0.31-1.05 — wider and higher than Good
- **Good-score** median LR = **0.72**; IQR = 0.35-1.10 — tighter distribution
- Bad-score policyholders represent only **1.3%** of portfolio but show greater loss volatility

::: notes
The Bad-score violin highlighted in blue is noticeably wider than the grey Good group. This confirms the bonus-malus system is working correctly. It identifies riskier customers. Even though the Bad group is only 1.3 percent of the portfolio, their loss distribution is more spread out, contributing disproportionately to aggregate volatility.
:::

## Insight 8: Comprehensive Policies Cost More Per Claim

```{r}
#| fig-width: 10
#| fig-height: 4.8
#| fig-align: center

insurance |>
  filter(!is.na(claim_severity), total_claims > 0) |>
  filter(claim_severity <=
           quantile(claim_severity, 0.99, na.rm = TRUE)) |>
  mutate(fill_col = ifelse(policy_type %in% c("COMP_E","COMP_N"),
                            clr_blue, clr_grey2)) |>
  ggplot(aes(x = policy_type, y = claim_severity, fill = fill_col)) +
  geom_violin(alpha = 0.8, show.legend = FALSE, trim = TRUE, colour = NA) +
  geom_boxplot(width = 0.1, outlier.shape = NA,
               show.legend = FALSE, fill = clr_white,
               colour = "#444444", linewidth = 0.6) +
  scale_y_continuous(labels = comma) +
  scale_fill_identity() +
  labs(title = "Claim Severity by Policy Type (Capped at 99th Percentile)",
       subtitle = "Highlighted: Comprehensive policies (COMP_E, COMP_N)",
       x = "Policy Type", y = "Claim Severity (EUR per claim)") +
  theme_minimal(base_size = 14) +
  theme(plot.title    = element_text(face = "bold"),
        plot.subtitle = element_text(colour = "#555555"),
        plot.margin   = margin(10, 10, 20, 10))
```

- **COMP_E** mean severity = **€891** per claim; **COMP_N** = **€793**
- **CC** mean severity = **€821**; **TP** = **€928** (high due to small sample)
- Wide IQR across all types confirms **individual claim costs are highly unpredictable**

::: notes
Not only do comprehensive policies have more claims, but each claim also costs more. COMP_E and COMP_N show the highest and most variable severity distributions. The wide interquartile range across all policy types tells us that predicting how much any individual claim will cost is genuinely difficult regardless of policy type.
:::

# Part 4: Volatility Analysis

## Insight 9: Young Drivers File Claims 2.7x More Often

```{r}
#| fig-width: 10
#| fig-height: 4.5
#| fig-align: center

insurance |>
  filter(!is.na(claim_frequency), total_exposure > 0) |>
  mutate(age_group = cut(driver_age,
                          breaks = c(18,25,35,45,55,65,100),
                          labels = c("19-25","26-35","36-45",
                                     "46-55","56-65","66+"))) |>
  filter(!is.na(age_group)) |>
  group_by(age_group) |>
  summarise(mean_freq = mean(claim_frequency), .groups = "drop") |>
  mutate(fill_col = ifelse(age_group == "19-25", clr_blue, clr_grey2)) |>
  ggplot(aes(x = age_group, y = mean_freq, fill = fill_col)) +
  geom_col(width = 0.65, show.legend = FALSE) +
  geom_text(aes(label = round(mean_freq, 3)),
            vjust = -0.5, size = 5, fontface = "bold") +
  coord_cartesian(ylim = c(0, 0.85)) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
  scale_fill_identity() +
  labs(title = "Mean Claim Frequency by Driver Age Group",
       subtitle = "Highlighted: Youngest drivers (19-25) with highest claim rate",
       x = "Driver Age Group", y = "Mean Claims per Policy Year") +
  theme_minimal(base_size = 14) +
  theme(panel.grid.major.x = element_blank(),
        plot.title    = element_text(face = "bold"),
        plot.subtitle = element_text(colour = "#555555"))
```

- Drivers aged **19-25** have mean claim frequency of **0.716** — highest of all groups
- Drivers aged **56-65** average only **0.289** — the safest segment
- The 19-25 group files claims **2.7x more often** than the 56-65 group

::: notes
This chart shows a pattern actuaries know well. The youngest drivers from 19 to 25 have the highest claim frequency at 0.716 by a clear margin. Then it drops steadily through the middle age groups. The 19-25 group files claims 2.7 times more often than the safest group. One note: the 19-25 bar is based on a relatively small sample as very few young drivers appear in this portfolio.
:::

## Insight 10: Urban Inland Drivers Have Highest Claim Rates

```{r}
#| fig-width: 10
#| fig-height: 4.5
#| fig-align: center

insurance |>
  filter(!is.na(claim_frequency), total_exposure > 0) |>
  group_by(municipality_type, circulation_area) |>
  summarise(mean_freq = mean(claim_frequency), .groups = "drop") |>
  mutate(fill_col = ifelse(circulation_area == "Urban", clr_blue, clr_grey2)) |>
  ggplot(aes(x = municipality_type, y = mean_freq,
             fill = fill_col, group = circulation_area)) +
  geom_col(position = position_dodge(0.75), width = 0.65,
           show.legend = FALSE) +
  geom_text(aes(label = round(mean_freq, 3)),
            position = position_dodge(0.75),
            vjust = -0.5, size = 4.5, fontface = "bold") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.2))) +
  scale_fill_identity() +
  annotate("text", x = 3.4, y = 0.36,
           label = "Blue = Urban\nGrey = Rural",
           hjust = 1, size = 3.8, colour = "#555555") +
  labs(title = "Mean Claim Frequency by Municipality & Circulation Area",
       subtitle = "Highlighted: Urban circulation (consistently higher claim rates)",
       x = "Municipality Type", y = "Mean Claims per Policy Year") +
  theme_minimal(base_size = 14) +
  theme(panel.grid.major.x = element_blank(),
        plot.title    = element_text(face = "bold"),
        plot.subtitle = element_text(colour = "#555555"))
```

- **Inland Urban**: mean frequency = **0.371** — highest among all combinations
- **Coastal Urban**: **0.314** vs Coastal Rural: **0.284** — urban always higher
- Urban drivers consistently file **10-25% more claims** than rural across all regions

::: notes
Urban drivers consistently file more claims than rural drivers across all three municipality types. Inland urban shows the highest frequency at 0.371. The blue bars are always taller than the grey bars within each municipality type. The urban vs rural gap is consistent across all regions, confirming that circulation area is a strong predictor of claim volatility.
:::

## Insight 11: COMP_N Loss Ratio Rose to 1.17 by 2024

```{r}
#| fig-width: 10
#| fig-height: 4.5
#| fig-align: center

trend_data <- insurance |>
  filter(!is.na(loss_ratio), total_premium > 0) |>
  group_by(year, policy_type) |>
  summarise(mean_lr = mean(loss_ratio), .groups = "drop")

ggplot(trend_data,
       aes(x = year, y = mean_lr,
           colour = ifelse(policy_type %in% c("COMP_E","COMP_N"),
                            clr_blue, clr_grey1),
           group = policy_type,
           linewidth = ifelse(policy_type %in% c("COMP_E","COMP_N"), 1.5, 0.8))) +
  geom_line() +
  geom_point(size = 3.5) +
  geom_hline(yintercept = 1, linetype = "dashed",
             colour = clr_red, linewidth = 0.9) +
  annotate("text", x = 2022.05, y = 1.03,
           label = "Break-even", hjust = 0,
           colour = clr_red, size = 4, fontface = "bold") +
  geom_text_repel(
    data = filter(trend_data, year == 2024),
    aes(label = paste0(policy_type, ": ", round(mean_lr, 2))),
    direction    = "y",
    nudge_x      =  0.1,
    size         = 3.5,
    show.legend  = FALSE,
    segment.color = clr_grey2,
    segment.size  = 0.3
  ) +
  scale_x_continuous(breaks = c(2022, 2023, 2024),
                     expand = expansion(add = c(0.2, 0.8))) +
  scale_colour_identity() +
  scale_linewidth_identity() +
  labs(title = "Mean Loss Ratio by Policy Type: 2022-2024",
       subtitle = "Highlighted: COMP_E and COMP_N (crossed break-even in 2024)",
       x = "Year", y = "Mean Loss Ratio") +
  theme_minimal(base_size = 14) +
  theme(plot.title    = element_text(face = "bold"),
        plot.subtitle = element_text(colour = "#555555"),
        legend.position = "none")
```

- **COMP_N** rose from **0.71 (2022)** to **1.17 (2024)** — crossed break-even in 2024
- **COMP_E** rose from **0.64 (2022)** to **1.11 (2024)** — also crossed break-even
- **TP, TPG, CC** remain below break-even — basic coverage types remain profitable

::: notes
This is arguably the most alarming finding. COMP_N and COMP_E both crossed the break-even line in 2024. COMP_N went from 0.71 in 2022 to 1.17 in 2024. This means the insurer is now paying out more in claims than it collects in premiums for these policy types. This could indicate rising repair costs or medical inflation. Basic coverage types remain comfortably below break-even throughout the period.
:::

## Key Findings Summary

::::: columns
::: {.column width="50%"}
**Profitability Drivers**

- 🔵 **Policy Type**: COMP_E/COMP_N crossed break-even in 2024
- 🔵 **Trend**: COMP_N LR rose from 0.71 to 1.17 (2022-2024)
- ⬜ **Bonus Score**: Bad-score holders have wider loss distributions
- ⬜ **88.1%** of policies have zero claims
:::

::: {.column width="50%"}
**Volatility Drivers**

- 🔵 **Young Drivers**: 19-25 group has 2.7x higher claim frequency
- 🔵 **Urban Circulation**: Inland Urban frequency = 0.371 (highest)
- ⬜ **Geography**: Urban always higher than rural in all regions
- ⬜ **Claim severity** highly variable within each policy type
:::
:::::

::: callout-important
**Core finding:** 88.1% of policies have zero claims. The portfolio's entire profitability challenge comes from 9% of policies that exceed break-even — and comprehensive policies are now crossing that line.
:::

::: notes
To wrap up. COMP_E and COMP_N have crossed break-even in 2024 — this is the most actionable finding. On volatility, young drivers file claims 2.7 times more than the safest group. Urban circulation consistently drives higher frequencies. The overarching finding is that 88 percent of policies never generate a claim. The challenge is managing the small but growing share of loss-making comprehensive policies.
:::

## References {.smaller}

- Espinosa, P., Lledó, J., & Atance, D. (2026). *A detailed dataset of motor insurance policies with coverage-specific financial information*. Mendeley Data. https://doi.org/10.17632/sw4jmdb2sm.1

- Swiss Re Institute. (2019). *sigma 4/2019: World insurance: the great pivot east continues*. Swiss Re.

- Wickham, H. (2016). *ggplot2: Elegant Graphics for Data Analysis*. Springer-Verlag New York.

- Wilke, C. O. (2019). *Fundamentals of Data Visualization*. O'Reilly Media.
