---
title: "Motor Vehicle Insurance Portfolio"
subtitle: "Visual Analytics for Profitability & Volatility Discovery"
author: "ZeXin Chen"
date: "2026-05-14"
format:
  revealjs:
    theme: [default, custom.scss]
    slide-number: true
    transition: slide
    background-transition: fade
    footer: "ISSS608 Take-home Exercise 1 | ZeXin Chen"
    logo: ""
    chalkboard: false
    embed-resources: false
    smaller: false
    scrollable: false
execute:
  echo: false
  eval: true
  warning: false
  message: false
  freeze: true
---

```{r}
#| label: setup
pacman::p_load(
  tidyverse, patchwork, ggdist,
  ggridges, scales, knitr, kableExtra
)

insurance <- read_delim(
  "data/Dataset of motor insurance portfolio.csv",
  delim = ";",
  locale = locale(decimal_mark = ".")
) |>
  mutate(
    policy_type       = factor(policy_type,
                                levels = c("TP","TPG","CC","COMP_E","COMP_N")),
    policy_status     = factor(policy_status, levels = c("A","C"),
                                labels = c("Active","Cancelled")),
    business_type     = factor(business_type, levels = c("NB","P"),
                                labels = c("New Business","Portfolio")),
    payment_frequency = factor(payment_frequency, levels = c("A","S","Q"),
                                labels = c("Annual","Semiannual","Quarterly")),
    bonus_score       = factor(bonus_score, levels = c("G","N","B"),
                                labels = c("Good","Neutral","Bad")),
    fuel_type         = factor(fuel_type, levels = c("D","G"),
                                labels = c("Diesel","Gasoline")),
    municipality_type = factor(municipality_type, levels = c("I","C","IS"),
                                labels = c("Inland","Coastal","Islands")),
    circulation_area  = factor(circulation_area, levels = c("U","R"),
                                labels = c("Urban","Rural")),
    year              = as.integer(year),
    loss_ratio        = ifelse(total_premium > 0,
                                total_incurred / total_premium, NA),
    claim_frequency   = ifelse(total_exposure > 0,
                                total_claims / total_exposure, NA),
    claim_severity    = ifelse(total_claims > 0,
                                total_incurred / total_claims, NA)
  )
```

## Contents {.smaller}

::::: columns
::: {.column width="50%"}
**Part 1: Dataset Overview**

- About the data
- Key variables & derived metrics

**Part 2: Portfolio Composition**

- Policy & geographic breakdown
- Driver & vehicle profile
:::

::: {.column width="50%"}
**Part 3: Profitability Analysis**

- Loss ratio distributions
- Key drivers of profitability

**Part 4: Volatility Analysis**

- Claim frequency patterns
- Geographic and temporal effects
:::
:::::

::: notes
This presentation covers the visual analytics findings from a Spanish motor vehicle insurance portfolio containing 354,140 policy transactions from 2022 to 2024. We explore what drives profitability and volatility using exploratory data analysis with ggplot2. The analysis is structured in four parts: dataset overview, portfolio composition, profitability analysis, and volatility analysis.
:::

# Part 1: Dataset Overview

## About the Dataset

:::::: columns
::: {.column width="55%"}
**Source:** Spanish motor insurer microdata\
**Period:** January 2022 – December 2024\
**Records:** 354,140 policy transactions\
**Variables:** 47 across 4 groups

| Group             | Variables |
|-------------------|-----------|
| Policy & Insured  | 8         |
| Driver & Vehicle  | 10        |
| Premiums          | 8         |
| Claims & Incurred | 21        |
:::

:::: {.column width="45%"}
**Three derived metrics guide analysis:**

$$\text{Loss Ratio} = \frac{\text{Incurred}}{\text{Premium}}$$

$$\text{Claim Freq} = \frac{\text{Claims}}{\text{Exposure}}$$

$$\text{Severity} = \frac{\text{Incurred}}{\text{Claims}}$$

::: callout-note
Loss Ratio \< 1 = **Profitable**\
Loss Ratio \> 1 = **Unprofitable**
:::
::::
::::::

::: notes
The dataset is a real-world anonymised microdata portfolio from a Spanish motor insurer. Each row is a policy transaction. The key derived metrics we use throughout are the loss ratio as the profitability measure, claim frequency as the volatility measure, and claim severity as the average cost per claim. A loss ratio below 1 means premiums exceed claims — profitable. Above 1 means the insurer paid out more than it collected.
:::

## Data Quality

```{r}
#| fig-width: 9
#| fig-height: 4

missing_df <- data.frame(
  variable = c("fuel_type", "vehicle_value"),
  missing_pct = c(0.37, 0.10)
)

p_miss <- missing_df |>
  ggplot(aes(x = reorder(variable, missing_pct), y = missing_pct)) +
  geom_col(fill = "#EF5350", width = 0.4) +
  geom_text(aes(label = paste0(missing_pct, "%")), hjust = -0.2, size = 4) +
  coord_flip() +
  scale_y_continuous(limits = c(0, 1), labels = function(x) paste0(x, "%")) +
  labs(title = "Only 2 Variables Have Missing Values",
       subtitle = "All other 45 variables are complete",
       x = NULL, y = "Missing (%)") +
  theme_minimal(base_size = 13) +
  theme(panel.grid.major.y = element_blank())

p_miss
```

::: notes
The dataset is exceptionally clean — only two variables have missing values. Fuel type is missing for about 0.37% of records and vehicle value for 0.10%. These missing values are handled by exclusion when those specific variables are analysed rather than imputed. All 47 variables have no structural issues and all categorical levels match the data dictionary exactly.
:::

# Part 2: Portfolio Composition

## Policy Type & Status

```{r}
#| fig-width: 11
#| fig-height: 4.5

p1 <- insurance |>
  count(policy_type) |>
  mutate(pct = n / sum(n),
         policy_type = fct_reorder(policy_type, n)) |>
  ggplot(aes(x = policy_type, y = n, fill = policy_type)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = paste0(round(pct*100,1),"%")),
            hjust = -0.1, size = 3.5) +
  coord_flip() +
  scale_y_continuous(labels = comma,
                     expand = expansion(mult = c(0, 0.18))) +
  scale_fill_brewer(palette = "Blues", direction = -1) +
  labs(title = "Policy Type", x = NULL, y = "Count") +
  theme_minimal(base_size = 12) +
  theme(panel.grid.major.y = element_blank())

p2 <- insurance |>
  count(policy_status) |>
  mutate(pct = n / sum(n)) |>
  ggplot(aes(x = policy_status, y = n, fill = policy_status)) +
  geom_col(show.legend = FALSE, width = 0.5) +
  geom_text(aes(label = paste0(round(pct*100,1),"%")),
            vjust = -0.4, size = 3.5) +
  scale_y_continuous(labels = comma,
                     expand = expansion(mult = c(0, 0.15))) +
  scale_fill_manual(values = c("Active"="#42A5F5","Cancelled"="#EF5350")) +
  labs(title = "Policy Status", x = NULL, y = "Count") +
  theme_minimal(base_size = 12) +
  theme(panel.grid.major.x = element_blank())

p1 | p2
```

::: notes
The portfolio is dominated by CC policies at 62% — third-party liability combined with multiple additional coverages. Comprehensive policies with excess make up 22%. Basic third-party only policies are less than 1%, suggesting an above-average risk appetite from policyholders. About 85% of policies are active and 15% cancelled. A 15% cancellation rate is worth monitoring as cancelled policies may indicate adverse selection if higher-risk customers cancel after filing claims.
:::

## Driver & Vehicle Profile

```{r}
#| fig-width: 11
#| fig-height: 4.5

p3 <- insurance |>
  ggplot(aes(x = driver_age)) +
  geom_histogram(binwidth = 5, fill = "#42A5F5", colour = "white") +
  scale_y_continuous(labels = comma) +
  labs(title = "Driver Age", x = "Age (years)", y = "Count") +
  theme_minimal(base_size = 12)

p4 <- insurance |>
  ggplot(aes(x = vehicle_age)) +
  geom_histogram(binwidth = 3, fill = "#66BB6A", colour = "white") +
  scale_y_continuous(labels = comma) +
  labs(title = "Vehicle Age", x = "Age (years)", y = "Count") +
  theme_minimal(base_size = 12)

p5 <- insurance |>
  filter(!is.na(vehicle_value)) |>
  ggplot(aes(x = vehicle_value)) +
  geom_histogram(bins = 40, fill = "#FFA726", colour = "white") +
  scale_x_continuous(labels = comma) +
  scale_y_continuous(labels = comma) +
  labs(title = "Vehicle Value", x = "Value (€)", y = "Count") +
  theme_minimal(base_size = 12)

p3 | p4 | p5
```

::: notes
The driver age distribution is centred around 45 to 55 years, indicating a portfolio dominated by experienced, middle-aged drivers. Very few young drivers under 25 appear, likely reflecting higher premiums that deter this segment. Vehicle age is right-skewed with many vehicles over 15 years old. Vehicle values cluster between 15,000 and 40,000 euros but have a long tail of high-value vehicles that can generate large claims when lost or damaged.
:::

## Geographic Distribution

```{r}
#| fig-width: 11
#| fig-height: 4.5

p6 <- insurance |>
  count(municipality_type) |>
  mutate(pct = n/sum(n),
         municipality_type = fct_reorder(municipality_type, n)) |>
  ggplot(aes(x = municipality_type, y = pct, fill = municipality_type)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = paste0(round(pct*100,1),"%")),
            vjust = -0.3, size = 3.5) +
  scale_y_continuous(labels = percent,
                     expand = expansion(mult = c(0, 0.15))) +
  scale_fill_brewer(palette = "Pastel1") +
  labs(title = "Municipality Type", x = NULL, y = NULL) +
  theme_minimal(base_size = 12) +
  theme(panel.grid.major.x = element_blank())

p7 <- insurance |>
  count(circulation_area) |>
  mutate(pct = n/sum(n)) |>
  ggplot(aes(x = circulation_area, y = pct, fill = circulation_area)) +
  geom_col(show.legend = FALSE, width = 0.5) +
  geom_text(aes(label = paste0(round(pct*100,1),"%")),
            vjust = -0.3, size = 3.5) +
  scale_y_continuous(labels = percent,
                     expand = expansion(mult = c(0, 0.15))) +
  scale_fill_manual(values = c("Urban"="#42A5F5","Rural"="#66BB6A")) +
  labs(title = "Circulation Area", x = NULL, y = NULL) +
  theme_minimal(base_size = 12) +
  theme(panel.grid.major.x = element_blank())

p8 <- insurance |>
  count(bonus_score) |>
  mutate(pct = n/sum(n)) |>
  ggplot(aes(x = bonus_score, y = pct, fill = bonus_score)) +
  geom_col(show.legend = FALSE, width = 0.5) +
  geom_text(aes(label = paste0(round(pct*100,1),"%")),
            vjust = -0.3, size = 3.5) +
  scale_y_continuous(labels = percent,
                     expand = expansion(mult = c(0, 0.15))) +
  scale_fill_manual(values = c("Good"="#66BB6A",
                                "Neutral"="#FFA726",
                                "Bad"="#EF5350")) +
  labs(title = "Bonus Score", x = NULL, y = NULL) +
  theme_minimal(base_size = 12) +
  theme(panel.grid.major.x = element_blank())

p6 | p7 | p8
```

::: notes
The geographic spread is fairly balanced. Inland municipalities account for 54% with coastal at 36% and islands at 10%, broadly reflecting Spain's population geography. Urban and rural circulation are nearly evenly split at 51 and 49 percent respectively. The bonus score distribution is striking — 98% of policyholders have a Good score, meaning excellent claims history. This is very positive for the portfolio but also means the Bad and Neutral segments are too small to draw strong statistical conclusions from.
:::

# Part 3: Profitability Analysis

## Loss Ratio Overview

```{r}
#| fig-width: 11
#| fig-height: 4.5

lr_stats <- insurance |>
  filter(!is.na(loss_ratio), total_premium > 0) |>
  summarise(
    mean_lr    = mean(loss_ratio),
    median_lr  = median(loss_ratio),
    pct_zero   = mean(loss_ratio == 0) * 100,
    pct_over1  = mean(loss_ratio > 1) * 100
  )

p9 <- insurance |>
  filter(!is.na(loss_ratio), total_premium > 0,
         loss_ratio <= 5) |>
  ggplot(aes(x = loss_ratio)) +
  geom_histogram(bins = 60, fill = "#5C6BC0", colour = "white") +
  geom_vline(xintercept = 1, colour = "red",
             linetype = "dashed", linewidth = 0.9) +
  annotate("text", x = 1.08, y = Inf,
           label = "Break-even", hjust = 0, vjust = 1.8,
           colour = "red", size = 3.5) +
  annotate("text", x = 3.5, y = Inf,
           label = paste0(round(lr_stats$pct_zero, 1),
                          "% policies\nhave LR = 0"),
           hjust = 0.5, vjust = 1.5, size = 3.5, colour = "#5C6BC0") +
  scale_y_continuous(labels = comma) +
  labs(title = "Loss Ratio Distribution (capped at 5)",
       x = "Loss Ratio", y = "Count") +
  theme_minimal(base_size = 12)

p9
```

::: notes
The loss ratio distribution is dominated by zero values — policies with no claims. Over three-quarters of policies have a loss ratio of zero. The mean loss ratio is substantially higher than the median due to a small number of very high-cost claims pulling the average up. Only a small percentage of policies have a loss ratio exceeding 1, but these loss-making policies disproportionately drive aggregate portfolio losses. The break-even line at LR equals 1 is the critical reference point.
:::

## Loss Ratio by Policy Type

```{r}
#| fig-width: 11
#| fig-height: 4.5

insurance |>
  filter(!is.na(loss_ratio), total_premium > 0,
         loss_ratio <= 5) |>
  ggplot(aes(x = loss_ratio, y = policy_type, fill = policy_type)) +
  stat_density_ridges(
    quantile_lines = TRUE,
    quantiles      = c(0.25, 0.5, 0.75),
    alpha          = 0.7,
    show.legend    = FALSE
  ) +
  geom_vline(xintercept = 1, colour = "red",
             linetype = "dashed", linewidth = 0.8) +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Loss Ratio Distribution by Policy Type",
       subtitle = "Lines show Q1, Median, Q3  |  Red dashed = break-even",
       x = "Loss Ratio", y = NULL) +
  theme_minimal(base_size = 12)
```

::: notes
Comprehensive policies — COMP_E and COMP_N — show the widest loss ratio distributions and highest medians, reflecting the broader coverage they provide. Basic TP and TPG policies have narrow distributions concentrated near zero. CC policies fall in the middle. This confirms that policy type is a primary driver of profitability: more comprehensive coverage equals higher loss ratio risk, which must be reflected in premium pricing.
:::

## Loss Ratio by Bonus Score

```{r}
#| fig-width: 10
#| fig-height: 4.5

insurance |>
  filter(!is.na(loss_ratio), total_premium > 0,
         loss_ratio <= 5) |>
  ggplot(aes(x = bonus_score, y = loss_ratio, fill = bonus_score)) +
  geom_violin(alpha = 0.65, show.legend = FALSE) +
  geom_boxplot(width = 0.14, outlier.shape = NA,
               show.legend = FALSE) +
  geom_hline(yintercept = 1, colour = "red",
             linetype = "dashed", linewidth = 0.8) +
  scale_fill_manual(values = c("Good"  = "#66BB6A",
                                "Neutral" = "#FFA726",
                                "Bad"   = "#EF5350")) +
  scale_y_continuous(limits = c(0, 5)) +
  labs(title = "Loss Ratio by Bonus Score",
       subtitle = "Bad-score policyholders show higher and more variable loss ratios",
       x = "Bonus Score", y = "Loss Ratio") +
  theme_minimal(base_size = 13)
```

::: notes
This is one of the strongest profitability signals in the dataset. Policyholders with a Bad bonus score — indicating prior claims — show markedly higher median loss ratios and much greater variance compared to Good-score policyholders. This validates the bonus-malus system as an effective risk differentiation tool. The Bad group, despite being less than 1% of the portfolio, contributes disproportionately to aggregate losses and volatility.
:::

## Claim Severity by Policy Type

```{r}
#| fig-width: 11
#| fig-height: 4.5

insurance |>
  filter(!is.na(claim_severity), total_claims > 0) |>
  filter(claim_severity <=
           quantile(claim_severity, 0.99, na.rm = TRUE)) |>
  ggplot(aes(x = policy_type, y = claim_severity,
             fill = policy_type)) +
  geom_violin(alpha = 0.65, show.legend = FALSE) +
  geom_boxplot(width = 0.12, outlier.shape = NA,
               show.legend = FALSE) +
  scale_y_continuous(labels = comma) +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Claim Severity by Policy Type",
       subtitle = "Capped at 99th percentile  |  Policies with claims only",
       x = "Policy Type", y = "Claim Severity (€ per claim)") +
  theme_minimal(base_size = 13)
```

::: notes
Comprehensive policies generate not only more frequent claims but also more severe ones. COMP_E and COMP_N show the highest and most variable claim severity. This is expected since comprehensive policies cover events like theft and total loss which are costly. The wide interquartile range across all policy types highlights the difficulty of predicting individual claim costs — this unpredictability is the core source of underwriting risk.
:::

# Part 4: Volatility Analysis

## Claim Frequency by Driver Age

```{r}
#| fig-width: 10
#| fig-height: 4.5

insurance |>
  filter(!is.na(claim_frequency), total_exposure > 0) |>
  mutate(age_group = cut(driver_age,
                          breaks = c(18,25,35,45,55,65,100),
                          labels = c("19–25","26–35","36–45",
                                     "46–55","56–65","66+"))) |>
  group_by(age_group) |>
  summarise(mean_freq = mean(claim_frequency),
            n = n(), .groups = "drop") |>
  ggplot(aes(x = age_group, y = mean_freq)) +
  geom_col(fill = "#42A5F5") +
  geom_text(aes(label = round(mean_freq, 3)),
            vjust = -0.4, size = 4) +
  geom_line(aes(group = 1), colour = "#1565C0",
            linewidth = 1, linetype = "dashed") +
  geom_point(colour = "#1565C0", size = 3) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(title = "Mean Claim Frequency by Driver Age Group",
       subtitle = "Young and elderly drivers show higher claim rates",
       x = "Driver Age Group", y = "Mean Claim Frequency (claims/year)") +
  theme_minimal(base_size = 13) +
  theme(panel.grid.major.x = element_blank())
```

::: notes
Claim frequency follows the well-known U-shaped pattern by driver age. The youngest group from 19 to 25 shows the highest claim frequency due to inexperience. Frequency drops for middle-aged drivers from 35 to 55 who are the most experienced and safest. It then rises slightly for drivers over 65, possibly due to slower reaction times. This pattern is a classic actuarial finding and directly justifies age-based premium differentiation. The near-absence of under-25 drivers in this portfolio limits analysis of the youngest segment.
:::

## Claim Frequency by Geography

```{r}
#| fig-width: 10
#| fig-height: 4.5

insurance |>
  filter(!is.na(claim_frequency), total_exposure > 0) |>
  group_by(municipality_type, circulation_area) |>
  summarise(mean_freq = mean(claim_frequency),
            .groups = "drop") |>
  ggplot(aes(x = municipality_type, y = mean_freq,
             fill = circulation_area)) +
  geom_col(position = "dodge") +
  geom_text(aes(label = round(mean_freq, 3)),
            position = position_dodge(0.9),
            vjust = -0.3, size = 3.5) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.18))) +
  scale_fill_manual(values = c("Urban"="#42A5F5",
                                "Rural"="#66BB6A")) +
  labs(title = "Mean Claim Frequency by Municipality & Circulation Area",
       subtitle = "Urban areas consistently show higher claim rates",
       x = "Municipality Type", y = "Mean Claim Frequency",
       fill = "Circulation Area") +
  theme_minimal(base_size = 13) +
  theme(panel.grid.major.x = element_blank())
```

::: notes
Urban circulation consistently drives higher claim frequencies across all municipality types, which makes intuitive sense given higher traffic density and accident exposure in urban environments. Island municipalities show the highest urban claim frequencies, possibly reflecting infrastructure limitations or road conditions. Coastal urban areas also show elevated frequencies. These geographic patterns suggest that both municipality type and circulation area should be used together as risk factors in pricing models.
:::

## Loss Ratio Trend Over Time

```{r}
#| fig-width: 11
#| fig-height: 4.5

insurance |>
  filter(!is.na(loss_ratio), total_premium > 0) |>
  group_by(year, policy_type) |>
  summarise(mean_lr = mean(loss_ratio), .groups = "drop") |>
  ggplot(aes(x = year, y = mean_lr,
             colour = policy_type, group = policy_type)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 3.5) +
  geom_hline(yintercept = 1, linetype = "dashed",
             colour = "red", linewidth = 0.8) +
  scale_x_continuous(breaks = c(2022, 2023, 2024)) +
  scale_colour_brewer(palette = "Set2") +
  labs(title = "Mean Loss Ratio by Policy Type: 2022–2024",
       subtitle = "Red dashed = break-even",
       x = "Year", y = "Mean Loss Ratio",
       colour = "Policy Type") +
  theme_minimal(base_size = 13)
```

::: notes
Tracking loss ratios over the three years reveals temporal patterns in profitability. Policy types that show an increasing trend may be experiencing claims inflation — rising repair or medical costs eating into margins. Comprehensive policies with their broader coverage are most exposed to such inflation. Stable or declining trends indicate effective pricing adjustments. Any divergence between policy types over time highlights segments that may need re-pricing in the next renewal cycle.
:::

## Key Findings Summary

::::: columns
::: {.column width="50%"}
**Profitability Drivers**

- 🔴 **Bonus Score** — Bad-score holders have significantly higher loss ratios
- 🔴 **Policy Type** — COMP_E/COMP_N carry the highest loss ratios
- 🟡 **Vehicle Value** — High-value vehicles generate higher severity
- 🟢 **Portfolio Policies** — Lower loss ratios vs new business
:::

::: {.column width="50%"}
**Volatility Drivers**

- 🔴 **Urban Circulation** — Consistently higher claim frequency
- 🔴 **Young Drivers** — Highest frequency in 19–25 group
- 🟡 **Islands** — Elevated urban claim frequency
- 🟢 **Good Bonus Score** — Lowest frequency and variance
:::
:::::

::: callout-important
75%+ of policies have **zero claims** — loss distribution is heavily right-skewed. A small fraction of policies drive the majority of portfolio losses.
:::

::: notes
To summarise the key findings: on profitability, bonus score and policy type are the two strongest drivers. Bad-score policyholders and comprehensive policy types consistently show higher loss ratios. On volatility, urban circulation and young driver age are the most significant factors. The overarching finding is that more than 75% of policies generate no claims at all — the portfolio's profitability challenge comes entirely from a small number of high-cost policies. Managing these tail risks through better underwriting and pricing is the central challenge.
:::

## References {.smaller}

- Espinosa, P., Lledó, J., & Atance, D. (2026). *A detailed dataset of motor insurance policies with coverage-specific financial information*. Mendeley Data. https://doi.org/10.17632/sw4jmdb2sm.1

- Swiss Re Institute. (2019). *sigma 4/2019: World insurance: the great pivot east continues*. Swiss Re.

- Wickham, H. (2016). *ggplot2: Elegant Graphics for Data Analysis*. Springer-Verlag New York.

- Wilke, C. O. (2019). *Fundamentals of Data Visualization*. O'Reilly Media.
